---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 19, part 3: Residual Diagnostics'
author: "Jinko Graham"
date: '2018-11-12'

output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```




## Residual Diagnostics

* Residuals are the main tool for checking model assumptions and 
identifying outliers
* Recall the model assumptions:
    1. The linear predictor is correctly specified.
    2. The random errors have constant SD.
    3. The random errors are normally distributed.
* The residuals are observed minus fitted values: $y_i-\hat{y}_i$, 
* In Chapter 18, we plotted residuals _vs._ 
fitted values to check assumptions 1 and 2, and also to informally
identify outliers.
* To check the normal errors assumption and detect
outliers more formally, we'll define
the _Q-Q plot_ and _standardized residuals_.
* But first let's check assumptions 1 and 2 
for the MLR model that we fit to 
low-birthweight-babies data, 
by plotting the residuals _vs._ fitted 
values.


## Residuals versus Fitted Values

* Load the data, fit the MLR model and do the plot $\ldots$

\scriptsize

```{r,fig.height=3.0}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
lfit2 <- lm(headcirc ~ gestage + birthwt,data=lbwt)
plot(lfit2,which=1)
```

\normalsize

## Comments

* There are no obvious missed trends. As far
as we can tell, the linear predictor looks properly specified. 

* There is no obvious funnel pattern in the residuals that might suggest that the error terms have non-constant SD.

* The 3 most extreme (farthest from zero)
residuals are labelled
by their case number. Case 31 in particular stands out.

* Note: Residual diagnostics can be subjective. 
    + Whether or not a plot suggests
    that an assumption is violated can depend on the person looking at it.
    + My concern is that you understand which plots
    check which assumptions and that you can form an opinion
    about the assumptions. 
    + Different people may have differing opinions. 
    

## Software Notes

* Recall that R's `plot()` function can do six
different diagnostic plots, specified by the
`which` argument.
    + The first plot (`which=1`) is the residual _vs_ fitted values.
    + The second plot (`which=2`) is the Q-Q plot which we haven't seen yet but which we will discuss next.
    + In this course, we won't be interested in the others.


## Q-Q Plots

- A quantile-quantile (Q-Q) plot is a plot of the _quantiles_ of one distribution _vs._ another.
    - If the two distributions have similar shape, the points
    on such a plot should fall roughly on a straight line.
    - We will define quantiles on the next slide.
- Our interest is in using Q-Q plots to 
compare the distribution of
residuals to the distribution they _should have_ under 
the model assumption of _normal random errors_.


## Quantiles

\small

* The $p$th quantile, $q_p$, of a distribution 
is the cutpoint 
such that the proportion $p$ of the distribution
is less than or equal to the cutpoint.

```{r,echo=FALSE, fig.height=2.5}
x<-seq(from=-4,to=+4, length=100)
plot(x,dnorm(x), type="l")
lines(x=c(-1.2,-1.2),y=c(-0.1,dnorm(-1.2)))
text(x=-1.5,y=.06, labels="p")
text(x=0.0,y=.06, labels="1-p")
axis(side=1, at=-1.2, labels=expression(q[p]))
```

* Examples: 
    1. The median is the 0.5 quantile, or $q_{.5}$,
    cutting the distribution into bottom and top halves
    2. The first quartile 
    is the 0.25 quantile, or $q_{.25}$, cutting 
    the distribution into the bottom quarter and the top 
    three quarters
  

## Distribution of Residuals

- We may _standardize_ the 
residuals to have a common distribution.
We'll skip the details.
<!--
- For the $i$th case, define the _standardized residual_ as
$$ r_i = \frac{y_i - \hat{y}_i}{\hat{\sigma}\sqrt{1-h_{i}}}, \quad \; {\rm where}$$
    + $\hat{\sigma}$ is the estimate of the error SD, i.e. of the SD of the distribution of the $\epsilon$'s, and
    + $0<1-h_{i}<1-1/n$ measures how typical case $i$'s explanatory variables are within the dataset. 
- Cases with atypical explanatory variables
have values of $1-h_{i}$ close to zero, and larger standardized residuals.
-->

\bigskip

* Under the model assumptions, the standardized residuals have 
    a $t$ distribution with $n-q-1$ df.

\bigskip

* **Rule of thumb**: Standardized residuals less than $-3$ or
    greater than 3 are considered to be 
    obvious outliers.
    
## Q-Q Plot of Standardized Residuals

* Idea: Plot the quantiles of the 
empirical distribution of the standardized residuals
against the quantiles of the $t$ distribution
with $n-q-1$ df.
    + Should get a straight line of slope 1 that 
    cuts through the origin.
    + If not, this suggests a violation of the assumption that the
    error terms are normally distributed
    with mean 0 and constant SD.
    
* When $n-q-1$ is of size 20 or more,
the $t$ distribution is similar to 
the standard normal distribution in shape.
    + Therefore, most software, such as R's `plot()` function, plots
    the quantiles of the standardized residuals against the quantiles
    of the standard normal distribution:

## Example Q-Q Plot

* For the low-birthweight babies,
$n=100$ babies and we fit $q=2$ explanatory
variables, `gestage` and `birthwt`.
    + So $n-q-1=97$ which is large enough
    to approximate the $t$ distribution by
    a standard normal distribution.

\scriptsize

```{r,fig.height=3.0}
plot(lfit2,which=2)
```

## Comments

* Mostly, the points on the Q-Q plot fall along 
the straight line that cuts through the origin with slope 1.
* The exceptions are in the upper tail of the 
distribution of residuals, and labelled as cases
31, 33 and 55.
    + More on outliers next.

## Identifying Outliers

- Standardized
residuals less than $-3$ or greater than 3 are 
considered to be obvious outliers.
- Extract the values of the standardized 
residuals with the `rstandard()` function; 

- E.G., `rstandard(lfit2)` gives the
    standardized residuals from the `lm()` object `lfit2` that
    fits the MLR model of `headcirc` as the response variable 
    and `gestage` and `birthwt` as explanatory 
    variables. 
    
<!--
    + Let's print out the standardized residuals
    for the first 10 babies in the dataset:
    
\scriptsize
```{r}
rstandard(lfit2)[1:10]
```

## 

+ We can `sort()` these standardized residuals so that they 
are ordered from smallest to largest. 

+ And then apply the `head()` function to the sorted 
  list, to get the smallest 7 residuals; 

  
\scriptsize

```{r}
head(sort(rstandard(lfit2)),n=7)
```

\normalsize

- Or, we can apply the `tail()` function to the sorted 
  list, to get the largest 7 residuals; 

  
\scriptsize

```{r}
tail(sort(rstandard(lfit2)),n=7)
```

\normalsize
-->

* From the resulting output, we see that cases 31 and 33 are outliers.
Their standardized residuals $r_{31}$ and $r_{33}$ are greater than 3.
As all other $r_i$'s have $|r_i|<3$, there are no other obvious outliers.


## Summary

* We've covered residual diagnostics including:
    1. A plot of residuals _vs._ fitted values to 
    check the assumptions that 
        + the linear predictor is correctly specified and 
        + the error SD is constant
        \medskip
        
    2. A Q-Q plot of the standardized residuals _vs._ the 
    quantiles of the standard normal to check the assumption of
    normal errors
    \bigskip
    
    3. A printout of the sorted list of standardized residuals
    (the head and tail ends are usually enough) to identify obvious outliers with extreme standardized residuals such that $r_i < -3$  or
    $r_i > 3$.

