---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 19, part 1: Multiple Linear Regression Models'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Multiple Regression
* Have been looking at _simple linear regression_, to understand the relationship between the response and a single explanatory variable.

* Now, transition to _multiple linear regression_, in which we study the relationship between the response and 
multiple explanatory variables.
* Why multiple regression?
    + A model with more than one explanatory variable may better explain the
    the mean of the response variable.
    + Allows for possible synergy between explanatory variables.
    + Allows adjustment for confounding variables 
        + (Recall: A confounding variable is an extraneous variable that is
associated with both the outcome and the exposure of interest).
    + Can improve the precision of our predictions.

## Example and Notation

Consider the `lbwt` dataset

\scriptsize
```{r,echo=FALSE,display=FALSE}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
```
\normalsize

* The response variable, $Y$, is head circumference,
    with observed values denoted by $y$.

* One explanatory variable is gestational 
    age, $X_1$, with observed values denoted by $x_1$.
    
* Let's consider a second explanatory variable, 
    birth weight, $X_2$, with observed values denoted by $x_2$.
    
\scriptsize
```{r}
head(lbwt,n=3)
```


## Multiple Linear Regression (MLR) Model 

* Let's fit a model with **two** explanatory variables:
    + $X_1$, the gestational age, as before, and a new variable
    + $X_2$, the birth weight.

\scriptsize

```{r}
fit2 <- lm(headcirc ~ gestage + birthwt, data=lbwt)
summary(fit2)$coefficients
```

\normalsize

* Consider the following interpretation 
of the fitted MLR model:
\begin{quote}
Among infants {\bf of the same birth weight},
a one week increase in gestational age is associated
with a 0.45cm increase in head circumference.
\end{quote}

* We will motivate this intepretation in what follows.

## Multiple Regression Overview

* We have one response variable, $Y$.
* We have $q$ explanatory variables denoted by
$X_1, X_2, \ldots, X_q$, with observed values
$x_1, x_2, \ldots, x_q$, respectively.
* The regression model describes how the average value of $Y$ changes
as $x_1, \ldots, x_q$ change.
* We use the method of least squares to fit the model to our data.
* Under modelling assumptions, we can
    * infer the slopes of the regression model in the population
    from the slopes fitted in our sample, and
    * make predictions from the model we have fitted to our data.
* Model assumptions are checked _after_ the model is fit to 
our sample of data.

## Model Overview

* Model components remain the same as in SLR: 
    (i) linear predictor, 
    (ii) normal error terms, and
    (iii) constant SD, $\sigma_y$.
* The linear predictor component is generalized to include more explanatory variables, but the 
normal errors and constant SD assumptions 
are as-before.
* Also, as in SLR,
we assume independent observations.

## Linear Predictor

* In SLR, the linear predictor for the population
mean response is
$$\mu_{y|x} =\alpha + \beta x,$$
where $x$ is a value of the single explanatory variable $X$.

* In MLR, the linear predictor is generalized to
$$\mu_{y|x_1,\ldots,x_q} = \alpha + \beta_1 x_1 + \ldots + \beta_q x_q$$
* $\mu_{y|x_1,\ldots,x_q}$ is the population mean value of $Y$ for all data points with 
$X_1=x_1, \ldots, X_q = x_q$.
* Individual regression coefficients 
$\beta_k$ are the change in $\mu_{y|x_1,\ldots,x_q}$ for a one-unit increase in $x_k$
_holding all other $x$'s fixed_.
    + In the above example with gestational age ($X_1$)
    and birth weight ($X_2$), the interpretation of 
    $\beta_1$ is the increase in $\mu_{y|x_1,x_2}$
    for a one week increase in gestational age, 
    holding birth weight fixed.





## Interpreting Fitted Regression Coefficients

* Fit MLR model to `lbwt` data. 

\scriptsize
```{r}
fit2 <- lm(headcirc ~ gestage + birthwt, data=lbwt)
summary(fit2)$coefficients
```
\normalsize

\vspace*{-.2in}

* Interpretation of fitted coefficient $\hat{\beta}_1$ for gestational age: 
    + "For a given birth weight, a one week
increase in gestational age is associated with 
a 0.45cm increase in head circumference."
* Interpretation of fitted coefficient $\hat{\beta}_2$ for birth weight:
    + "For a given gestational age, a one gram
increase in birth weight is associated with a 0.0047cm increase in head circumference."
    + Or (since a 1g increase in weight is too fine-grained)
    "For a given gestational age, a **100** gram
increase in birth weight is associated with
a 0.47cm increase in head circumference."

## Software Notes

* `lm()` will fit simple _and_ multiple regression models.
\bigskip

* To use `lm()` to fit multiple regression models, we include multiple explanatory variables on the
right-hand side of the formula, separated by the `+`
sign.
    + In the example, the model formula    
    `headcirc ~ gestage + birthwt`    
    includes both `gestage` and `birthwt` as 
    explanatory variables.
    