---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 19, part 4: Statistical Interaction and Confounding'
author: "Jinko Graham"
date: '2018-11-12'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Overview

\small

- We often study the relationship between a response variable, $Y$, 
and explanatory variable, $X_1$, adjusted for other variables.
- Example: Study of low birthweight babies.
    - Response $Y$ is head circumference    
    - Explanatory variable $X_1$ is gestational age.
- However, an extraneous variable $X_2$, such as birth weight, can ***modify*** the effect of $X_1$ on $Y$.
    - We looked at ***effect modification*** previously, when studying association between a categorical outcome variable, $Y$, and 
    a categorical exposure variable, $X_1$. 
    - Also
    referred to as ***statistical interaction***.


## Steps

- Suppose we're primarily interested in the association between $Y$ and $X_1$.
- Have also collected data on an extraneous variable, $X_2$.
- Suggested steps are:

1. First consider whether $X_2$ ***modifies*** the effect of $X_1$ on $Y$ 
    +  Called statistical interaction between 
    $X_1$ and $X_2$.
2. If there is no statistical interaction, we can 
consider $X_2$ as a potential confounding variable.
    + $X_2$ could change the association
    between $Y$ and $X_1$ when it is included in our MLR model.

## 

* We'll be using the data on low birthweight babies to illustrate ideas.

\scriptsize

```{r, warning=FALSE, message=FALSE}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
head(lbwt)
```

## Statistical Interaction

- Easiest when $X_2$ is binary; i.e., takes values of 0 or 1. 
- $X_2$ _modifies_ the effect of $X_1$ on $Y$ if the 
slope of the regression line
of $Y$ on $X_1$ differs in the $X_2=0$ and $X_2=1$ subgroups. 
- Illustrate with the variable `toxemia` in the low birthweight babies dataset.
    - `toxemia=1` if the mother is toxic during pregnancy and 0 otherwise
    - If we stratify the analysis by toxemia and find different 
slopes for gestational age in the two toxemia groups, there is statistical interaction 
between gestational age and toxemia.




## MLR Model with Statistical Interaction 

* Consider the MLR model with _linear predictor_: 
  $\mu_{Y|X_1,X_2} = \alpha + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2,$
where 
    * $Y$ is head circumference, `headcirc`.
    * $X_1$ is gestational age, `gestage`.
    * $X_2$ is `toxemia` (1 is toxic, 0 is not)
    * $X_1 \times X_2$ is the statistical interaction between gestational age and toxemia.
    * $\beta_1$, $\beta_2$, and $\beta_3$ are the corresponding regression coefficients:
        + $\beta_1$ is the gestational-age ***main effect*** 
        + $\beta_2$ is the toxemia ***main effect***
        + $\beta_3$ is the gestational-age-by-toxemia ***interaction effect***
    
    
## Separate Lines
- Our linear predictor is $$\mu_{Y|X_1,X_2} = \alpha + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2,$$
- This model allows separate lines for the two toxemia groups.
    - Line for no-toxemia group ($X_2= 0$): $\alpha + \beta_1 X_1$
        - intercept $\alpha$ and slope $\beta_1$ for `gestage`.
    - Line for toxemia group ($X_2= 1$): $\alpha + \beta_1 X_1 + \beta_2 + \beta_3 X_1$
        - intercept $\alpha+\beta_2$ and slope $\beta_1+\beta_3$ for `gestage`.
- Focusing on the slopes, we see that the difference between `gestage` slopes for the two toxemia groups is $\beta_3$. 
    +  $\beta_3=0$ implies that the slopes are the same in the two groups; 
    + i.e., toxemia doesn't modify the effect of gestational age on head circumference.
    
## 

- To assess the evidence for statistical interaction
between toxemia and gestational age,
we test the hypotheses $$H_0:\beta_3=0 
\;\; vs. \;\; H_a:\beta_3 \not= 0.$$ 

+ If $H_0$ is retained, we conclude that there is insufficient
    statistical evidence that toxemia modifies the effect of 
    gestational age on head circumference.
    
##

- If we retain the no-interaction of hypothesis $H_0:\beta_3=0$,
our linear predictor becomes
$$\mu_{Y|X_1,X_2} = \alpha + \beta_1 X_1 + \beta_2 X_2$$

- This model allows separate lines for the two toxemia groups but with the same slope for `gestage`:
    - Line for no-toxemia group ($X_2= 0$): $\alpha + \beta_1 X_1$ 
        - intercept $\alpha$ and slope $\beta_1$ for `gestage`.
    - Line for toxemia group ($X_2= 1$): $\alpha + \beta_1 X_1 + \beta_2$.
        - intercept $\alpha+\beta_2$ and slope $\beta_1$ for `gestage`.

- No interaction between `gestage` and `toxemia` means that each toxemia group has its own line with different intercepts, but with the same slope for `gestage` 

## Fitted Model

\small

- Let's fit the MLR model with interaction between gestational age and toxemia:

\footnotesize

```{r}
lfit <- lm(headcirc ~ gestage+toxemia+gestage:toxemia,data=lbwt)
summary(lfit)$coefficients
```

\small

- From the row of output for `gestage:toxemia`, we see that the $t$-test of $H_0:\beta_3=0$ _vs._  $H_a: \beta_3 \not=0$ 
retains $H_0$ at the 5% level ($p=0.78$).
- No statistical evidence that toxemia modifies the effect of gestational age on head circumference.
- However, toxemia may still be a confounding variable.

## Software Notes

* In the model formula      
`headcirc ~ gestage + toxemia + gestage:toxemia`
    + The interaction term between `gestage` and `toxemia` is indicated by `gestage:toxemia`.
    +  The main effect terms are indicated by `gestage` and `toxemia`.
    
* In the model summary:
    + Information about the slope $\beta_3$ for the interaction term is in the row labelled `gestage:toxemia`.
    + Information about the slopes $\beta_1$ and $\beta_2$ for the
    main effect terms are in the rows labelled `gestage` and `toxemia`, respectively.
    

## Statistical Interaction More Generally

* Interaction terms appear as products of main-effect terms.
    + E.G. in the MLR with linear predictor 
    $\mu_{Y|X_1, X_2} = \alpha + \beta_1 X_1 + \beta_2X_2 + \beta_3 X_1X_2$,
    the interaction term $X_1 X_2$ is a product of the main
effect terms $X_1$ and $X_2$.  
* Interaction means that the slopes for $X_1$ can
depend on the value of the modifying variable $X_2$.
* E.G. Say $X_2$ is a continuous variable taking
on values between 5 and 10. 
* Focus on two values $X_2=5$ and $X_2=6$ one unit
apart:  
    + Line for $X_2 = 5$ group is $\alpha + \beta_1 X_1 + \beta_2 5 + \beta_3 X_1 5$  
         + intercept: $\alpha+\beta_2 x_2 = \alpha + \beta_2 5$, and 
         + slope for $X_1$: $\beta_1 + \beta_3 x_2 = \beta_1 + \beta_3 5$  
    + Line for $X_2 = 6$ group is $\alpha + \beta_1 X_1 + \beta_2 6 + \beta_3 X_1 6$  
         + intercept: $\alpha+\beta_2 x_2 = \alpha + \beta_2 6$, and 
         + slope for $X_1$: $\beta_1+\beta_3 x_2 = \beta_1 + \beta_3 6$  
* Difference in slopes for $X_1$ for the two groups is $\beta_3 6 - \beta_3 5 = \beta_3$.
    
##

- In general, we interpret the slope $\beta_3$ 
for the interaction term $X_1 X_2$ as the difference between 
the slopes for $X_1$ in two groups that are defined by a one-unit change in $X_2$.

- If $\beta_3=0$, the slopes for $X_1$ are the same.
    - Therefore, $X_2$ does **not** modify the effect of $X_1$ on $Y$.
    
- To assess the statistical interaction of $X_1$ and $X_2$, test the hypotheses that $H_0:\beta_3=0$ _vs._ $H_a:\beta_3 \not= 0$.

- This is equivalent to testing whether or not $X_2$ modifies
the effect of $X_1$ on $Y$.


## Example: Interaction of Gestational Age and Birthweight

\scriptsize

```{r}
lfit <- lm(headcirc ~ gestage+birthwt+gestage:birthwt,data=lbwt)
summary(lfit)$coefficients
```

\small

- $\hat \beta_3=-.000314$ is the estimated difference between the slopes for `gestage`, in 2 groups defined by a one-unit change in `birthwt`.

- E.G. Define
two groups: one for babies with the median `birthwt` of 1155g and another
for babies with `birthwt` 1156g.

    + In babies with `birthwt` 1156g, the slope for `gestage` 
    is estimated to be 0.000314 **less** than in babies with `birthwt` 1155g (since $\hat \beta_3$ is negative).

## What is the effect of `gestage` on `headcirc` in babies with a `birthwt` of 1156g?

\small

- The linear predictor or population mean is 
$$\mu_{Y|X_1, X_2} = \alpha + \beta_1 X_1 + \beta_2X_2 + \beta_3 X_1X_2.$$
- In babies with `birthwt` of $x_2=1156$g this simplifies to
\begin{eqnarray*}
\mu_{Y|X_1, 1156}&=& \alpha + \beta_1 X_1 + \beta_2 \, 1156 + \beta_3 X_1 \, 1156 \\
& = & \alpha + \beta_2 \, 1156 + \beta_1 X_1 + \beta_3 X_1 \, 1156 \\ 
& = & \alpha + \beta_2 \, 1156 + (\beta_1 + \beta_3 \, 1156) X_1
\end{eqnarray*}

* The slope for `gestage` ($X_1$) in babies with 
a `birthwt` of
$x_2=1156$g is therefore $\beta_1 + \beta_3 \, 1156$


* **In babies with a `birthwt` of $1156$g**, we estimate that the effect of `gestage` on `headcirc` is 
$$\hat{\beta}_1 + \hat{\beta}_3 \, 1156 = 0.787-0.000314 \times 1156 = 0.424;$$
i.e., a one-week increase in `gestage` is associated with an
estimated $0.424$cm
increase in `headcirc`


## Does `birthwt` modify the effect of `gestage` on `headcirc`?

* To address this question, let's test for statistical interaction
between `birthwt` and `gestage` at the 5\% level.

\scriptsize

```{r}
summary(lfit)$coefficients
```

\normalsize 

* Compare the $p$-value for the interaction term to the level 0.05.
    - Since the $p$-value is 0.09, we retain the null hypothesis
    of no interaction.

* Conclude that `birthwt` does **not** modify the effect of 
`gestage` on `headcirc`.

* Though `birthwt` is not an effect modifier, it could still confound the association between `gestage` and `headcirc` $\ldots$



# Confounding Variables 

## Changing the role of `birthwt`

- Previously, we had been thinking of `birthwt` as a potential
modifier of the effect of `gestage` on `headcirc`. 

- Having declared `birthwt` not to be an effect modifier, we may now consider it as a potential **confounder** of the association
between `gestage` and `headcirc`. 
     
- Look at 
the relationship between head circumference, $Y$, and gestational age, $X_1$, adjusted for birth weight, $X_2$.
- If analyses of an association between $Y$ and $X_1$
with and without $X_2$ give
"meaningfully different" estimates of the slope for $X_1$,
then $X_2$ is declared to be a confounder.
- The definition of "meaningfully different" depends on the context.
- One rule-of-thumb: If the estimated slope $\hat{\beta}_1$ changes by more than 10\%
when $X_2$ is excluded, then $X_2$ is a confounder
(Budtz-Jorgensen et al. 2007, Annals of Epidemiology).
    + **Note**: No statistical test for confounding is involved.


## Example: `birthwt` as confounder

\footnotesize

```{r}
coefficients(lm(headcirc ~ gestage + birthwt,data=lbwt))
coefficients(lm(headcirc ~ gestage,data=lbwt))
```

\small

- Measure change in the estimate of $\beta_1$ relative to the fitted 
model that _includes the confounding variable_, as this is 
considered the safer estimate of the true effect.

    - Specifically, look at change as \% of this estimate.

- The percent change in $\hat{\beta}_1$ is
$|0.445-0.780|/|0.445| \times 100\% = 75\%$. 
    + As this is larger than 10%,  `birthwt` would be considered a confounder by the rule-of-thumb.
    
## Interpreting slopes when `birthwt` is a confounder

\scriptsize

```{r}
coefficients(lm(headcirc ~ gestage + birthwt,data=lbwt))
```

\normalsize

* The interpretation of the slope for `gestage` is:
    + For a given birth-weight, 
    a one-week increase in the gestational age is associated
    with an estimated 0.449cm increase in the head circumference.
    
* The interpretation of the slope for `birthwt` is:
    + For a given gestational age, 
    a one-gram increase in birth weight is associated
    with an estimated 0.005cm increase in the head circumference.

  