---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 11 : Inference for Two Means'
author: "Jinko Graham"
date: '2018-09-13'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Example Data: Low Birthweight Infants

* Data on 100 infants born with birth weight less than
1500g. 
    + Variables are: head circumference (cm), birth length (cm), gestational
    age (wks), birth weight (g), mother's age (yrs), and 
    toxemia (1=high blood pressure during pregnancy, 0=not)
    
\scriptsize
```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
head(lbwt) 
```

 
## Gestational Age by Toxemia: Boxplots

* Explore differences graphically with boxplots:

\scriptsize

```{r,fig.height=2,fig.width=3,message=FALSE,warning=FALSE}
library(dplyr)
lbwt <- mutate(lbwt,toxCateg = factor(toxemia))
library(ggplot2)
ggplot(lbwt,aes(x=toxCateg,y=gestage)) +
         labs(x="Toxemia",y="Gestational Age") + geom_boxplot()
```

## Software Notes

* Use `mutate()` to make changes to a dataset.
    + In the example, we added a new variable called `toxCateg` to 
the `lbwt` dataset.
    + Whereas `toxemia` is a numeric variable with values 0 and 1,
    `toxCateg` is an explicitly categorical (or `factor`) variable,
    still having values 0 and 1.
    + In
    R, categorical variables are known as "factors".
* We have used `ggplot()` to do the boxplots of 
`gestage` by toxemia categories (`toxCateg`). 
    + For a boxplot, the call to aesthetic argument, `aes`, must specify an x-variable that is a factor.
    + `labs()` specifies the x- and y-axis labels.
    + `geom_boxplot()` adds the boxplots.
* Note: Data "wrangling", processing and 
graphics take up 90\% of an analyst's time (that's why data scientists get paid
big bucks). Not to worry though because we will have templates to work from in this class.

## Gestational Age by Toxemia: Histograms

\scriptsize

```{r,fig.height=2.5,fig.width=3.5,message=FALSE,warning=FALSE}
ggplot(lbwt,aes(x=gestage,fill=toxCateg)) +
         labs(x="Gestational Age") + geom_histogram(binwidth=1)
```

## Software Note

* For a histogram, setting the aesthetic argument `fill` to `fill=toxCateg` specifies that the bars
of the histogram are to be filled with different colors
for the different categories of `toxCateg`. 
    + Gives the impression of histograms stacked one upon the 
    other.
    

## Gestational Age by Toxemia: Summary Statistics

* The sample means and SDs of gestational age for
each toxemia category are summarized below. 

\footnotesize
```{r,warning=FALSE,message=FALSE}
library(dplyr) 
lbwt %>% 
  group_by(toxCateg) %>%
  summarize(mean=mean(gestage),sd=sd(gestage))
```

## Software Notes

\footnotesize
```{r,warning=FALSE,message=FALSE,eval=FALSE}
lbwt %>% 
  group_by(toxCateg) %>%
  summarize(mean=mean(gestage),sd=sd(gestage))
```

\normalsize

* The code that produced the summaries should be read as:
    + Start with the `lbwt` dataset,
    + Group observations in this dataset by the variable `toxCateg`; i.e., partition
    the observations into groups defined 
    by the categories of `toxCateg`, and
    + Summarize the sample mean and SD of the `gestage` 
    variable within each group.
    + The "forward pipe" `%>%` is the "glue" that connects
    these steps together.
    
* This short [video](https://www.youtube.com/watch?v=qxF_W0R4tfs) (time 5:22) gives a nice explainer on the forward pipe. 

<!--
The `dplyr` package
is for manipulating dataframes. Note its "5 verbs": `select`, `filter`, `arrange`, `mutate` and `summarize`.
--> 
    
## `t.test()` for Gestational Age by Toxemia

\scriptsize

```{r}
t.test(gestage ~ toxCateg,data=lbwt,conf.level=0.90)
```

\normalsize

* In R, a "formula" such as `gestage ~ toxCateg` says
to think of `gestage` as a function of `toxCateg`.
    + We will make extensive use of R formulas, later, 
    when we study regression.
* The argument `conf.level` sets the level, or 
coverage probability, $C$, of the CI.

## Getting the $p$-value

\scriptsize

```{r, echo=FALSE}
t.test(gestage ~ toxCateg,data=lbwt,conf.level=0.90)
```

\normalsize

* The output tells us that, under $H_0$, the test statistic $T$'s distribution is approximated by a t distribution on $\nu=31.465$ degrees of freedom.
* The software compares the observed value -4.4745
of the test statistic
to this reference distribution, to get a $p$-value of .00009366115.
