---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 11 : Inference for Two Means'
author: "Jinko Graham"
date: '2018-09-13'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Comparison of Two Means (Chapter 11)

Context

* We have measurements sampled from two populations. 
* Want to make inference about the difference between the populations.
* In particular, interested in the difference
between the two population means, denoted $\mu_1$ and $\mu_2$.

Notation

* Let $x_{11},\ldots,x_{1n_1}$ denote a sample from the first population
and $x_{21},\ldots,x_{2n_2}$ denote a sample from the second.
* The sample averages $\bar{x}_1$ and $\bar{x}_2$ estimate the population means
$\mu_1$ and $\mu_2$, respectively; 
so $\bar{x}_1 - \bar{x}_2$ estimates $\mu_1 - \mu_2$.
* We're interested in confidence intervals for $\mu_1-\mu_2$
and tests of $H_0: \mu_1 - \mu_2 = 0$
vs. $H_a: \mu_1 - \mu_2 \not= 0$.


## Example: Low Birthweight Infants

* Data on 100 infants with birth weight less than
1500g. 
    + Variables are: head circumference (cm), birth length (cm), gestational
    age (wks), birth weight (g), mother's age (yrs), and mother's status for
    toxemia (1=high blood pressure during pregnancy, 0=not)
    
* Compare the distribution of variables such as age and birth weight in moms with to mom's without toxemia 

* The first few rows of the data set are as follows:

\scriptsize
```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
head(lbwt) 
```

## Gestational Age by Toxemia

* Question: Does the distribution of 
gestational age differ in moms with toxemia vs. moms without toxemia? 

* Explore differences by toxemia status graphically, using boxplots (below) and histograms (next slide) in the sample.

\scriptsize

```{r,fig.height=2,fig.width=3,message=FALSE,warning=FALSE}
library(dplyr)
lbwt <- mutate(lbwt,toxCateg = factor(toxemia))
library(ggplot2)
ggplot(lbwt,aes(x=toxCateg,y=gestage)) +
         labs(x="Toxemia",y="Gestational Age") + geom_boxplot()
```

## Gestational Age by Toxemia: Histograms

\scriptsize

```{r,fig.height=2.5,fig.width=3.5,message=FALSE,warning=FALSE}
ggplot(lbwt,aes(x=gestage,fill=toxCateg)) +
         labs(x="Gestational Age") + geom_histogram(binwidth=1)
```

## Gestational Age by Toxemia: Summary Statistics

* The sample means and SDs of gestational age for
each toxemia category are summarized below. 

\footnotesize
```{r,warning=FALSE,message=FALSE}
library(dplyr) 
lbwt %>% 
  group_by(toxCateg) %>%
  summarize(mean=mean(gestage),sd=sd(gestage))
```

\normalsize

* The sample means of the gestational ages differ between the toxemia groups, but the sample SDs look the same.

* Could the difference in the sample means 
be due to chance?

## Outline of Inference Approach

* Same basic approach to inference as in the one-sample problem:
    + Inference is based on the sampling distribution of the statistic $\bar{X}_1 - \bar{X}_2$
* Transform $\bar{X}_1 - \bar{X}_2$ to a _pivotal quantity_,
    $Z$, if population SDs $\sigma_1$ and $\sigma_2$ are known.
*  When $\sigma$'s are unknown, as is typically the case, we substitute estimates to obtain
    a _pivotal quantity_ $T$.
* Confidence intervals and hypothesis tests follow from the sampling 
distribution of $T$.
* Note: We omit the following topics in this course:
    + Paired-samples $t$-test (Section 11.1 of text) 
    + Two-sample $t$-test assuming equal SDs (Section 11.2.1)


## Sampling Distribution of $\bar{X}_1 - \bar{X}_2$

* We have simple random samples (SRSs) of 
size $n_1$ for group 1 and $n_2$ for group 2. 
* These samples are independent.
* Then the distribution of 
$\bar{X}_1-\bar{X}_2$ has
    + mean $\mu_1-\mu_2$ and
    + SD $\sqrt{\sigma_1^2/n_1 + \sigma_2^2/n_2}$, 
    where $\sigma_1$ and $\sigma_2$ are the population SDs for group 1 and group 2, respectively.
* If the sample sizes are large enough,
the CLT tells us that the shape of this distribution 
is approximately normal.


## $Z$ Transformation

* "Standardizing" a random variable by subtracting its population mean 
and dividing by its population SD gives a random variable
with mean 0 \& SD 1.
* For normal random variables, the transformation 
does not change the distribution; the standardized random variable is still normally distributed.
* Conclude that if $\bar{X}_1-\bar{X}_2$ is 
approximately normal with mean $\mu_1-\mu_2$ and
SD $\sqrt{\sigma_1^2/n_1 + \sigma_2^2/n_2}$, then
$$
Z = \frac{ (\bar{X}_1-\bar{X}_2) - (\mu_1-\mu_2)}{\sqrt{\sigma_1^2/n_1 + \sigma_2^2/n_2}}
\sim N(0,1),
$$
where $\sim$ means "is distributed as".

## $T$ Transformation

* Inserting sample SDs $s_1$ and $s_2$ for the parameters 
$\sigma_1$ and  $\sigma_2$ in $Z$ gives
$$
T = \frac{(\bar{X}_1-\bar{X}_2)-(\mu_1-\mu_2)}{\sqrt{s_1^2/n_1 + s_2^2/n_2}}
$$
* What is the distribution of $T$? 
    + Turns out we can 
    approximate it by a $t$-distribution with $\nu$ df. 
    + We won't ever use the formula for $\nu$ but if you're curious it is given on page 270 of the text.
* Instead, computer software such as R automatically calculates $\nu$ for us,
as shown next.



## Illustration with Gestational Age and Toxemia

* The following is the output of the `t.test()` function
for these data (see the R demo for details):

\footnotesize

```{r}
t.test(gestage ~ toxCateg,data=lbwt)
```

\normalsize 

* The df is $\nu = 31.465$.
* We can also see that the software computed a 95% confidence
interval and p-value. More on these in a few slides.

## Confidence Intervals for $\mu_1-\mu_2$

* The level-$C$ CI for $\mu_1-\mu_2$ is of the form 
$$\mbox{estimate }  \pm  \mbox{margin of error}$$
* The estimate is $\bar{x}_1 - \bar{x}_2$
* The margin of error is $t^* \times SE$ where 
    + $t^*$ is the upper $(1-C)/2$ critical value of 
    the $t$-distribution with $\nu$ df, and
    + $SE$ is the **estimated SD** of $\bar{x}_1 - \bar{x}_2$:  $\sqrt{s_1^2/n_1 + s_2^2/n_2}$

## Application to Gestational Age and Toxemia

* Calculate a 90% CI for the difference between 
mean gestational age in the toxemia and non-toxemia groups.
* The relevant sample statistics to four digits are as follows:

group | sample mean ($\bar{x}$) | sample SD ($s$) | sample size ($n$) 
----- | ----------- | --------- | -----------
1: non-toxemia | 28.35 | 2.321 | 79 
2: toxemia | 30.90 | 2.322 | 21

* Estimate is $\bar{x}_1 - \bar{x}_2 = 28.35-30.90 = - 2.55$

* Margin of error is $t^* \times SE$ where
    + computer software calculates a critical value of
    $t^* = 1.695$ (see the R demo).
    + the SE is $\sqrt{s_1^2/n_1 + s_2^2/n_2} =
    \sqrt{2.321^2/79 + 2.322^2/21} = 0.570$.
    + Hence the margin of error is $1.696\times 0.570 = 0.966$.

* CI is $(-2.55 - 0.966, -2.55+0.966) = (-3.516,-1.584)$.

    
## Interpretation

* The 90% CI is approximately $(-3.5,-1.6)$.
* The text suggests an interpretation such as: 
    + "90% of intervals constructed in this way cover the 
true difference between mean gestational age in 
the **non-toxemia** and **toxemia** 
groups."
* Another common style of interpretation is:
    + "We are 90% confident that the true difference between 
    mean gestational age in the **non-toxemia** and **toxemia** groups
    is between $-3.5$ and $-1.6$."
* Or, we might find it more natural to switch the order of 
the groups, which would switch the sign of the difference:
    + "We are 90% confident that the true difference between 
    mean gestational age in the **toxemia** and **non-toxemia**
    groups is between $1.6$ and $3.5$"



## Hypothesis Test

* For the null hypothesis $H_0: \mu_1-\mu_2=0$ of no difference between the groups, 
the formula for the observed $t$-statistic is
$$
t = \frac{(\bar{x}_1-\bar{x}_2) - 0}{\sqrt{s_1^2/n_1 + s_2^2/n_2}}
$$
* Observed values of $t$ that are extreme
in the sense of being more compatible with $H_a$ are taken as evidence 
against $H_0: \mu_1-\mu_2=0$.
* The $p$-value is the chance of a value that is as 
or more extreme than what we observed, under $H_0$.
* Using the same logic as in the one-sample problem (Ch 10), and taking $T$ to be a
random variable with a $t$ distribution on $\nu$ df, we have:
    + for $H_a: \mu_1 -\mu_2 \not= 0$,
    the $p$-value is $p = 2 P( T\geq |t|)$,
    + for $H_a: \mu_1-\mu_2 > 0$, 
    the $p$-value is $p=P(T \geq t)$ and
    + for $H_a: \mu_1-\mu_2 < 0$, the $p$-value is $p=P(T \leq t)$.

## Application to Gestational Age and Toxemia

* We can re-use the calculations from the CI example (page 13), for the difference 
in sample means ($\bar{x}_1 - \bar{x}_2 = -2.55$) and the SE of the difference
($\sqrt{s_1^2/n_1 + s_2^2/n_2} = 0.570$):

$$
t = \frac{(\bar{x}_1-\bar{x}_2) - 0}{\sqrt{s_1^2/n_1 + s_2^2/n_2}} = \frac{-2.55}{0.570} = -4.474.
$$

* Computer software (R demo) returns 
a $p$-value of $9.4\times 10^{-5}$, or 0.000094,
for a 2-sided
alternative hypothesis.

## Interpretation

* Small $p$-values (e.g. $<.05$) 
are taken as evidence against the null hypothesis.

* Our $p$-value of $9.4\times 10^{-5}$ is very small.

* If we had set a level of $\alpha=10$% for the test, we'd 
declare that: 
\begin{quote}
"We reject the null hypothesis that the mean 
    gestational age is the same in the toxemia and 
    non-toxemia groups at the 10\% level."
\end{quote}

* If we hadn't set a level of the test in advance, we 
might report our results as:
  \begin{quote}  
    "There is strong evidence that 
    the mean gestational age is different in the 
    toxemia and non-toxemia groups ($p<0.001$)."
  \end{quote}

  
## Cause and Effect

* Our two-sample t test has revealed that toxemia and 
gestational age are **associated**.
    + The distribution of gestational age is different 
    in the two toxemia groups (different means, lower in toxemia group).
* But, an association does **not** mean that toxemia has a causal effect on gestational age.
    + It could be that gestational age affects toxemia.
    + Or, there could be a hidden _confounding variable_ 
    that affects both toxemia and gestational age, that
    accounts for their association.
    
    \bigskip
    (More on confounding later, when we study multiple regression.)


## Relationship Between Confidence Intervals and Tests

* In the low birth weight example, the 90% CI does not cover zero,
and the hypothesis test of $H_0: \mu_1-\mu_2=0$
_vs_ $H_a: \mu_1-\mu_2 \not= 0$ rejects the 
null hypothesis at the 10% level.
* Conversely, when a 90% CI **does** 
cover zero, the corresponding test against a 
two-sided alternative
will **retain** the null hypothesis at the 10\% level.
* This is a general property of tests of a population parameter $\theta$.
    + A level-$\alpha$
test of $H_0: \theta = \theta_0$ 
    _versus_ $H_a: \theta \not= \theta_0$
    retains the null hypothesis if and only if
    the level $(1-\alpha)\times 100\%$ CI 
    covers $\theta_0$.

## Assumptions

* The assumptions for inference are that:
    + The data are two SRSs from the two parent 
    populations (e.g. mothers with toxemia and mothers without toxemia).
    + Either 
        1. the data measurements in 
    the parent populations are normally distributed 
    with mean $\mu_i$ and sd $\sigma_i$, written
    $N(\mu_i,\sigma_i)$, or
        2. the sample size
    $n=n_1+n_2$ is "large" enough to rely on the CLT for the sample means $\bar{X}_1$ and $\bar{X}_2$.
* Guidelines for $n$ (from the text _Basic Practice of Statistics_
by D. Moore):
    + For $n < 15$, use the $t$-based CI and hypothesis test
    if the data look to be approximately normally distributed.
    + For $15 \leq n < 40$, use the $t$-based CI and hypothesis test,
    _except_ in the presence 
    of outliers or strong skewness in the data distribution.
    + For large samples ($n\geq 40$), you can use the $t$-based CI and hypothesis test, 
    even for clearly skewed distributions (because of the CLT).

## Checking the assumptions for the low-birth-weight example

* There were $n=100$ babies in this data set.
* According to the rules-of-thumb on the previous slide, we can use the $t$-based 
CI and hypothesis test even if the 
population distributions are skewed.

## Summary

* Inference for the difference 
between two population
means is based on either $Z$ (SDs known) or 
$T$ (SDs unknown).
* Confidence intervals are of the form estimate $\pm$ margin of error
    + the margin of error is a critical value times SE
* To test the null hypothesis $H_0:\mu_1 - \mu_2 = 0$  against an alternative
$H_a$ we compute a test statistic $t$ (or $z$ if SDs known) and $p$-value
    + can compare $p$-value to a significance level $\alpha$ to obtain a test
* Inference is considered reliable when the parent populations are normal, or when rules of thumb for sample sizes are satisfied.