---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 11 : Inference for Two Means'
author: "Jinko Graham"
date: '2018-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Example Data: Low Birthweight Infants

* Data on 100 infants born with birth weight less than
1500g. 
    + Variables are: head circumference (cm), birth length (cm), gestational
    age (wks), birth weight (g), mother's age (yrs), and 
    toxemia (1=high blood pressure during pregnancy, 0=not)
    
\scriptsize
```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
head(lbwt) 
```

## Software Notes

* Recall that `read.csv()` reads comma-separated-values (CSV) files into R.
<!--    + By default, reads files from the "working" directory
    in which R is running (e.g., the project directory 
    of your RStudio project, or the folder of
    your Jupyter notebook), but
    it can read files from URLs too.
    + The `url()` function takes a quoted URL as input and
    returns an object that `read.csv()` can use to fetch
    the file.
-->
* A more general data-reading function is `read.table()`,
which can read many different plain-text formats; e.g., 

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt2 <- read.table(uu,header=TRUE,sep=",")
head(lbwt2,n=3) 
```

\normalsize

* The `header=TRUE` argument tells `read.table` that
the first row of the data file contains the variable
names.
* The `sep` argument specifies the delimiter that separates values in the data file; in this case it is a comma.
    
## Gestational Age by Toxemia: Boxplots

* Explore differences graphically with boxplots:

\scriptsize

```{r,fig.height=2,fig.width=3,message=FALSE,warning=FALSE}
library(dplyr)
lbwt <- mutate(lbwt,toxCateg = factor(toxemia))
library(ggplot2)
ggplot(lbwt,aes(x=toxCateg,y=gestage)) +
         labs(x="Toxemia",y="Gestational Age") + geom_boxplot()
```

## Software Notes

* Use `mutate()` to make changes to a dataset.
    + In the example, we added a new variable called `toxCateg` to 
the `lbwt` dataset.
    + Whereas `toxemia` is a numeric variable with values 0 and 1,
    `toxCateg` is an explicitly categorical (or `factor`) variable,
    still having values 0 and 1.
    + In
    R, categorical variables are known as "factors".
* We have used `ggplot()` to do the boxplots of 
`gestage` by toxemia categories (`toxCateg`). 
    + For a boxplot, the call to aesthetic argument, `aes`, must specify an x-variable that is a factor.
    + `labs()` specifies the x- and y-axis labels.
    + `geom_boxplot()` adds the boxplots.
* Note: **Don't panic!** "Data wrangling" and 
statistical graphics in R can be
difficult (that's why data scientists get paid
big bucks), but we will have templates to work from in this class.

## Gestational Age by Toxemia: Histograms

\scriptsize

```{r,fig.height=2.5,fig.width=3.5,message=FALSE,warning=FALSE}
ggplot(lbwt,aes(x=gestage,fill=toxCateg)) +
         labs(x="Gestational Age") + geom_histogram(binwidth=1)
```

## Software Note

* For a histogram, setting the aesthetic argument `fill` to `fill=toxCateg` specifies that the bars
of the histogram are to be filled with different colors
for the different categories of `toxCateg`. 
    + Gives the impression of histograms stacked one upon the 
    other.
    

## Gestational Age by Toxemia: Summary Statistics

* The sample means and SDs of gestational age for
each toxemia category are summarized below. 

\footnotesize
```{r,warning=FALSE,message=FALSE}
library(dplyr) 
lbwt %>% 
  group_by(toxCateg) %>%
  summarize(mean=mean(gestage),sd=sd(gestage))
```

## Software Notes

\footnotesize
```{r,warning=FALSE,message=FALSE,eval=FALSE}
lbwt %>% 
  group_by(toxCateg) %>%
  summarize(mean=mean(gestage),sd=sd(gestage))
```

\normalsize

* The code that produced the summaries should be read as:
    + Start with the `lbwt` dataset,
    + Group observations in this dataset by the variable `toxCateg`; i.e., partition
    the observations into groups defined 
    by the categories of `toxCateg`, and
    + Summarize the sample mean and SD of the `gestage` 
    variable within each group.
    + The "forward pipe" `%>%` is the "glue" that connects
    these steps together.
    
## `t.test()` for Gestational Age by Toxemia

\scriptsize

```{r}
t.test(gestage ~ toxCateg,data=lbwt,conf.level=0.90)
```

\normalsize

* In R, a "formula" such as `gestage ~ toxCateg` says
to model `gestage` as a function of `toxCateg`.
    + We will make extensive use of R formulas, later, 
    when we study regression.
* The argument `conf.level` sets the level, or 
coverage probability, of the CI.

## Using R to Calculate $p$-values

* In the lecture notes we calculated the t-statistic,
$t$, from sample means and SDs and found
the corresponding $p$-value, $2 P( T\geq |t|)$, for a 2-sided alternative hypothesis.
    + Here $T$ is a random variable having
    a t distribution on $\nu=31.465$ degrees of freedom.
* Used to look up p-values in 
statistical tables; now we use computer software:

\scriptsize

```{r}
tstat <- -4.4745; nu <- 31.465
2*pt(abs(tstat),df=nu,lower.tail=FALSE)
```
\normalsize

* the pvalue is .00009366115

* `pt()` with the `lower.tail=FALSE` argument calculates upper-tail probabilities for the t distribution.
* `abs(tstat)` is the absolute value of `tstat`.
* The `df` argument specifies the number of degrees of freedom.