---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 14 : Inference for Proportions'
author: "Jinko Graham"
date: '2018-09-13'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Test and CI 

* Can use `prop.test()` function to get  $p$-value and CI.


\scriptsize
```{r}
prop.test(c(166,122),c(8506,8102),conf.level=0.95,correct=FALSE)
```

\normalsize

## Arguments 

\bigskip

\scriptsize
`prop.test(c(166,122),c(8506,8102),conf.level=0.95,correct=FALSE)`

\normalsize

\bigskip

In the call above to `prop.test()`:
    
+ The first argument contains the numbers of "successes" (cancers) in the EP and placebo groups,
    respectively.

+ The 2nd argument contains the numbers of "trials" (women) in the EP and placebo groups, respectively. 
    
+ `conf.level` is the level or coverage probability $C$ of the interval 
    (default = 0.95).
    
+ `correct` specifies whether to apply a "continuity correction"
    that improves the statistical inference
     when the total size of the sample is small. 
    The default is `correct=TRUE`,
    but I set `correct=FALSE` to re-create
    the results from using the formulas in the text. 
    
## Output

\scriptsize
```{r, echo=FALSE}
prop.test(c(166,122),c(8506,8102),conf.level=0.95,correct=FALSE)
```

\normalsize

+ Mostly like the output of `t.test()`

+ `X-squared` is the square of the test statistic
    $Z$ that we have discussed. This has a chi-squared
    ($\chi^2$)
    distribution with one df. 

+ When the alternative hypothesis is two-sided, the p-value from the 
    $\chi^2$ test is equivalent
    to the p-value from the $Z$-test. 

