---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 14 : Inference for Proportions'
author: "Jinko Graham"
date: '2018-08-23'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Confidence Intervals

* Recall: WHI has 16,608 women aged 50 to 79 randomized 
to receive either estrogen plus progestin (EP; $n_1=8506$), or a placebo
($n_2=8102$). After five years, 166 or those in the EP group had developed 
invasive breast cancer, compared to 122 in the placebo group.

\scriptsize

```{r}
numCancer1 <- 166; n1 <- 8506 #EP group
p1hat <- numCancer1/n1
numCancer2 <- 122; n2 <- 8102 #Placebo group
p2hat <- numCancer2/n2

phatDiff <- p1hat - p2hat
phatDiff
```

* The difference in proportions is about 0.004

## 

\normalsize

* 95\% CI is estimate $\pm$ margin of error, where
    + estimate of $p_1-p_2$ is $\hat{p}_1 - \hat{p}_2 = 0.0044$ 
    + margin of error is a critical value times standard error of difference.
    
\scriptsize

```{r}

zstar <- qnorm((1-.95)/2,lower.tail=FALSE) #get critical value
se <- sqrt(p1hat*(1-p1hat)/n1 + p2hat*(1-p2hat)/n2) #get se
me <- zstar*se #get margin of error

CI <- c(phatDiff - me,phatDiff + me)
CI
```
\normalsize

The 95\% CI is about $(0.0005,0.008)$

## R Lookup Table 

CI for Difference of Proportions 

\scriptsize

* $n_1=8506$ women receive EP and $n_2=8102$ receive placebo. 
166 in EP group and 122 in placebo group get breast cancer.
Table to help understand R code:

\begin{tabular}{lcl}
R variable & Notation & Value \\ \cline{1-3} 
{\tt numCancer1}& -- & 166\\
{\tt n1} & $n_1$ & 8506 \\
{\tt p1hat} & $\hat{p}_1$ & 166/8506 \\
{\tt numCancer2} & -- & 122 \\
{\tt n2} & $n_2$ &  8102 \\
{\tt p2hat} & $\hat{p}_2$ & 122/8102 \\
{\tt phatDiff} & $\hat{p}_1 - \hat{p}_2$ & 0.0044 \\
{\tt zstar} & $z^*$ &  1.96 \\
{\tt se} & $\sqrt{\hat{p}_1 * ( 1- \hat{p}_1)/n_1 +  \hat{p}_2 * ( 1- \hat{p}_2)/n_2}$ & 0.00202 \\
{\tt CI} & $(\hat{p}_1 - \hat{p}_2)  \pm z^* * \sqrt{\hat{p}_1 * ( 1- \hat{p}_1)/n_1 +  \hat{p}_2 * ( 1- \hat{p}_2)/n_2}$  
&  (0.0005,\\ 
& & ~~~0.008)\\ 
\end{tabular}

## Test statistic 

* Under $H_0: p_1-p_2=0$ or $p_1=p_2=p$, and so both the EP and placebo populations have the same proportion $p$ of cancer cases.

* Calculating the test statistic requires the pooled-sample
estimate of $p$. 

\scriptsize

```{r}
phat <- (numCancer1+numCancer2)/(n1+n2) #pooled-sample estimate of p
se <- sqrt(phat*(1-phat)*(1/n1+1/n2)) #based on phat
z <- phatDiff/se
z
```

\normalsize

* The test statistic is about 2.2

## Hypothesis test

* For the hypothesis test of $H_0: p_1 - p_2 = 0$ vs. $H_a: p_1 - p_2 \not= 0$, we have: 

\scriptsize

```{r}
pval<-2*pnorm(abs(z),lower.tail=FALSE) 
pval
```

\normalsize
* The pvalue is about 0.03.

* We therefore reject $H_0$ at the 5\% level.
    + There is statistical evidence that women taking EP have a higher 
risk of invasive breast cancer than those taking the placebo.

## R Lookup Table

Testing $H_0: p_1 - p_2 = 0$ vs. 
$H_a: p_1 - p_2 \not= 0$.

\scriptsize

* $n_1=8506$ women receive EP and $n_2=8102$ receive placebo. 
166 in EP group and 122 in placebo group get breast cancer.

\begin{tabular}{lcl}
R variable & Notation & Value \\ \cline{1-3} 
{\tt numCancer1}& -- & 166\\
{\tt n1} & $n_1$ & 8506 \\
{\tt p1hat} & $\hat{p}_1$ & 166/8506 \\
{\tt numCancer2} & -- & 122 \\
{\tt n2} & $n_2$ &  8102 \\
{\tt p2hat} & $\hat{p}_2$ & 122/8102 \\
{\tt phatDiff} & $\hat{p}_1 - \hat{p}_2$ & 0.0044 \\
& & \\
& & \\
{\tt phat} & $\hat{p}$ & $\frac{166+122}{8506 + 8102}$\\
& & \\
{\tt se} & $\sqrt{\hat{p} * ( 1- \hat{p})*( 1/n_1 +  1/n_2)}$ & 0.00203 \\
& & \\
{\tt z} & $\frac{(\hat{p}_1 - \hat{p}_2)}{\sqrt{\hat{p} * ( 1- \hat{p})*( 1/n_1 +  1/n_2)}}$ & 2.2 \\
& & \\
{\tt pval} & $2* P(Z \geq |z|)$ & 0.03 \\ \cline{1-3}
\end{tabular}

## Test and CI using prop.test()

* Can also use `prop.test()` function to get  $p$-value and CI.

\scriptsize

```{r}
numCancer <- c(numCancer1,numCancer2)
n <- c(n1,n2)
prop.test(numCancer,n,conf.level=0.95,correct=FALSE)
```

## Software Notes

* Arguments to `prop.test()`:
    + The first two arguments are the numbers of successes (cancers)
    and number of trials (women), respectively.
    + `conf.level` is the confidence-interval level 
    (default = 0.95).
    + `correct` specifies whether to apply a "continuity correction"
    that improves the statistical inference
     when the total size of the sample is small. 
    The default is `correct=TRUE`,
    but I set `correct=FALSE` to re-create
    the results from using the formulas in the text. 
* Output:
    + Mostly like the output of `t.test()`
    + `X-squared` is the square of the test statistic
    $Z$ that we have discussed. This has a chi-squared
    ($\chi^2$)
    distribution with one df. 
    + When the alternative hypothesis is two-sided, the p-value from the 
    $\chi^2$ test is equivalent
    to the p-value from the $Z$-test. 

