---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapters 2 & 3: Data presentations and summaries'
author: "Jinko Graham"
date: '2018-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Data Presentation (Chapter 2)

## Types of Variables

* Categorical versus quantitative variables
* Categorical variables: variable has categories or classes
    + Nominal data: unordered categories or classes 
    (e.g., gender, cancer type).
    + Ordinal data: categories with a natural order 
    (e.g., cancer stage, injury severity score)
* Quantitative variables: numbers represent measurable quantities
    + Discrete variables: Restricted values, e.g. integers. Some examples included number of days in hospital, parity. 
    + Continuous data: No restriction on values (though in practice,
    the measuing device may impose restrictions). Some examples include blood pressure, height, weight.

## Tables

* Tables can be used to display the frequency distribution of a 
categorical variable 
* Example: Frequency distribution of gender among 
21,737 bladder cancer patients. Data from [Mungan et al. (2000)](https://www.ncbi.nlm.nih.gov/pubmed/10699612)

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/mung.csv")
Mungan <- read.csv(uu)
with(Mungan,table(Gender))
```

## Tables, cont.

* Joint frequency distribution of two categorical variables:
```{r}
with(Mungan,table(Gender,Cancer.Stage))
```

* More on analysis of two-way tables in Chapter 15.

## Tabulating Quantitative Variables

* Can tabulate quantitative variables after
``binning''
* Divide the range of possible values into bins
    + For example, with age data ranging from 15 to 30 years, 
    could create three 5-year bins for age
* Count the number of subjects in each bin
* Histograms are a graphical display of quantitative 
data that makes use of binning.
    + We will see an example histogram in two slides.

## Graphs

* Bar plots to display categorical variables or discrete variables with a limited 
number of possible values
* Histograms to display discrete variables with many
possible values and for continuous variables
* Bars above categories (or bins) indicate number of observations in that category (or bin)

## Example Histogram

* Data on the numbers of hysterectomies performed by 15 male Swiss doctors:    
\footnotesize `20 25 25 27 28 31 33 34 36 37 44 50 59 85 86`

```{r,results='asis',fig.height=2.3,fig.width=3.4,message=FALSE,warning=FALSE}
hyst <- data.frame(numHyster = c(20,25,25,27,28,31,33,34,
                                 36,37,44,50,59,85,86))
library(ggplot2)
ggplot(hyst,aes(x=numHyster)) + geom_histogram(binwidth=10)
```

\small

* Observations far from the others are called outliers

## Notes on Histograms

* The purpose is a graphical representation of a distribution. 
    + The details of how the picture are drawn are not that
    important.
    + But you may find yourself wondering ...
* The bins in the previous example were of width 10.
    + It looks like they were set at 15 to 25, 25 to 35,
    etc.
* Which bin do the 25's go in?
    + The default is in the 15-25 bin. Bins don't include
    their left end-point, but do include their right end-point
    
# Summary Statistics (Chapter 3)

## Summary Statistics Overview

* For quantitative variables, numerical summaries
are used to measure different aspects of a data distribution.
    + Mean and median measure centre (central tendancy) 
    of the distribution
    + Inter-quartile range and standard deviation measure 
    spread (dispersion)

* _Five-number summary_ to summarize a distribution: 
min, max, median,
1st and 3rd quartiles. Graphed with a boxplot (more later).

## Centre: The mean

* The sample mean is the ordinary arithmetic average of the observations in the sample.
    + Notation: Let $x_1,x_2,\ldots,x_n$ denote the observed values of a variable measured
on $n$ individuals. The sample mean, $\overline{x}$ is 
$$
\overline{x} = \frac{x_1+x_2+\ldots+x_n}{n} = \frac{1}{n}\sum_{i=1}^n x_i
$$
* Hysterectomy example data: 
```
20 25 25 27 28 31 33 34 36 37 44 50 59 85 86 
```
has sample mean 
$$
\frac{20+25+\ldots+86}{n}=41.3
$$

## Centre: The Median

* The sample median is the ``middle value'' of the variable 

\footnotesize

20, 25, 25, 27, 28, 31, 33, **34**, 36, 37, 44, 50, 59, 85, 86  

\normalsize

* The centre observation is the median, $M=34$.
* The median represents a ``typical'' observation.

## Spread: The Standard Deviation (SD) and Variance

* The SD measures the spread about the mean.
* The sample variance, $s^2$, can be viewed as an average (almost) of the squared 
deviations about the sample mean
* The sample SD, $s$, is the square root of the sample variance.
    +  Notation: $x_1,x_2,\ldots,x_n$  are observed measurements
    on $n$ individuals with sample mean $\overline{x}$. 
\begin{eqnarray*}
s^2 & = & \frac{1}{n-1}\sum_{i=1}^n (x_i-\overline{x})^2 \\
s & = & \sqrt{s^2} =   \sqrt{\frac{1}{n-1}\sum_{i=1}^n (x_i-\overline{x})^2}
\end{eqnarray*}

* Hysterectomy example: $s=20.6$ 

\scriptsize

## Spread: The Inter-Quartile Range (IQR)

* The first and third quartiles mark the first and third quarters of the observations
    + These are also called the 25th and 75th percentile
* For the hysterectomy data, R
calculates these quartiles to be 27.5 and 47, respectively.
    +  Details of how R calculates quartiles are unimportant to us. 
    
* The middle half of the data lies between. 
* The range of the middle half, or IQR, is 47-27.5=19.5.

## Boxplots

* The five-number summary is the minimum, 
maximum, median, 1st and 3rd quartiles; graphed with a boxplot:

\scriptsize

```{r,fig.height=3,fig.width=4}
with(hyst,boxplot(numHyster)) 
```

\small

* Box represents the IQR, thick horizontal line the median.
* Whiskers extend out to the min/max if these are within a certain distance from the box,
* However, **outliers** that are far from the box are plotted separately (as above)
* There are different ways to draw boxplots; details unimportant

## Further Examples

* Summary statistics and graphics can be very powerful.
* See, for example, the late Hans Rosling's work

http://www.ted.com/talks/hans_rosling_shows_the_best_stats_you_ve_ever_seen.html
