---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demos for Chapters 8-10: Review of Statistical Inference'
author: "Jinko Graham, Brad McNeney"
date: '2019-09-09'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Confidence Intervals (Chapter 9)

## Example

* Example from the text, page 223, summarizes data on plasma aluminum levels,
in $\mu g/l$, for $n=10$ infants receiving antacids that contain aluminum.
    + The sample mean of the plasma aluminum levels is $\bar{x} = 37.20 \mu g/l$, and the sample SD is $s=7.13$
* I can't access the data from the text, but I have
simulated a data set of $n=10$ subject with
similar properties for this demo.
    + Read these data into R:

\scriptsize

<!--
```{r}
set.seed(123)
plasmaAlu <- data.frame(alu = rnorm(n=10,mean=37.2,sd=7.13))
write.csv(file="pa.csv",plasmaAlu,row.names=FALSE)
```
\normalsize
-->
```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/pa.csv")
plasmaAlu <- read.csv(uu)
head(plasmaAlu)
```

## Simulated Data Summary Statistics

* Our simulated data are similar to the data from 
the text, but the sample mean and SD are different:

\small

```{r,message=FALSE,warning=FALSE}
library(dplyr)
summarize(plasmaAlu,mean(alu))
summarize(plasmaAlu,sd(alu))
```

## Software Notes 

* Recall from the chapter 2-3 demo that
`dplyr` is an add-on package for R that includes useful 
tools for manipulating datasets in R.
    + To use `dplyr` functions we must first load the 
    package with `library(dplyr)`.

* The package consists of 5 main "verbs" for manipulating 
a dataframe: 
1. `select()`: 	select columns
2. `filter()`: 	filter rows
3. `arrange()`: 	re-order or arrange rows
4. `mutate()`: 	create new columns
5. `summarize()`: 	summarise columns

* The `summarize()` function
    takes the dataframe as its first
    argument, and the summaries to compute as additional
    arguments.

## Do-It-Yourself CIs in R

* In R, the quantile function `qt()` for the t distribution can be used to find 
the quantile or critical value ($t^*$) for the CI when $\sigma$ is unknown.

    + When the argument `lower.tail=FALSE`, the first argument, `p`, is the upper-tail area to the right of the desired quantile, under the t distribution.  

\scriptsize

```{r}
tstar<-qt(p=(1-0.95)/2,df=9,lower.tail=FALSE)
tstar
```

\normalsize

* Once we have the quantile or critical value, we can calculate the 95% CI.

## 

* Use assignment operator `<-` to put the critical value into an R object called `tstar`.

* Then use `summarize()` function in `dplyr` package to get sample mean, sample sd and the number of observations, $n$. Assign these to an R object `mysummary`. 

* Use the saved objects to calculate lower and upper bounds of CI and put them into the R object `CI`.
   
\scriptsize

```{r,warning=FALSE}
tstar <- qt(p=(1-0.95)/2,df=9,lower.tail=FALSE)
mysummary <- plasmaAlu %>% 
  summarize(xbar=mean(alu),s=sd(alu),n=n()) %>%
  mutate(CI.lower=xbar-tstar*s/sqrt(n),
         CI.upper=xbar+tstar*s/sqrt(n))
mysummary
```

## Chaining Data Manipulations

* In the above code chunk we used the "forward pipe", `%>%`
to push the dataset from one data manipulation to another.
* Push the `plasmaAlu` dataframe through the pipe
to use as input to `summarize()`, then take the dataframe
output by summarize and push it through the pipe to 
use as input to `mutate()`.
    + Each data-processing step is a verb, and we 
    are to read the result like a sentence:
    Start with `plasmaAlu`, summarize it 
    (mean, SD and sample size), then mutate it to 
    add the lower and upper CI limits.
    + Notice that we do not need to explicitly specify the 
    input data to `summarize()` and `mutate()` when
    they receive these data through the forward pipe.

## CIs with the `t.test()` Function

* R's `t.test()` uses data to test hypotheses 
about a mean, or about differences between two means,
but the function also returns summary statistics and 
a CI.
* Set the level or coverage probability of the CI with the argument
    `conf.level` (default $C=0.95$):

\scriptsize

```{r}
with(plasmaAlu,t.test(alu,conf.level=0.95))
```



# Hypothesis Tests (Chapter 10)


## Example

* In the **population** of infants **not taking antacids**,
the mean plasma-aluminum levels are known to be $\mu_0 = 4.13$  $\mu$g/l.
* Want to assess whether the mean level $\mu$ in infants taking
antacids is the same as $\mu_0$; i.e. whether 
$$H_0: \mu=4.13.$$
* Our alternative hypothesis is $H_a:\mu \not= 4.13$; i.e.,the mean plasma-aluminum levels of infants taking
antacids is different from infants not taking antacids.


## Do-It-Yourself Hypothesis Test

* The $t$-statistic is
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}
$$
* We can implement this as follows using
the summaries in the object `mysummary` from
the CI demo:

\scriptsize

```{r}
mysummary
with(mysummary,(xbar - 4.13)/(s/sqrt(n)))
```
\normalsize

* The $p$-value is 
$2P(T \geq |15.63|)$ for $T$ with $10-1=9$ df:

\scriptsize
```{r}
2*pt(15.63,df=9,lower.tail=FALSE)
```

## Using the `t.test()` Function

* The `t.test()` function will also perform the test.
    + Specify the null hypothesis 
$H_0:\mu=4.13$ with the `mu` argument.
    + The function's default is to use the two-sided
    alternative hypothesis, which in this case is
    $\mu \not= 4.13$:


\scriptsize
```{r}
with(plasmaAlu,t.test(alu,mu=4.13))
```
