---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demos for Chapters 8-10: Review of Statistical Inference'
author: "Jinko Graham"
date: '2018-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Confidence Intervals (Chapter 9)

## Example

* Example from the text, page 223, summarizes data on plasma aluminum levels,
in $\mu g/l$, for $n=10$ infants receiving antacids that contain aluminum.
    + The sample mean of the plasma aluminum levels is $\bar{x} = 37.20 \mu g/l$, and the sample SD is $s=7.13$
* I can't access the data from the text, but I have
simulated a data set of $n=10$ subject with
similar properties for this demo.
    + Read these data into R from my website:

\scriptsize

<!--
```{r}
set.seed(123)
plasmaAlu <- data.frame(alu = rnorm(n=10,mean=37.2,sd=7.13))
write.csv(file="pa.csv",plasmaAlu,row.names=FALSE)
```
\normalsize
-->
```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/pa.csv")
plasmaAlu <- read.csv(uu)
head(plasmaAlu)
```

## Simulated Data Summary Statistics

* Our simulated data are similar to the data from 
the text, but the sample mean and SD are different:

\small

```{r,message=FALSE,warning=FALSE}
library(dplyr)
summarize(plasmaAlu,mean(alu))
summarize(plasmaAlu,sd(alu))
```

## Software Notes 

* Recall from the chapter 2-3 demo that
`dplyr` is an add-on package for R that includes useful 
tools for manipulating datasets in R.
    + To use `dplyr` functions we must first load the 
    package with `library(dplry)`.
    + The `summarize()` function from `dplyr`
    takes the dataset as its first
    argument, and the summaries to compute as additional
    arguments.

## CIs with the `t.test()` Function

* R's `t.test()` uses data to test hypotheses 
about a mean, or about differences between two means,
but the function also returns summary statistics and 
a CI.
    + Set the CI with the argument
    `conf.level` (default 0.95):

\scriptsize

```{r}
with(plasmaAlu,t.test(alu,conf.level=0.95))
```


## Do-It-Yourself CIs in R

* In R, the quantile function `qt()` for the t distribution can be used to find 
the critical value for the CI when $\sigma$  is unknown.

    + The probability argument, `p`, is the upper-tail area to the right of the desired quantile, under the t distribution.  

\scriptsize

```{r}
qt(p=(1-0.95)/2,df=9,lower.tail=FALSE)
```

\normalsize

* Use R to calculate the 95% CI:

    + Put the quantile into `tstar` with the assignment operator, `<-`
    + Extract the variable `alu` from the data frame `plasmaAlu` with the 
    `$` operator, and find its sample mean and SD.

\scriptsize

```{r}
tstar <- qt(p=(1-0.95)/2,df=9,lower.tail=FALSE)
alu <- plasmaAlu$alu
xbar <- mean(alu); s <- sd(alu); n <- 10
CI<- c( xbar - tstar*s/sqrt(n), xbar + tstar*s/sqrt(n))
CI
```


# Hypothesis Tests (Chapter 10)


## Example

* Go back to the example on plasma aluminum levels in infants.
* Suppose that in the population of infants **not** taking antacids,
the mean plasma aluminum levels are known to be $4.13 \mu g/l$.

* Recall that in the sample of 10 infants taking antacids, the plasma aluminum levels had sample mean and sample SD of $\bar{x} = 37.20 \mu g/l$ and $s=7.13$, respectively.
    
* Let the null hypothesis be $H_0: \mu=4.13$, that the mean plasma aluminum for infants taking 
antacids is the same as for the population of
infants not taking antacids.
* Let the alternative hypothesis be $H_a:\mu \not= 4.13$, that  
the mean plasma levels of aluminum for infants taking
antacid is different from the population of infants not taking antacids.


## Do-It-Yourself Hypothesis Test

* The $t$-statistic is
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}
$$
\footnotesize
```{r}
(37.20 - 4.13)/(7.13/sqrt(10))
```
\normalsize
* The $p$-value is 
$2P(T \geq |14.67|)$ for $T$ with $10-1=9$ df:
\footnotesize
```{r}
2*pt(14.67,df=9,lower.tail=FALSE)
```

## Using the Data

* Let's use the fake data `alu` from the confidence
interval example.
* The `t.test()` function will perform the test.
    + Specify the null hypothesis 
$H_0:\mu=4.13$ with the `mu` argument.
    + The function's default is to use the two-sided
    alternative hypothesis, which in this case is
    $\mu \not= 4.13$:


\scriptsize
```{r}
with(plasmaAlu,t.test(alu,mu=4.13))
```
