---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demos for Chapters 8-10: Review of Statistical Inference'
author: "Jinko Graham"
date: '2018-09-01'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Confidence Intervals (Chapter 9)

## Example

* Example from the text, page 223, summarizes data on plasma aluminum levels,
in $\mu g/l$, for $n=10$ infants receiving antacids that contain aluminum.
    + The sample mean of the plasma aluminum levels is $\bar{x} = 37.20 \mu g/l$, and the sample SD is $s=7.13$
* I can't access the data from the text, but I have
simulated a data set of $n=10$ subject with
similar properties for this demo.
    + Read these data into R from my website:

\scriptsize

<!--
```{r}
set.seed(123)
plasmaAlu <- data.frame(alu = rnorm(n=10,mean=37.2,sd=7.13))
write.csv(file="pa.csv",plasmaAlu,row.names=FALSE)
```
\normalsize
-->
```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/pa.csv")
plasmaAlu <- read.csv(uu)
head(plasmaAlu)
```

## Simulated Data Summary Statistics

* Our simulated data are similar to the data from 
the text, but the sample mean and SD are different:

\small

```{r,message=FALSE,warning=FALSE}
library(dplyr)
summarize(plasmaAlu,mean(alu))
summarize(plasmaAlu,sd(alu))
```

## Software Notes 

* Recall from the chapter 2-3 demo that
`dplyr` is an add-on package for R that includes useful 
tools for manipulating datasets in R.
    + To use `dplyr` functions we must first load the 
    package with `library(dplyr)`.
    + The `summarize()` function from `dplyr`
    takes the dataset as its first
    argument, and the summaries to compute as additional
    arguments.

## Do-It-Yourself CIs in R

* In R, the quantile function `qt()` for the t distribution can be used to find 
the quantile or critical value for the CI when $\sigma$ is unknown.

    + When the argument `lower.tail=FALSE', the first argument, `p`, is the upper-tail area to the right of the desired quantile, under the t distribution.  

\scriptsize

```{r}
qt(p=(1-0.95)/2,df=9,lower.tail=FALSE)
```

\normalsize

* Once we have the quantile or critical value, we can calculate the 95% CI.

## 

* First, use the assignment operator `<-` to put the quantile into an R object `tstar`.

* Next, use the `summarize()` function in the `dplyr` package to get the sample mean, sample sd and the number of observations, $n$, 
and assign these to an R object `mysummary`. 

* Use the saved objects to calculate the lower and upper bounds of the confidence interval and put them into an R object `CI`.
   
\scriptsize

```{r}
tstar <- qt(p=(1-0.95)/2,df=9,lower.tail=FALSE)
mysummary <- summarize(plasmaAlu,xbar=mean(alu),s=sd(alu),n=n())
mysummary
lower<- with(mysummary, xbar - tstar*s/sqrt(n))
upper<- with(mysummary, xbar + tstar*s/sqrt(n))
CI<- c(lower, upper)
CI
```


## CIs with the `t.test()` Function

* R's `t.test()` uses data to test hypotheses 
about a mean, or about differences between two means,
but the function also returns summary statistics and 
a CI.
* Set the level or coverage probability of the CI with the argument
    `conf.level` (default $C=0.95$):

\scriptsize

```{r}
with(plasmaAlu,t.test(alu,conf.level=0.95))
```



# Hypothesis Tests (Chapter 10)


## Example

* Suppose that in the population of infants **not** taking antacids,
the mean plasma aluminum levels are known to be $4.13 \mu g/l$.
* Let the null hypothesis be $H_0: \mu=4.13$, that the mean plasma aluminum for infants taking 
antacids is the same as for the population of
infants not taking antacids.
* Let the alternative hypothesis be $H_a:\mu \not= 4.13$, that  
the mean plasma levels of aluminum for infants taking
antacid is different from the population of infants not taking antacids.


## Do-It-Yourself Hypothesis Test

* The $t$-statistic is
$$
t = \frac{\bar{x} - \mu_0}{s/\sqrt{n}}
$$
* We can implement this as follows using
the summaries in the object `mysummary` from
the CI demo:

\scriptsize

```{r}
mysummary
with(mysummary,(xbar - 4.13)/(s/sqrt(n)))
```
\normalsize
* The $p$-value is 
$2P(T \geq |15.63|)$ for $T$ with $10-1=9$ df:
\scriptsize
```{r}
2*pt(15.63,df=9,lower.tail=FALSE)
```

## Using the `t.test()` Function

* The `t.test()` function will also perform the test.
    + Specify the null hypothesis 
$H_0:\mu=4.13$ with the `mu` argument.
    + The function's default is to use the two-sided
    alternative hypothesis, which in this case is
    $\mu \not= 4.13$:


\scriptsize
```{r}
with(plasmaAlu,t.test(alu,mu=4.13))
```
