---
title: "Sampling R Demo"
author: "Brad McNeney"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
```

## The `survey` package

* Computing standard errors, 
confidence intervals and test statistics from
complex survey data
is tricky and requires information on clusters and strata.
    + This information is not included in the CCHS-HA data files.
* However, we can use the survey weights to take the
unequal sampling probabilities into account
and get some "quick and dirty" CIs and tests. 
    + A useful tool for doing so is the `survey` package.
* Use the `Tools -> Install Packages` menu in 
RStudio to install.

\scriptsize

```{r}
uu<-url("http://people.stat.sfu.ca/~mcneney/Teaching/Stat305/Data/HUI.csv")
hui <- read.csv(uu)
library(survey)
```


## Specifying the Design in `survey`

* Details of the survey design and the data are
encapsulated in a `survey.design` object by 
the `svydesign()` function.
* We pass `svydesign()` the variables in the data frame that 
identify clusters and strata, if any, and the survey 
weights.
    + Argument `id` specifies cluster IDs, `strata` the
    stratum variables, and `weights` the sampling weights.
    + For our CCHS-HA data we don't know the clusters or
    strata, and pass only the weights.

\scriptsize

```{r}
dd <- svydesign(id=~1,strata=NULL,weights=~WTS_M,data=hui)
```

## Software Notes

* Cluster ID, stratum and weight variables are specified
as one-sided "formulas", of the form `~variable`.
    + We saw two-sided formulas for relationships between
    a response and grouping variable in `t.test()`.
    

## Means and SEs

* Use `svymean()` to calculate means and SEs that 
account for the design. 
    + Remove missing values from the calculation with
    `na.rm=TRUE`
* Use `svyby()` to stratify means on a grouping variable.

\scriptsize

```{r}
svymean(~HUIDHSI,design=dd,na.rm=TRUE)
svyby(~HUIDHSI,by=~DHH_SEX,design=dd,FUN=svymean,na.rm=TRUE)
```

## Confidence Intervals

\scriptsize

```{r}
confint(svymean(~HUIDHSI,design=dd,na.rm=TRUE))

confint(svyby(~HUIDHSI,by=~DHH_SEX,design=dd,FUN=svymean,na.rm=TRUE))
```

## t-tests

* `svyttest()` does t-tests.

\scriptsize

```{r}
svyttest(HUIDHSI~DHH_SEX,design=dd,na.rm=TRUE)
```