---
title: "Sampling"
author: "Brad McNeney"
date: '2019-01-28'
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,message=FALSE)
```

## Introduction to Sampling: Definitions

Reference: Sampling: Design and Analysis, by S. Lohr (1999).

+ **Observational unit** An object on which a measurement
is taken.
+ **Target population** The complete collection of 
observations that we want to study.
+ **Sample** A subset of the population
+ **Sampled population** The population from which the 
sample was taken.
+ **Sampling unit** The unit we actually sample.
    * E.G., we may want to study individuals, but do
    not have a list of individuals in the target population.
    Instead we sample households. The observational units
    are the individuals living in the households.
+ **Sampling frame** The list of sampling units.

## Selection Bias

* Selection bias occurs when some part of the target population
is not in the sampled population.
    + For example, if a survey of household income omits
    transient people, the estimates of average household
    income are likely too large.
* Convenience samples are often biased, since the units
that are easiest to select or are most likely to 
respond are usually not representative.
    + Mall surveys, mail-in surveys, web surveys.
    + Nonresponders often differ from responders. 

## Probability Sampling

+ In a probability sample, each unit in the population
has a known probability of selection.
+ In a **simple random sample** (SRS), every unit in 
the population has the same probability of being
included in the sample.
+ In a **stratified random sample** the population 
is divided into subgroups, or strata, and an SRS is
taken from each stratum.
+ In a **cluster sample**, observation units are aggregated
into larger sampling units called clusters. We draw a
sample of clusters and then subsample all or some 
observational units within the sampled clusters.

## Stratified Random Sampling

* One example is case-control sampling, where we
draw separate samples from the population of cases
and controls.
    + The motivation is our need for a representative
    sample of a rare segment (cases) of the population.
* Other reasons to draw a random sample include
protecting against a bad SRS, lower cost of administering
the survey, and ensuring comparable precision 
of estimates (e.g., of means) within strata.

## Cluster Sampling

* We aggregate the observational units into clusters,
sample clusters, and then sample all or some observational
units within clusters.
* The clusters are called primary sampling units (psu's)
and the samples within clusters are called secondary
sampling units (ssu's).
* The primary motivation for cluster sampling is to 
save money when sampling units from a population
that is geographically dispersed (like Canada)
or naturally grouped into clusters (like schools,
or hospitals). 

## Complex Surveys

* Complex surveys may employ a combination of 
clustering and stratification. 
* Example: The Canadian Community Health Survey -- Healthy
Aging.
* See the documentation available on canvas.

## Demographic and HUI Variables

* Illustrate survey information 
with a selection of demographic and
health-utility-index (HUI) variables.

\scriptsize

```{r}
uu<-url("http://people.stat.sfu.ca/~mcneney/Teaching/Stat305/Data/HUI.csv")
hui <- read.csv(uu)
head(hui,n=3)
```

## 

\tiny

```{r}
summary(hui)
```

## Sampling Weights

* The sampling weight for a sample member is
the number of units in the population represented
by the sample member.
    + The sampling weight is the inverse of the 
    probability that the sample member
    was included in the sample.
* For example, if the population has 1600 men and
400 women, and a stratified sample is of 
200 men and 200 women, then each sampled man represents
8 and each sampled woman represents 2.
    + Each man has a 1/8 chance
    and each woman a 1/2 chance of being included in the study.
* For complex surveys, determining sampling weights 
requires specialized expertise.
    + In the CCHS-HA, sampling weights are provided
    in the `WTS_M` variable.

## Sampling Weights for CCHS-HA

\scriptsize

```{r}
library(dplyr)
hui %>% group_by(GEO_PRV,GEOGCMA2) %>%
  summarize(Q1=quantile(WTS_M,.25),
            Q2=quantile(WTS_M,.5),
            Q3=quantile(WTS_M,.75))
```

## Sampling Weights by Age

\scriptsize

```{r}
hui %>% group_by(DHHGAGE) %>%
  summarize(Q1=quantile(WTS_M,.25),
            Q2=quantile(WTS_M,.5),
            Q3=quantile(WTS_M,.75))
```

## Weighted Means

* The estimate of the population mean from a weighted
sample $y_1,\ldots,y_n$ with weights $w_1,\ldots,w_n$
is 
$$ \frac{\sum_{i=1}^n w_i y_i}{\sum_{i=1}^n w_i}$$

\scriptsize

```{r}
hui %>% group_by(DHH_SEX) %>% 
  summarize(unwtd.mean=mean(HUIDHSI,na.rm=TRUE),
            wtd.mean=weighted.mean(HUIDHSI,w=WTS_M,na.rm=TRUE))
```

