---
title: "Stat 305 Exercise Set 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Chapter 15, #10 but with modification. Survey of Italian physicians on breast cancer. Physicians in different specialties were asked about
their recommendations for radical (mastectomy) vs.
conservative (lumpectomy) surgery. In particular, 
would they recommend radical surgery regardless
of a patient's age (RR), conservative surgery for 
younger patients only (CY), or conservative surgery regardless of age (CR)?


To answer this question we need a data-frame 
that contains the results of the survey.  
Create one called `survey` with 3 
variables `RR`, `CY` and `CR` as follows:
```{r}
survey <- data.frame(
  RR=c(6,23,2,1,1), 
  CY=c(22,61,3,12,12), 
  CR=c(42,127,54,43,31))
row.names(survey)<-c("Internal", "Surgery", "Radiotherapy",
              "Oncology", "Gynecology")
survey
```

(a) At the 0.05 level, test the null hypothesis
that there is no association between physician
specialty and recommended treatment. (Note: Ignore
the warning message from R about the chi-squared approximation.)

**Answer:**

```{r}
# Add your R commands here.  
```

*Insert your written comments here.*

(b) Report the expected cell
counts in the table under the null hypothesis. Comment on whether the chi-square
approximation is reliable for these data.

**Answer:**

```{r}
# Add your R commands here.  
```

*Insert your written comments here.*


(c) What do you conclude?

**Answer:**

*Insert your written comments here.*

## Question 2

Chapter 15, #12 but with modification. Intra-observer
variability in assessing cervical smears. This study
screened 3325 pap-smear slides for the presence or
absence of abnormal squamous cells. Each slide
was screened by a particular observer and then
rescreened 6 months later by the same observer.
The results of the study are shown below:

<table cellpadding="10">
<tr> 
 <td> &nbsp; </td>  
 <td colspan="2" align="center"> &nbsp; <b>2nd screen</b> &nbsp; </td>
 <td> &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; <b>1st screen</b> &nbsp; </td> 
 <td> &nbsp; Present &nbsp; </td> 
 <td> &nbsp; Absent &nbsp; </td> 
 <td> &nbsp; Total &nbsp; </td>
</tr>
<tr>
 <td> &nbsp; Present &nbsp; </td> 
 <td> &nbsp; 1763 &nbsp; </td> 
 <td> &nbsp; 489 &nbsp; </td> 
 <td> &nbsp; 2252 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Absent &nbsp; </td> 
 <td> &nbsp; 403 &nbsp; </td> 
 <td> &nbsp; 670 &nbsp; </td> 
 <td> &nbsp; 1073 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Total &nbsp; </td> 
 <td> &nbsp; 2166 &nbsp; </td> 
 <td> &nbsp; 1159 &nbsp; </td> 
 <td> &nbsp; 3325 &nbsp; </td> 
</tr>
</table>
<p>


(a) Do these data support the null hypothesis
that there is no association between the time of 
screening and diagnosis? 

 **Answer:**

```{r}
# Add your R commands here.  
```

*Insert your written comments here.*

The data could also be displayed in the following
manner:
<table cellpadding="10">
<tr> 
 <td> &nbsp; </td>  
 <td colspan="2" align="center"> &nbsp; <b>Screening</b> &nbsp; </td>
 <td> &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; <b>Abnormal cells </b> &nbsp; </td> 
 <td> &nbsp; First &nbsp; </td> 
 <td> &nbsp; Second &nbsp; </td> 
 <td> &nbsp; Total &nbsp; </td>
</tr>
<tr>
 <td> &nbsp; Present &nbsp; </td> 
 <td> &nbsp; 2252 &nbsp; </td> 
 <td> &nbsp; 2166 &nbsp; </td> 
 <td> &nbsp; 4418 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Absent &nbsp; </td> 
 <td> &nbsp; 1073 &nbsp; </td> 
 <td> &nbsp; 1159 &nbsp; </td> 
 <td> &nbsp; 2232 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Total &nbsp; </td> 
 <td> &nbsp; 3325 &nbsp; </td> 
 <td> &nbsp; 3325 &nbsp; </td> 
 <td> &nbsp; 6650 &nbsp; </td> 
</tr>
</table>
<p>


(b) Is there anything wrong with this presentation?
Please describe briefly.

**Answer:**

*Insert your written comments here.*

(c) If you didn't know there was anything wrong with the
presentation in part b, how would you analyze the data? For instance, what type of test would you apply and what would be the null hypothesis that you'd be testing? (Note: You don't actually have to do the test; just say what you would do.)

**Answer:**

*Insert your written comments here.*


## Question 3

Chapter 15, #18. Alcohol and miscarriages. A study investigating the association between spontaneous abortion (miscarriage) and alcohol consumption obtained the data below. Note that the second column labeled "pregnancies" counts the total number of women who got pregnant whereas the third column labeled "miscarriages" counts the number (out of the total in the second column) who subsequently lost their pregnancy due to spontaneous abortion. 

<table cellpadding="10">
<tr> 
 <td> <b>drinks/week</b> &nbsp; </td>
 <td> <b>pregnancies</b> &nbsp; </td>
 <td> <b>miscarriages</b> &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 0 &nbsp; </td>
 <td> &nbsp; 33164 &nbsp; </td>
 <td> &nbsp; 6793 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 1-2 &nbsp; </td>
 <td> &nbsp; 9099 &nbsp; </td>
 <td> &nbsp; 2068 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 3-6 &nbsp; </td>
 <td> &nbsp; 3069 &nbsp; </td>
 <td>  &nbsp; 776 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 7-20 &nbsp; </td>
 <td> &nbsp; 1527 &nbsp; </td>
 <td>  &nbsp; 456 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 21+ &nbsp; </td>
 <td> &nbsp; 287 &nbsp; </td>
 <td> &nbsp; 98 &nbsp; </td>
</tr>
</table>
<p>

Below is a revised code chunk using the style of
the revised chapter 15, part 4 R demo.
For parts a-d below, you may **modify** this R code chunk as necessary (and then run it) if you find it helpful.
```{r, message=FALSE, warning=FALSE}
group=c("1-2", "3-6", "7-20", "21+")
preg <- c(9099,3069,1527,287)
mis <- c(2068,776,456,98)
mydf <- data.frame(group=group,a=mis,b=preg-mis,c=6793,d=33164-6793)
mydf # comment: the values of c and d are repeated as necessary
lev <- 0.80 #confidence level or coverage probability
critval <- qnorm( (1-lev)/2, lower.tail=FALSE)
library(dplyr)
mydf <- mutate(mydf, OR=a*d/(b*c),
               logOR = log(OR),
               se.logOR = sqrt(1/a + 1/b + 1/c + 1/d),
               ci.lower = exp(logOR-critval*se.logOR),
               ci.upper = exp(logOR+critval*se.logOR))
print(mydf,digits=3)
```

(a) For each level of alcohol consumption, estimate
the probability that a woman who becomes pregnant
will have a miscarriage.

**Answer:**

*Insert your written comments here.*

(b) For each category of alcohol use, estimate the
relative odds of miscarriage 
for women who consume some amount of alcohol versus
those who do not consume any. 

**Answer:**

*Insert your written comments here.*

(c) In each case, calculate a 95\% confidence 
interval for the odds ratio

**Answer:**

*Insert your written comments here.*

(d) What do you conclude? Do the data suggest a dose-response 
relationship between alcohol use and the odds of miscarriage?

**Answer:**

*Insert your written comments here.*

## Question 4

Chapter 15, #21, with modification. Beer-drinking questionnaires. Two different questionnaires addressing drinking habits
were administered to the same set of men and women aged 50-65 years from a certain community. The first questionnaire format concerned
general diet and lifestyle while the second format specifically targeted alcohol use. For each questionnaire, the responses were processed to classify a respondent as either a non-drinker or drinker of beer.  Categories for the general questionnaire are saved under the variable `genques` and those for the questionnaire targeting alcohol use are saved under the variable `alcques`.
The data are available as a plain text file; see the R code below for accessing it.

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/alcohol.txt")
alc <- read.table(uu,header=TRUE)
head(alc) 
```

a) Are the drinking statuses on the two different types of questionnaires associated? Use a 5\% level in your test. What do you conclude?


**Answer:**

```{r}
# Put your R commands here.
```

*Insert your written comments here.*

b) Is the type of questionnaire associated with drinking status? Use a 5\% level and continuity correction in your test. What do you conclude?

**Answer:**

```{r}
# Put your R commands here.
```

*Insert your written comments here.*
