---
title: "Assignment 1, Question 1"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To answer this question we need a data frame 
that contains the numbers of hysterectomies for the
15 Swiss doctors in the class notes.
We can create a data frame called `hyst` with one 
variable `numHyster`, as follows:
```{r}
hyst <- data.frame(numHyster = c(20,25,25,27,28,31,33,34,
                                 36,37,44,50,59,85,86))
```

Notice that I'm cutting-and-pasting from the `ch2-3Demo.Rmd`
demo. You
will find lots of useful R code in the demos to draw from, with little or no modification.

(a) Draw a histogram of the data. Do you see any outliers?

**Answer:**

```{r}
library(ggplot2)
# Add your R commands to draw a histogram here -- the R demos in 
# the lecture notes should be helpful. Note that the Workshop is
# there to help you through assignments, including getting 
# comfortable with R and statistical thinking; please make use 
# of this excellent resource. Go early to avoid the last-minute 
# rush and crashing the servers. Watch out especially for Crowdmark # crashing if you are trying to upload your assignment at the 
# last minute when there is high volume. The Workshop is not
# responsible for late assignments due to server crashes.  
```

*Insert your written comments here about the outliers that you see
in the histogram.*

(b) Compute the sample mean, standard deviation, median, 25th
percentile and 75th percentile from these data. Comment on the difference
between the mean and the median, and the difference 
between the SD and the IQR.

**Answer:**

```{r}
library(dplyr)
# Add your R commands to do this question here. You can use
# the summarize() function as in the demo in ch2-3Demo.Rmd.
```

*Add your written comments that interpret the output of the R 
commands here*

