---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 15, part 3: McNemar''s Test'
author: "Jinko Graham"
date: '2018-10-22'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
```


## Paired Data and McNemar's Test

<!--
* The chi-square test is not appropriate when the 
observations being counted in the table are paired in some way.
-->
* The chi-square test is not appropriate when the study is
designed to collect paired data.
* Example: Study of smoking habits over time.
    + A random sample of 2110 people were questioned about smoking status
    in 1980 and again in 1982.
    + Are smoking status and year associated?
i.e, does the population proportion of smokers 
differ by year?
    + Test $H_0: p_{1980} - p_{1982} = 0$ vs.
    $H_a: p_{1980} - p_{1982} \not = 0$,
    where $p_{1980}$ and $p_{1982}$ are the population
    proportions of smokers in 1980 and 1982, respectively. 

* The data might look as follows:

\footnotesize

\begin{center}
\begin{tabular}{cll}
       &\multicolumn{2}{c}{\underline{Smoking}}\\ 
person & 1980 & 1982 \\
1 & no & yes \\
2 & no & no \\
3 & yes & no \\
$\vdots$ & $\vdots$ & $\vdots$ \\
2110 & yes & yes \\
\end{tabular}
\end{center}

\normalsize

##         

* The data could be arranged in a table as:

\footnotesize

\begin{center}
\begin{tabular}{ll|c|c|r}
& \multicolumn{1}{c}{}& \multicolumn{2}{c}{Year} & \\
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{1980} & \multicolumn{1}{c}{1982} & \\ \cline{3-4}
Smoke & Yes & 717 & 696 & 1413 \\ \cline{3-4}
& No &  1393 & 1414 & 2807 \\ \cline{3-4}
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{2110} & \multicolumn{1}{c}{2110} & 4220 \\
\end{tabular}
\end{center}

\normalsize 

* But this table is misleading: the 4220 observations that it 
reports are from 2110 double-counted people.

* The 4220 observations counted in the table are not independent, 
but rather paired observations from just 2110 people.

## 

* To clarify the paired nature of the data, they are typically  arranged as:

\footnotesize

\begin{center}
\begin{tabular}{ll|c|c|r}
& \multicolumn{1}{c}{}& \multicolumn{2}{c}{1982 (after)} & \\
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{Smoker} & \multicolumn{1}{c}{Nonsmoker} & \\ \cline{3-4}
1980 (before) & Smoker & 620 & 97 & 717 \\ \cline{3-4}
& Nonsmoker &  76 & 1317 & 1393 \\ \cline{3-4}
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{696} & \multicolumn{1}{c}{1414} & 2110 \\
\end{tabular}
\end{center}

\normalsize

* The observations that are counted in the cells of this table 
are people and are independent.
* The cells with the same before- and after-status
of the subject are called _concordant_.
* The cells with different before- and after-status 
are called _discordant_.
* McNemar's test is a contrast between the discordant cells, and ignores
the concordant cells.



## 
\footnotesize

\begin{center}
\begin{tabular}{ll|c|c|r}
& \multicolumn{1}{c}{}& \multicolumn{2}{c}{1982 (after)} & \\
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{Smoker} & \multicolumn{1}{c}{Nonsmoker} & \\ \cline{3-4}
1980 (before) & Smoker & 620 &  $r=97$ & 717 \\ \cline{3-4}
& Nonsmoker &  $s=76$ & 1317 & 1393 \\ \cline{3-4}
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{696} & \multicolumn{1}{c}{1414} & $n=2110$ \\
\end{tabular}
\end{center}

\normalsize

* Our sample of 2110 individuals has 620 continuing smokers, $r=97$ quitters and $s=76$ starters in 1982.
* In 1980 and 1982, the sample proportions of smokers are, respectively,
    + $\hat{p}_{1980} = (620+r)/n=(620+97)/2110$
    + $\hat{p}_{1982}=(620+s)/n = (620+76)/2110$

* So, $\hat{p}_{1980} \not= \hat{p}_{1982}$
when $r \not= s$; 
or when $\hat{p}_{start} \not= \hat{p}_{quit}$, where
    + $\hat{p}_{start} = s/n$ is the proportion starting in 1982
    + $\hat{p}_{quit}=r/n$ is the proportion quitting in 1982. 
    
* Re-express 
 $H_0: p_{1980} - p_{1982} = 0$ vs.
    $H_a: p_{1980} - p_{1982} \not = 0$ as
    + $H_0': p_{quit} = p_{start}$ vs.
$H_a':p_{quit} \not= p_{start}$.


## McNemar's Test

* Base the hypothesis test on the difference in observed proportions
$\hat{p}_{start} - \hat{p}_{quit}=(r-s)/n$
and its standard error.
* We'll skip the derivation but the test statistic ends up being:
$$
X^2 = \frac{(r-s)^2}{r+s} \sim \chi^2_1 .
$$
* If the number of quitters, $r$,
is very different from the number of starters, $s$, 
the statistic $X^2$ is **big** and we reject $H_0$ in favour of $H_a$. 
* An alternate form  that uses a
continuity correction for small samples (text, page 351)  is
$$
X^2 = \frac{(|r-s|-1)^2}{r+s} \sim \chi^2_1 .
$$


## McNemar's Test for the Smoking Data

* We have $r=97$ and $s=76$. The test statistic with 
continuity correction is 
$$ \frac{(|97-76|-1)^2}{97+76} = 2.31$$
and the corresponding p-value 
is 0.128 (see R demo).

<!--
\scriptsize

```{r}
X2 <- (abs(97-76)-1)^2/(97+76)
X2
pval<-pchisq(X2,df=1,lower.tail=FALSE)
pval
```

\normalsize
-->

*  Taking $\alpha=.05$, there is insufficient statistical evidence to conclude that smoking status is associated with year (the pvalue 0.128 is $>0.05$).

## Other Examples of Paired Data

* Scoring individuals from the same matched pair.
    + e.g. case-control pairs in which the control has been matched to the case on a number of characteristics. 
* Scoring the same experimental unit with two different techniques
* Ratings of the same experimental unit by two different raters
* Genetic variants that were transmitted/not transmitted 
from a parent (experimental unit) to its offspring

## Example: Transmission/Disequilibrium Test (TDT)

* [Spielman _et al._, 1993](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1682161/) \tiny click \normalsize
developed a test for genetic linkage that is an application
of McNemar's test
* DNA segments that are
physically close together on a chromosome, or "genetically linked", tend
to be co-transmitted from parent to offspring.
    + A DNA marker that is genetically linked to a 
    disease-causing mutation (i.e. they are physically close on the chromosome) should tend to be co-transmitted with the disease.
* Application to type 1 autoimmune diabetes (T1D):
    + A genetic marker called 5'FP located 
    near the insulin gene was associated with
    T1D in the population. Is 5'FP linked to a
    disease-causing mutation?
    + If so, we should see certain variants of 5'FP
    over-represented in transmissions from parents to 
    children affected by T1D.
    
## Diabetes Data from [Spielman _et al._](http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/tdt.csv)

* The DNA marker had two variants denoted 1 and X.
* Study of 124 parents of children with T1D
    + All parents selected to carry a 1 and an X at the DNA marker. 

* The dataset has
    + one row for each parent, and
    + two columns, one for the variant that was transmitted 
    from the parent to the affected child, and one for the 
    variant that was not transmitted.

<!--
```{r}
tdtex <- data.frame(transmitted = factor(c(rep("1",78),rep("X",46))),
                   untransmitted = factor(c(rep("X",78),rep("1",46))))
write.csv(tdtex,file="tdt.csv",row.names=FALSE)
# Now from a terminal, upload tdt.csv to JG website:
#  scp tdt.csv jgraham@rcg-linux-ts1.rcg.sfu.ca:WWW_HOME/Teaching/S305_17/Data/.
```
-->

\scriptsize

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/tdt.csv")
tdtex <- read.csv(uu)
head(tdtex,n=8)
```

## McNemar's test on T1D Data

* Cross tabulating `transmitted` and `untransmitted` 
variables for the 
heterozygous parents yields the following table. 

\scriptsize

```{r}
tt <- table(tdtex)
tt
```

\normalsize

* McNemar's test applied to the T1D data (see R demo):

\scriptsize

```{r}
mcnemar.test(tt,correct=FALSE)
```

\normalsize

* There is strong evidence that the 1 variant is
preferentially transmitted over the X variant 
to the affected child.
    + The marker near the insulin gene is
    genetically linked to T1D.
    
<!--
* Recent work [e.g., Pugliese (2005)](https://www.ncbi.nlm.nih.gov/pubmed/16081366)
suggests that variation in gene expression in the thymus is
a risk factor for T1D, so perhaps the causal locus is in 
the promoter region of the insulin gene.
-->
