---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Demo for Chapter 15, part 4: Inference for Odds Ratios'
author: 'Jinko Graham, Brad McNeney'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## Testing whether $OR=1$

* The chi-square test assesses the null hypothesis
that $OR=1$ (no association between exposure and disease)
against the alternative hypothesis that $OR\not=1$ (an association). 

\scriptsize

```{r}
mydf <- data.frame(case=c(1350,7),control=c(1296,61)) # Doll and Hill's data
rownames(mydf) <- c("smoker","non-smoker")
mydf
chisq.test(mydf)
```

## Using R to get point and interval estimates of the OR. 


* For a $2\times 2$ table, the calculations are easy enough
to do with a hand calculator or spreadsheet.
* Below we present some R code that illustrates
data frame manipulation in R.
    + This will give us code 
    that will generalize to the case of multiple exposures,
    as in the Doll and Hill dataset where smoking status
    has four levels.
    
## Data Frame for Confidence Intervals

\scriptsize
 
```{r}
library(dplyr) # for the mutate() function
mydf <- data.frame(a=1350,b=1296,c=7,d=61)
zstar <- qnorm((1-.95)/2,lower.tail=FALSE)
mydf <- mutate(mydf,
               OR=a*d/(b*c),
               logOR=log(OR),
               SE=sqrt(1/a+1/b+1/c+1/d),
               logci.lower=logOR-zstar*SE,
               logci.upper=logOR+zstar*SE,
               ci.lower=exp(logci.lower),
               ci.upper=exp(logci.upper))
round(mydf,2) # Round all numbers to 2 decimals when printing
```

\normalsize

* `mutate()` is used to create new variables from existing ones
and add them to our data frame.
* In this example, variables such as `OR` and `logOR`
are created and added to `mydf`.
    + Notice that the calculation of `logOR` can use
    the newly-created variable `OR`.


## More Than Two Exposure Levels

* Doll and Hill's data with smokers classified by the average number
of cigarettes per day:
\begin{center}
\begin{tabular}{llcc} 
 & &  case & control \\ \hline
Number of   & 25+    & 340 & 182  \\
cigarettes  & 15-24  & 445 & 408  \\ 
per day     & 1-14    & 565 & 706  \\ 
            & 0      & 7 & 61  \\ \hline
\end{tabular}
\end{center}
* Can use the last row with 0 cigs per day (unexposed) as a baseline group, and calculate our ORs for 
each level of exposure. 

* Here is where the R code we wrote can pay off. We 
essentially repeat the code, but with different
definitions of `a`, `b`, `c` and `d`.

##

\scriptsize

```{r}
mydf <- data.frame(a=c(340,445,566),
                   b=c(182,408,706),
                   c=c(7,7,7),
                   d=c(61,61,61))
mydf <- mutate(mydf,
               OR=a*d/(b*c),
               logOR=log(OR),
               SE=sqrt(1/a+1/b+1/c+1/d),
               logci.lower=logOR-zstar*SE,
               logci.upper=logOR+zstar*SE,
               ci.lower=exp(logci.lower),
               ci.upper=exp(logci.upper))
round(mydf,2) # 
```

