---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 15, part 3: McNemar''s Test'
author: "Jinko Graham"
date: '2018-10-06'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## McNemar's Test for the Smoking Data

* We have $r=97$ and $s=76$. The test statistic with 
continuity correction is 
$$ \frac{(|97-76|-1)^2}{97+76} = 2.31$$
and the corresponding p-value 
is 0.128:

\scriptsize

```{r}
X2 <- (abs(97-76)-1)^2/(97+76)
X2
pval<-pchisq(X2,df=1,lower.tail=FALSE)
pval
```

\normalsize


## McNemar's test on T1D Data

* Read in the data and cross tabulate:

\scriptsize

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/tdt.csv")
tdtex <- read.csv(uu)
head(tdtex,n=8)
tt <- table(tdtex)
tt
```

\normalsize

##

*  Now apply McNemar's test *without* the continuity correction to the cross-tabulated data:

\scriptsize

```{r}
mcnemar.test(tt,correct=FALSE)
```

\normalsize

* Apply McNemar's test *with* the (default) continuity correction:

\scriptsize

```{r}
mcnemar.test(tt)
```


