---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 15, part 4: Inference for Odds Ratios'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
```

## Estimating Odds Ratios

* For Doll and Hill's lung-cancer data, 
we estimated the odds ratio from the sample proportions
of smokers in the cancer (case) and non-cancer (control) groups.

\begin{center}
\begin{tabular}{llll} 
 ~ & ~ &  case & control  \\ \cline{3-4}
Smoke & Yes & $a=1350$ & $b=1296$  \\
 (E)  & No  & $c=7$ & $d=61$  \\ \cline{3-4}
~ & ~ & $a+c=1357$ & $b+d = 1357$
\end{tabular}
\end{center}

* OR estimate is
$$ 
\widehat{OR} = \frac{\frac{a}{a+c}}{1-\frac{a}{a+c}}
\Bigg /
\frac{\frac{b}{b+d}}{1-\frac{b}{b+d}} = \frac{ad}{bc}
$$

* For Doll and Hill's data, we have
$$
\widehat{OR} = \frac{ad}{bc}=\frac{1350 \times 61}{1296 \times 7}=9.1.
$$

## Testing whether $OR=1$

* The chi-square test assesses the null hypothesis
that $OR=1$ (no association between exposure and disease)
against the alternative hypothesis that $OR\not=1$ (an association). 

\scriptsize

```{r}
mydf <- data.frame(case=c(1350,7),control=c(1296,61)) # Doll and Hill's data
rownames(mydf)=c("smoker", "non-smoker")
mydf
chisq.test(mydf)
```

## 
\normalsize

* The $p$-value for testing $H_0: OR=1$ vs.
$H_a: OR \not= 1$ is $7.552 \times 10^{-11}$.

* Strong evidence of association between lung cancer and smoking!

* Recall from the Chapter 6 notes that $OR \approx RR$, provided that the disease is rare.

* Assuming lung cancer is rare, we may approximate the relative risk of lung cancer by the odds ratio.  
    + We estimated the OR to be 
    $\widehat{OR}= \frac{a*d}{b*c}=
    \frac{1350*61}{1296*7}=9.1$
    + So we estimate that the risk of lung cancer in smokers is about 9.1 times the risk of lung cancer in non-smokers (i.e. $\widehat{RR}=9.1$). 


## Confidence Intervals for ORs
* Recall: The natural logarithm $\log_e(x)$ is defined 
so that $x=e^{\log_e(x)}$, where the base $e \approx 2.718$.
    + To get $x$, we **exponentiate** the natural logarithm $\log_e(x)$; i.e., we raise the base $e$ to the power of the exponent $\log_e(x)$.
* For large samples, it turns out that
$\log_e(\widehat{OR})$ is approximately normally distributed.
* This approximation leads to CIs for 
$\log_e( OR)$ of the form 
$$\mbox{estimate} \; \pm m.e., \; \; \mbox{where}$$
the margin of error term, m.e., is an SE times a critical value.
    + The standard error of $\log_e(\widehat{OR})$ is:
$$
SE = \sqrt{\frac{1}{a} + \frac{1}{b} + \frac{1}{c} + \frac{1}{d}}
$$
    + For a level-$C$ CI for 
 $\log_e(OR)$, the critical value is $z^*$, the upper $(1-C)/2$ critical value of the standard normal distribution.
 
## 

* A level-$C$ CI for $\log_e(OR)$  is thus 
$\log_e(\widehat{OR}) \pm z^* SE$, or
$$\left ( \log_e(\widehat{OR}) - z^* SE, \;\; \log_e (\widehat{OR)} + z^* SE \right ),$$
* To get the level-$C$ CI for $OR$, we 
exponeniate the lower and upper bounds of the CI above:
$$\left ( e^{\log_e (\widehat{OR}) - z^* SE}, \quad
e^{\log_e(\widehat{OR}) + z^* SE} \right )$$

<!--

## Using R to get point and interval estimates of the OR. 

* We write our own function in R to automate calculations:

\scriptsize
```{r}
orfunc<-function(a,b,c,d,conf.level=0.95) {
  or<-a*d/(b*c)
  se<-sqrt(1/a+1/b+1/c+1/d)
  zstar <- qnorm((1-conf.level)/2,lower.tail=FALSE)
  ci<-exp(c(log(or)-zstar*se,log(or)+zstar*se))
  output<-list(or=or,ci=ci, conf.level=conf.level)
  return(output)
  }
```

\normalsize

* We will apply ```orfunc()``` to the data in

\scriptsize
```{r}
mydf
```

\normalsize

* So $a=1350, b=1296 , c=7, d=61$.

## 

* Let's go ahead and use our function:

\scriptsize


```{r}
orfunc(a=1350,b=1296,c=7,d=61)
```

-->

## Application to Doll and Hill's Lung-Cancer Data

* The estimated OR is
$\widehat{OR}= \frac{1350*61}{1296*7}=9.1$, and its logarithm is
$\log_e(9.1) = 2.21$.
* The SE of $\log_e(\widehat{OR})$ is
$$
SE = \sqrt{ \frac{1}{1350} + \frac{1}{1296} + \frac{1}{7} + \frac{1}{61}} = 0.401
$$

* For a 95\% CI, the critical value is 
$z^* = 1.96$.

* The 95\% CI for $\log_e(OR)$ is therefore
$$
\left( 2.21 - 1.96\times 0.401, 2.21 + 1.96\times 0.041 \right)
= (1.42,3.00)
$$

* The 95\% CI for $OR$ is then
$$
\left( e^{1.42}, e^{3.00} \right) = (4.14,20.1)
$$


## Interpretation of Point Estimates of OR

* Smoking is associated with an estimated 9.1-fold increase in the odds of lung cancer.
* Or, if lung cancer is **rare**, we can interpret the OR as an RR and say:
    + smoking is associated with an estimated 9.1-fold increase in the ***risk*** of lung cancer. 

## Interpretation of Interval Estimates of OR

* "With 95% confidence, 
smoking is associated with an estimated 4.1 to 20-fold
increase in the odds of lung cancer"
    +  You can use the statement above, but keep in mind that it really means that: In 95 out of 100 datasets, we expect the CI, such as 4.1-20 for this dataset, to cover the true OR.
    
* Assuming that lung cancer is rare, we can interpret the OR as an RR and say: 
    + "With 95% confidence, 
smoking is associated with an estimated 4.1 to 20-fold
increase in the _risk_ of lung cancer"
    + i.e, in 95 out of 100 datasets, we expect the CI,  such as 4.1-20 for this dataset, to cover the true RR.
    
## More Than Two Exposure Levels

* Doll and Hill's data with smokers classified by the average number
of cigarettes per day:
\begin{center}
\begin{tabular}{llcc} 
 & &  case & control \\ \hline
Number of   & 25+    & 340 & 182  \\
cigarettes  & 15-24  & 445 & 408  \\ 
per day     & 1-14    & 565 & 706  \\ 
            & 0      & 7 & 61  \\ \hline
\end{tabular}
\end{center}
* Can use the last row with 0 cigs per day (unexposed) as a baseline group ($c=7, $d=61$), and calculate our ORs for 
each level of exposure. 
    + E.G. Estimated OR for 25+ vs. 0 cigs per day:
\begin{eqnarray*}
OR & = & \frac{\mbox{odds of lung cancer in exposed ($25+$ cigs/day)}} {\mbox{odds of lung cancer in unexposed ($O$ cigs/day)}} \\
&=& \frac{a * d}{b*c} 
= \frac{340 * 61}{182*7}
= 16.28
\end{eqnarray*}


## Odds-ratios with Multiple Exposure Levels

* Add estimated $OR$s to the table: 
\begin{center}
\begin{tabular}{llccr} 
 & &  case & control & $\widehat{OR}$ \\ \hline
Number of   & 25+    & 340 & 182 & 16.28  \\
cigarettes  & 15-24  & 445 & 408 & 9.50 \\ 
per day     & 1-14    & 565 & 706 & 6.97 \\ 
            & 0      & 7 & 61  & -- \\ \hline
\end{tabular}
\end{center}
* The increase in estimated $OR$s with exposure level suggests a "dose-response" 
relationship.
* For observational data such as these, a dose-response relationship is one of the criteria for establishing causality. 
    + In this study, the dose-response relationship with number of cigarettes per day was used to argue that smoking **causes** lung cancer.
    


## Including Confidence Intervals

<!--
* Let's apply the R function we wrote
to get point and interval estimates of the OR; e.g.,

\scriptsize

```{r}
orfunc(a=340,b=182,c=7,d=61)
```

\normalsize

* Include confidence intervals in the table
-->

\begin{center}
\begin{tabular}{llccrc} 
 & &  case & control &  $\widehat{OR}$ & 95\% CI \\ \hline
Number of   & 25+    & 340 & 182 & 16.28 & (7.30,36.32)  \\
cigarettes  & 15-24  & 445 & 408 & 9.50 & (4.30,21.02) \\ 
per day     & 1-14    & 565 & 706 & 6.97 & ( 3.17,15.37) \\ 
            & 0      & 7 & 61  & -- & -- \\ \hline
\end{tabular}
\end{center}

## Historical notes

* Doll and Hill's study of lung cancer was 
published in 1954
    + Turned the tide of public-health opinion on smoking.

* Famously, the iconic geneticist and statistician RA Fisher was strongly opposed (to the point of campaigning against it for the tobacco industry). 
<!--
    + Statistics comes out of genetics, through Fisher
    + He is credited with inventing experimental design in agricultural field experiments
    + In Genetics, he unified the concepts of natural selection and Mendel’s rules of inheritance.
-->
    
* Fisher was a heavy pipe smoker
    + Argued that correlation (association) is not causation 
    

* Died aged 72 in 1962, following complications from cancer surgery. 
  
    
## Chapter 15 Summary

* Contingency tables summarize the joint distribution of two categorical variables.
* The chi-square test tests for association between two categorical variables
* For data that are paired in some way, we use McNemar's test, which contrasts the 
discordant cells in a table that counts each pair just once.
* Testing for association in a $2\times 2$ table amounts to testing whether or not the
OR is 1. 
    + Can extend to $r \times 2$ tables with $r$ levels of the exposure.
* When the disease outcome is rare, the OR can
be interpreted as a relative risk (RR).
* Can obtain approximate CIss for the OR.
* Omitted Berkson's fallacy in text; beyond scope of course.