---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 15, part 2: Chi-Square Tests'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
```

## Chi-square tests of association

* In chapter 14, we tested for association between two categorical variables by testing for
differences between proportions
    + Call the test from Chapter 14 the $Z$ test.


* Applied the $Z$ test to data from the WHI.
    + Recall that the WHI randomized 16,608 post-menopausal women aged 50-79 years to receive either hormone replacement therapy (estrogen plus 
    progestin EP$+$;$n_1=8506$), or a placebo (EP$-$; $n_2=8102$).
    
* We tested for a difference in the proportions of women with invasive breast cancer (BC$+$) in the hormone replacement therapy (EP+) and placebo (EP$-$) groups.
    

## The [WHI Data](https://en.wikipedia.org/wiki/Women%27s_Health_Initiative) \tiny click \normalsize

<!--

```{r}
WHI <- data.frame(EP = factor(c(rep("EP+",8506),rep("EP-",8102))),
                  BC = factor(c(rep("BC+",166),
                                rep("BC-",8340),
                                rep("BC+",122),
                                rep("BC-",7980))))
head(WHI)
table(WHI)
write.csv(WHI,file="whi.csv",row.names=FALSE)
# Now from a terminal, upload whi.csv to JG website:
#  scp whi.csv jgraham@rcg-linux-ts1.rcg.sfu.ca:WWW_HOME/Teaching/S305_18/Data/.
```

-->

* The first few rows of the dataset are as follows: 

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/whi.csv")
WHI <- read.csv(uu)
head(WHI,n=8)
```

\normalsize

* A cross-tabulation of the BC and EP variables in the dataset is:

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/whi.csv")
WHI <- read.csv(uu)
wtab <- table(WHI)
wtab
```


## Association between HRT and breast cancer

* The table of proportions below gives the conditional distributions of BC status given EP status.
    + The proportions in each row add to 1.

* BC and EP are associated if their conditional distributions
    are different.

\scriptsize

```{r}
prop.table(wtab,margin=1)
```

\normalsize

* Previously, we used the Z test for different conditional distributions 
   + Looks for differences in the proportion of BC$+$ in the 
EP$-$ and EP$+$ groups.
   + Can be applied to data in $2 \times 2$ tables


## Chi-square test of association

* When applied to $2\times 2$ tables, 
the $Z$ test for a difference in proportions is
equivalent to the chi-square ($\chi^2$) test.
* But the chi-square test has the advantage of 
generalizing from $2\times 2$ tables to $r\times c$ 
tables, for $r\geq2$ rows and $c\geq 2$ columns. 
* Compares _observed_ cell counts to _expected_ counts
    + The expected count is the count we would expect if the null hypothesis of no 
    association were true (details deferred).
* The form of the statistic is
$$
X^2 = \sum_{\mbox{cells}} \frac{ \left( \mbox{observed} - \mbox{expected}
\right)^2}{\mbox{expected}}
$$

## Sampling distribution of $X^2$

* Under the null hypothesis of no association between 
row and column variables, the test statistic $X^2$ is approximately
distributed as a chi-square distribution with 
$(r-1)\times (c-1)$ degrees of freedom (df).
* Computer software gives upper tail probabilities 
of different chi-square distributions.

## Chi-square test for WHI example

* We can perform the chi-square test in R (see demo).

\scriptsize

```{r}
cc <- chisq.test(wtab,correct=FALSE)
cc
```

\normalsize

* At the 5\% level, there is statistical evidence of an association between hormone-replacement therapy and invasive breast cancer. 

## Continuity correction for $2 \times 2$ tables

* The continuity correction to the $X^2$ test improves
the $\chi^2$ approximation for
$2\times 2$ tables.
* The corrected version of the statistic is:
$$
X^2 = \sum_{\mbox{cells}} \frac{ \left( |\mbox{observed} - \mbox{expected}| - 0.5
\right)^2}{\mbox{expected}}
$$

* Please note that this correction applies only to $2\times 2$ tables and
does not generalize to $r \times c$ tables.

## Chi-square test with continuity correction for WHI example

\scriptsize

```{r}
cc <- chisq.test(wtab) #apply the default continuity correction
cc
```

\normalsize

* At the 5\% level, there is still evidence of an
association between hormone replacement therapy and invasive breast cancer.

## Expected counts


* As mentioned earlier, these are calculated under 
the null hypothesis of no association between the 2
variables in the table.

* Let's first discuss expected counts for the WHI 
example. Later, we'll generalize to arbitrary $r\times c$ tables.

\scriptsize

```{r,echo=FALSE}
wtab
```

\normalsize


* If $H_0$ holds and HRT has no effect on breast cancer,
the proportion of BC$+$ in each EP group should be the same and can be estimated
by pooling: $$\hat{p} = \frac{122+166}{7980+122+8340+166} = \frac{288}{16608}=0.01734.$$

## 

* Here is the table of counts with column and row margins added:

\scriptsize

```{r,echo=FALSE}
mytab=cbind(wtab, apply(wtab,1,sum))
colnames(mytab)<-c(colnames(wtab),"rowTot")
mytab2=rbind(mytab,apply(mytab,2,sum))
rownames(mytab2)<-c(rownames(mytab),"colTot")
mytab2
```

\normalsize


* Focusing on the 1st row of the table, under $H_0$ of no association, we expect that, of the 8102 women who are EP$-$,
    + $8102 \times  \hat{p} = 140.5$ would be BC$+$, and 
    + $8102 \times (1-\hat{p})=7961.5$ would be BC$-$.

* Similarly, focusing on the 2nd row, we expect that, under $H_0$ of no association, of the 8506 women who are EP$+$,
    + $8506 \times \hat{p} = 8506 \times \frac{288}{16608} = 147.5$ would be BC$+$, and 
    + $8506 \times (1-\hat{p})=8506 \times \frac{16320}{16608}=
    8358.5$ would be BC$-$.


## Expected counts, notation

* Notation from the text: 
\begin{center}
\begin{tabular}{ll|c|c|r}
& \multicolumn{1}{c}{}& \multicolumn{2}{c}{BC} & \\
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{BC$+$} & \multicolumn{1}{c}{BC$-$} & \\ \cline{3-4}
EP & EP$+$ & $a$ & $b$ & $a+b$ \\ \cline{3-4}
& EP$-$ &  $c$ & $d$ & $c+d$ \\ \cline{3-4}
&\multicolumn{1}{c}{} & \multicolumn{1}{c}{$a+c$} & \multicolumn{1}{c}{$b+d$} & $n$ \\
\end{tabular}
\end{center}
where $n=a+b+c+d$

* The pooled estimate of the proportion of BC+ women is $\hat{p} = (a+c)/n$.
* Expected count for the EP$+$ and BC$+$ cell: 
    +  Of the $(a+b)$ women who are EP$+$, we expect that $(a+b)\times \hat{p} =
(a+b)(a+c)/n$ would be BC$+$

* Notice that the expected count is of the form
    row total ($a+b$) times column total ($a+c$) 
    divided by table total ($n$). This is a generalizable pattern $\ldots$


## Expected counts: general formula

* For 
$r\times c$ tables, the expected count for the cell
in the $i$th row and $j$th column is the $i$th row total
times the $j$th column total divided by table total.

## Accuracy of the $\chi^2$ approximation ($r \times c$ tables)

* The $\chi^2$ approximation for the null distribution of the test statistic is considered
accurate when 
    1. No more than 20\% of cells have expected counts $<5$, and
    2. All expected cell counts are $\geq 1$.
* Note: These rules-of-thumb are intended regardless
of whether or not we use the continuity 
correction for $2\times 2$ tables.

## Accuracy of the $\chi^2$ approximation in WHI example

* The expected cell counts are as follows:

\scriptsize

```{r}
cc <- chisq.test(wtab)
with(cc,expected)
```

\normalsize

* _All_ expected cell counts are greater than 5, 
and so the $\chi^2$ approximation
is considered accurate.

## Sampling

* The chi-square test is appropriate under different sampling schemes
such as:
    1. Take simple-random samples (SRSs) from each of $c$ populations and classify
individuals in each SRS according to one categorical variable with $r$ levels
    2. Take one SRS from a single population and classify individuals according to two 
categorical variables, one with $c$ levels and the other with $r$ levels

* The first scheme includes 
case-control sampling ($c=2$)
    + e.g. an SRS of size $n_1=500$ from the case population for non-Hodgkin lymphoma and an SRS of size $n_2=500$ from the control population and then classify them according to whether
or not they are exposed to some pesticide ingredient.

* The second scheme pertains to the WHI study 
    + A sample of size $n=16608$ was drawn from the population of post-menopausal women and then cross-classified according to whether or not they were 
randomized to receive HRT and whether or not they developed invasive breast cancer.

