---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 15, part 1: Contingency Tables'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## Contingency Tables 

* Example: Data of [Mungan et al. 2000](https://www.ncbi.nlm.nih.gov/pubmed/10699612) \tiny click
\normalsize on 21,737 bladder cancer patients
    + Two categorical variables: **gender**, which has 2 levels,
    and **cancer stage**, which has 4 levels.


\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/mung.csv")
Mungan <- read.csv(uu)
head(Mungan,n=8)
```



## Tabulating Data in R

* Two useful functions in R for cross tabulating:
    + `table()` - easiest to use and
    + `xtabs()` - more flexible, but flexibility only needed when we
    have more than two variables -- see Chapter 16

\scriptsize

```{r}
table(Mungan)
```


## Row and Column Margins

* In R, the **row margin** is the tabulation of the row variable and 
the **column margin** is the tabulation of the column variable. 
    + The row margin is indexed by the number 1 and the
    column margin by the number 2.

\scriptsize

```{r}
mtab <- table(Mungan)
margin.table(mtab,1) # R's row margin
margin.table(mtab,2) # R's column margin
```

<!-- NTS: Row margin is marginal distn of the row variable
and col margin is marginal distn of the column variable.
-->

## Cancer stage distribution given gender

\scriptsize

```{r}
mtab <- table(Mungan)
mtab
```
\normalsize

* For each gender category, we can divide the 
counts in each row by
    the row total to get proportions.
    
\scriptsize

```{r}
prop.table(mtab,margin=1) # margin=1 refers to dividing by row totals
```

## Gender distribution given cancer stage 

* Likewise, for each cancer stage category we can divide
the counts in each column by the column total 
to get proportions.

\scriptsize

```{r}
prop.table(mtab,margin=2)
```


<!--
## Software Notes

* `prop.table()` requires a data type known as a `matrix` as input. 
* Sometimes we are given the cross-tabulated data and want
to enter the table into R ourselves as a `data.frame`.
* Then need to "coerce" the `data.frame` to a `matrix`:

\scriptsize

```{r}
mydf <- data.frame(Stage1=c(3926,12418), Stage2=c(402,995),
                    Stage3=c(356,883), Stage4=c(852,1905)) 
row.names(mydf) <- c("Female","Male")
mymat <- as.matrix(mydf) # coerce data.frame to matrix
mymat 
prop.table(mymat,margin=2)
```

## Software Notes, cont.

* DIY proportions: It is easy to compute with columns of a data frame; e.g.,

\scriptsize

```{r}
library(dplyr)
mydf <- 
  mutate(mydf,
    rTot=Stage1+Stage2+Stage3+Stage4, # row totals
    p1=Stage1/rTot,p2=Stage2/rTot,p3=Stage3/rTot,p4=Stage4/rTot) # row propns
select(mydf,p1,p2,p3,p4) # select() selects named columns of a data frame
```
-->