---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Demo for Chapter 15, part 4: Inference for Odds Ratios'
author: "Jinko Graham"
date: '2018-10-06'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## Testing whether $OR=1$

* The chi-square test assesses the null hypothesis
that $OR=1$ (no association between exposure and disease)
against the alternative hypothesis that $OR\not=1$ (an association). 

\scriptsize

```{r}
mydf <- data.frame(case=c(1350,7),control=c(1296,61)) # Doll and Hill's data
rownames(mydf) <- c("smoker","non-smoker")
mydf
chisq.test(mydf)
```

## Using R to get point and interval estimates of the OR. 


* For a $2\times 2$ table, the calculations are easy enough
to do with a hand calculator or spreadsheet.
* Below we present some R code that illustrates
data frame manipulation in R.
    + This will give us code 
    that will generalize to the case of multiple exposures,
    as in the Doll and Hill dataset where smoking status
    has four levels.
    + If you are not interested in learning R, 
    don't struggle to understand each line of code.
    + Instead focus on understanding the *purpose* of each 
    line of code. 
        + There will be similar code in assignment 2 and you
    will need to understand the steps to make
    simple modifications as necessary.
    
## Add Odds of Case to Data Frame

\scriptsize
 
```{r}
mydf <- data.frame(group=c("smoker","non-smoker"),
                   case=c(1350,7),
                   control=c(1296,61)) # Doll and Hill's data
library(dplyr) # for the mutate() function
mydf <- mutate(mydf,total = case+control,prcase  = case/total,
               odds = prcase/(1-prcase))
mydf
```

\normalsize

* `mutate()` is used to create new variables from existing ones
and add them to our data frame.
* In this example, the variables `total`, `prcase` and
`odds` are created and added to `mydf`.
    + Notice that the calculation of `odds` can use
    the newly-created variable `prcase`.

## Extract the Baseline Odds

* As a baseline for comparison, we will use the group of non-smokers.
Let's extract their estimated odds of lung cancer.

\scriptsize

<!--
```{r}
odds0 <- mydf %>% filter(group=="non-smoker") %>% 
  select(odds) %>% unlist()
```
-->

```{r}
odds0 <- mydf[mydf$group=="non-smoker","odds"]
odds0
```

\normalsize

* We can grab elements of a dataframe by referencing
the desired rows and columns inside square brackets.
* E.G, `mydf[1,2]` will grab the element in the 
1st row and 2nd column of `mydf`.
* Above, we indicate the desired row with 
the logical condition `mydf$group=="non-smoker"`,
    + `mydf$group` refers to the variable `group` in the 
    data frame `mydf`. 
    + Grab only the row for the non-smoking group.
* The desired column has the variable name `odds`.

## Extracting Other Baseline Data

* Extract the number of non-smoker cases and controls
analogously.

\scriptsize

```{r}
c <- mydf[mydf$group=="non-smoker","case"] 
c
d <- mydf[mydf$group=="non-smoker","control"] 
d
```

## Add the SE for the log-OR

\scriptsize

```{r}
mydf <- mutate(mydf, OR=odds/odds0, 
               se.logOR = sqrt(1/case + 1/control + 1/c + 1/d))
mydf
```

\small

## Add the CI

\scriptsize

```{r}
critval <- qnorm( 0.025, lower.tail=FALSE)
mydf <- mutate(mydf, 
               lowerCI=round(exp(log(OR) - critval*se.logOR),3), 
               upperCI=round(exp(log(OR) + critval*se.logOR),3))
mydf <- mutate(mydf, prcase=round(prcase,3), 
               odds=round(odds,3), OR=round(OR,3),
               se.logOR=round(se.logOR,3))
mydf
```

## 

* The `se.logOR` and CI for the non-smokers are not defined because we are using the non-smokers as the baseline group in our calculations.
    + Technically, the `se.logOR` is 0 and the CI is exactly 1, by definition.
* As a result, we set these to be the missing data code `NA` 
in the baseline group:

\scriptsize

```{r}
mydf[mydf$group=="non-smoker",
     c("se.logOR", "lowerCI", "upperCI")] <- 
  c(NA, NA, NA)
mydf
```

\normalsize

* Notice how we referenced multiple columns at once
with `c("se.logOR","lowerCI","upperCI")`.


## More Than Two Exposure Levels

* Doll and Hill's data with smokers classified by the average number
of cigarettes per day:
\begin{center}
\begin{tabular}{llcc} 
 & &  case & control \\ \hline
Number of   & 25+    & 340 & 182  \\
cigarettes  & 15-24  & 445 & 408  \\ 
per day     & 1-14    & 565 & 706  \\ 
            & 0      & 7 & 61  \\ \hline
\end{tabular}
\end{center}
* Can use the last row with 0 cigs per day (unexposed) as a baseline group, and calculate our ORs for 
each level of exposure. 

* Here is where the R code we wrote can pay off. We 
essentially repeat the code, but now refer to the baseline
group as `"0"` instead of `"non-smoker"`.

##

\scriptsize

```{r}
mydf <- data.frame(group=c("25+", "15-24", "1-14", "0"),
                   case=c(340,445,565,7),
                   control=c(182,408,706,61))

library(dplyr)
mydf <- mutate(mydf,total = case+control,
               prcase  = case/total,odds = prcase/(1-prcase))
odds0 <- mydf[mydf$group=="0","odds"]
c <- mydf[mydf$group=="0","case"]
d <- mydf[mydf$group=="0","control"]
mydf <- mutate(mydf, OR=odds/odds0, 
               se.logOR = sqrt(1/case + 1/control + 1/c + 1/d))
critval <- qnorm( 0.025, lower.tail=FALSE)
mydf <- mutate(mydf, 
             lowerCI = round(exp(log(OR) - critval*se.logOR),3), 
             upperCI = round(exp(log(OR) + critval*se.logOR),3))
mydf <- mutate(mydf, prcase=round(prcase,3), odds=round(odds,3), 
               OR=round(OR,3), se.logOR=round(se.logOR,3))
mydf[mydf$group=="0",c("se.logOR", "lowerCI", "upperCI")] <- 
  c(NA, NA, NA)
```

##

\scriptsize

```{r}
mydf
```

