---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 15, part 2: Chi-Square Tests'
author: "Jinko Graham"
date: '2018-10-06'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## Association and the WHI data

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/whi.csv")
WHI <- read.csv(uu)
wtab <- table(WHI)
```

* The table of proportions below gives the conditional distributions of BC status given EP status.

\scriptsize

```{r}
prop.table(wtab,margin=1)
```

## Categorical Variables in R

* R calls a categorical variable a ```factor``` 
and refers to its categories as ```levels```.

* For example, the categorical variables EP and BC, for hormone replacement therapy and breast cancer status, respectively, are called ```factors``` by R.

* When we cross-tabulate factors,
R chooses the order of the columns and rows in our table.
    + The order is set by the order of the categories in the 
    EP and BC factors.
    + Generally the categories, or levels, of a factor
    are ordered alphabetically

\scriptsize

```{r}
wtab
```

\normalsize

  
## Chi-square test for WHI example

\scriptsize

```{r}
cc <- chisq.test(wtab,correct=FALSE)
cc
```

\normalsize

* The argument `correct=FALSE` specifies that we 
do **not** want to do a continuity correction.

## Another way: use a dataframe rather than a table. 

* Say that we are not given the data; all we have to work with are the counts:
\begin{center}
\scriptsize
\begin{tabular}{ll}
EP+, BC+ 166 & EP+, BC$-$ 8340 \\
EP$-$, BC+ 122 &  EP$-$, BC$-$ 7980
\end{tabular}
\end{center}
\normalsize

* Then we can do the chi-square test as follows:

\scriptsize

```{r}
mydf <- data.frame(BCpos=c(166,122),BCneg=c(8340,7980)) #WHI data
rownames(mydf)=c("EP+", "EP-") 
mydf
chisq.test(mydf, correct=FALSE)
```


## Chi-square test with continuity correction for WHI example

\scriptsize

```{r}
cc <- chisq.test(wtab) #apply the default continuity correction
cc
```


## Expected Cell Counts

* The expected cell counts under the null hypothesis
of no association can be extracted from 
the output of `chisq.test()`.
* In general, you can find the names of an R object 
with `names()` and extract components with `with()`.

\scriptsize


```{r}
names(cc)
with(cc,expected)
```
