---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 15, part 1: Contingency Tables'
author: "Brad McNeney"
date: '2019-01-31'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)
```


## Contingency Tables 

* Contingency tables display
the joint frequency distribution of two categorical variables.
* E.G.: Let's consider the data of [Mungan et al. 2000](https://www.ncbi.nlm.nih.gov/pubmed/10699612) \tiny click
\normalsize on 21,737 bladder cancer patients
    + Two categorical variables: **gender**, which has 2 levels,
    and **cancer stage**, which has 4 levels.
    + The first few lines of the data file are as follows:


\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/mung.csv")
Mungan <- read.csv(uu)
head(Mungan,n=8)
```

## 

* The contingency table made by cross-tabulating
the gender and cancer stage variables of the Mungan data is as follows:

```{r}
table(Mungan)
```


## Terminology: Cells of a Table

* The cells of the table are its entries.
    + In the table cross-tabulating the gender and cancer 
    stage variables of the Mungan data, the first cell of the table is 3926
    
    
\scriptsize

```{r}
table(Mungan)
```

## Terminology: Row and Column Variables

* The row variable in a table defines the rows, the 
column variable the columns. 
    + In the table below, the row variable is `Gender` 
    and the column variable is `Cancer.Stage`.
    
\scriptsize

```{r}
table(Mungan)
```

## Terminology: Row and Column Margins

* The **row margin** is the tabulation of the row variable and 
the **column margin** is the tabulation of the column variable. 
* For the Mungan data, 
    + the row margin (tabulation of Gender) is 5536 and 
    16201 Females and Males, respectively
    + The column margin (tabulation of Cancer Stage) is
    16344, 1397, 1239, 2757 for cancer stages I through IV,
    respectively.
* Exercise: verify these table margins yourself.

## Adding Margins to a Table

* It is common practice to add margins to a contingency table.
* In the following, the row margins (first table) and 
column margins (second table) have been added:

```{r}
mtab <- table(Mungan)
cbind(mtab, Total = margin.table(mtab,1)) # Add row margin to the table
mtab <- table(Mungan)
rbind(mtab, Total = margin.table(mtab,2)) # Add col margin to the table
```

<!-- NTS: Row margin is marginal distn of the row variable
and col margin is marginal distn of the column variable.
-->

## Conditional distribution of cancer stage given gender

\scriptsize

```{r}
mtab <- table(Mungan)
mtab
```
\normalsize

* For each gender category, we can divide the 
counts in each row by
    the row total to get proportions.
    
\scriptsize

```{r}
prop.table(mtab,margin=1) # margin=1 refers to dividing by row totals
```

\normalsize

* This gives an estimate of the distributions of cancer stage within each gender.
    
## Conditional distribution of gender given cancer stage 

* Likewise, for each cancer stage category we can divide
the counts in each column by the column total 
to get proportions.

\scriptsize

```{r}
prop.table(mtab,margin=2)
```


\normalsize

* This gives an estimate of the distributions of gender within each cancer stage.

<!--
## Software Notes

* `prop.table()` requires a data type known as a `matrix` as input. 
* Sometimes we are given the cross-tabulated data and want
to enter the table into R ourselves as a `data.frame`.
* Then need to "coerce" the `data.frame` to a `matrix`:

\scriptsize

```{r}
mydf <- data.frame(Stage1=c(3926,12418), Stage2=c(402,995),
                    Stage3=c(356,883), Stage4=c(852,1905)) 
row.names(mydf) <- c("Female","Male")
mymat <- as.matrix(mydf) # coerce data.frame to matrix
mymat 
prop.table(mymat,margin=2)
```

## Software Notes, cont.

* DIY proportions: It is easy to compute with columns of a data frame; e.g.,

\scriptsize

```{r}
library(dplyr)
mydf <- 
  mutate(mydf,
    rTot=Stage1+Stage2+Stage3+Stage4, # row totals
    p1=Stage1/rTot,p2=Stage2/rTot,p3=Stage3/rTot,p4=Stage4/rTot) # row propns
select(mydf,p1,p2,p3,p4) # select() selects named columns of a data frame
```
-->

## Independence of Row and Column Variables.


* Suppose the conditional distribution of gender given cancer stage 
is 25% female and 75% male, regardless of cancer stage.

\bigskip

* What is the unconditional distribution of gender in this case (i.e., ignoring cancer stage)?

## 

* If the conditional gender distribution is 
    25% female and 75% male in each cancer stage then, ignoring
    cancer stage and considering the unconditional distribution of gender, there will also be 25% females and 75% males.

\bigskip

* In this case, we say that gender and cancer stage are 
_independent_.

## More generally

* If the conditional distributions of the row variable 
given the column variable are all the same,
they will also be the same as
the unconditional distribution of the row variable.
    + E.G., if the conditional gender distribution is 
    25% female and 75% male in each cancer stage, we will
    also have 25% females and 75% males unconditionally (i.e. ignoring cancer stage).
* We say that the column and row variables are
_independent_ because:
    + Knowing the value of the column variable 
    tell us nothing about the row
    variable; 
    + E.G. Knowing cancer stage tells us nothing about
    gender; so 
    $P({\rm Gender=Female}\mid {\rm Stage=I}) = P({\rm Gender=Female})$

## 

* One can use the definition of conditional probability to show that independence of row and column variables is 
equivalent to the following two statements:
    1. The conditional distributions given the different levels of the row variable are all equal
    2. The conditional distributions given the different levels of the column variable are all equal.
* The opposite of independence is dependence, or an
_association_.
* We next discuss how to test for association.
    