---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 16: Multiple $2\times 2$ Contingency Tables'
author: "Jinko Graham"
date: '2018-10-06'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## Associations and extraneous variables

* Want to study the association between two
dichotomous variables: an exposure variable
and a disease outcome.


* However, a third, extraneous variable with $g$ different values (levels) is also present
    + Does ignoring it lead to incorrect conclusions?

* When associating the disease outcome and the exposure variable, need to account for the possibility 
that the extraneous variable is either
    1. an **effect modifier** or 
    2. a **confounder**.
   
## Example: Aortic Stenosis (AS) Study

* What is the association between AS and smoking? How does sex affect this association?  
    + AS: disease outcome 
    + smoking: exposure variable
    + sex: extraneous variable with $g=2$ levels.

* First few lines of the data:

\scriptsize

```{r,echo=FALSE}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/sten.txt")
stenosis <- read.table(uu,header=TRUE)
head(stenosis)
```

\normalsize

* When associating AS and smoking, we should account for the possibility 
that sex is either
    1. an **effect modifier** or 
    2. a **confounder**.

#  Effect Modification

## Effect Modification

* When the ORs for different strata of the extraneous variable are different, 
the extraneous variable is an 
**effect modifier**.

* Inference (estimates, CIs, tests)
    of these different ORs must then be based on 
    the stratified tables.

* That is, we estimate the ORs, perform tests, and 
    construct CIs with each stratum-specific table, separately.
    
##  Testing for Effect Modification

* When the ORs for different strata of the extraneous variable are the same, 
they are said to be  **homogeneous**.

* Can test for this with.
    + $H_0$: homogeneity of ORs, versus
    + $H_a$: effect modification. 

* If we reject $H_0$, declare the extraneous variable to be 
an effect modifier.

## Test of Homogeneity (Text, Section 16.2.1)
* Suppose the extraneous variable has $g$ levels or **strata**
* Let $\widehat{OR}_i$ be the estimated OR between
the exposure and the disease outcome in the
$i$th stratum of the extraneous variable.
* Let $y_i = \log_e \widehat{OR}_i$, with variance estimated by $$se^2(\log_e \widehat{OR}_i)=1/a + 1/b + 1/c + 1/d.$$
* Let $w_i = 1/se^2(\log_e \widehat{OR}_i)$.
    + The larger the $se(\log_e \widehat{OR}_i)$,
    the smaller the weight $w_i$ accorded to stratum $i$ of the extraneous variable. 
* A weighted average of the $y_i$'s that diminishes the contribution of values with high standard errors is:
$$
Y = \sum_{i=1}^g \left ( \frac{w_i}{\sum_{j=1}^g w_j} \right ) y_i
=
\frac{\sum_{i=1}^g w_i y_i}{\sum_{j=1}^g w_j}.
$$

##
* The test statistic tallies up the deviations of the estimated log-ORs from this average:
$$
X^2 = \sum_{i=1}^g w_i(y_i - Y)^2 
$$
Large deviations suggest heterogeneous ORs.

* Under $H_0$: homogeneous ORs, the
statistic $X^2$ has
a chi-square distribution with $g-1$ df.


## Test of Homogeneity in the AS Example

* In the AS example, the gender-specific tables are:


\scriptsize

```{r,echo=FALSE}
xtabs(~AS+smoke+sex, data=stenosis)
```

\normalsize

## 

* We test for homogeneity of the ORs for AS across the 
strata defined by sex at level 5\%.

* We get (see R demo):

\scriptsize

```{r, echo=FALSE}
stratumStats <- function(a,b,c,d) {
  y<-log(a*d/(b*c))
  w <- 1/(1/a+1/b+1/c+1/d)
  return(list(y=y,w=w))
}
s1 <- stratumStats(47,19,29,14)
s2 <- stratumStats(20,24,25,37)

homogStat <- function(ss1, ss2) {
  Y <- (ss1$w*ss1$y + ss2$w*ss2$y)/(ss1$w+ss2$w)
  X2 <- ss1$w*(ss1$y-Y)^2 + ss2$w*(ss2$y-Y)^2
  p <- pchisq(X2,df=1,lower.tail=FALSE)
  return(list(X2=X2,pval=p))
}
homogStat(s1,s2)
```

\small
\vspace*{-0.2in}

* Retain the null hypothesis of homogeneous ORs at 5\% level.  
Insufficient evidence that sex modifies the effect of 
smoking on AS.
    + See text, section 16.2.1 for another example.

* Next question (and topic): Sex doesn't appear to be an effect modifier but is it a confounder? 
    
# Confounding


## What is a Confounder?

* A confounding variable is an extraneous variable
that is associated with both the disease outcome and the exposure of interest.

* Why is this important?

* Ignoring a confounder leads to biased inference of the effect of the exposure on the disease outcome (i.e., incorrect estimate, test and CI) and potentially **wrong conclusions**. 

* Confounding is thus a **big concern** in any observational study. 
    + E.G. Confounding by health-care coverage
    confused researchers in the observational
    studies leading up to the WHI.
    + Women with better health-care coverage tended to have better disease outcomes and also to have better access to HRT. Hence HRT seemed to be associated with better disease outcomes.
    
## Simpson's Paradox

* Confounding explains "Simpson's paradox":
    + When the effect of an exposure on the odds of the disease outcome differs in the stratified and pooled
analyses.

* In the AS example, the estimated ORs summarizing the association between AS and smoking, stratified by
sex, are about the same: 1.19 for females and 1.23 for males.
    + These results suggest that, for both females and males, smoking _slightly_ increases the odds of AS.

* However, the estimated OR from the pooled table across genders is 1.47.
    + This result suggests a _stronger_ effect of smoking than
    we obtain from the stratified tables!
    

## Effect Modifier vs. Confounder

* When considering the OR summarizing the association between the 
disease outcome and the exposure variable, the extraneous variable is:

    1.  an _effect modifier_ if the ORs for different strata of the extraneous variable are **different**.
    2.  a _confounder_ if the ORs for different strata of the extraneous variable are the **same** 
        but different from the OR that we get when we ignore the extraneous variable altogether.


## The Mantel-Haenszel (MH) estimator

* Need to adjust the OR for the effects of a confounding variable. 

* The Mantel-Haenszel (MH) estimator of the OR adjusts for a confounding variable when the
    disease outcome and exposure variable are dichotomous (i.e. have 2 categories)
    and the confounding variable is categorical with 2 or more categories.

## MH Summary OR (Section 16.2.2)

* The MH summary estimator is a weighted average of the stratum-specific OR estimators
    + The MH estimator was developed over 50 years ago, 
    is common practice in Epidemiology, and has good statistical
    properties.
    + Hence it is worth knowing.
* The MH estimate of the OR is
$$
\widehat{OR} = \frac{\sum_{i=1}^g a_i d_i/T_i}{\sum_{i=1}^g b_i c_i/T_i}
$$
where $a_i$, $b_i$, $c_i$, $d_i$ are from the table for the $i$th stratum of the extraneous variable, and $T_i$ is the table total for this $i$th table.
* See the text, page 381 for a worked example of this calculation.


## The MH Test
* Once we have the MH estimate of the common OR across the strata of the confounding variable, we may test the null hypothesis that this common OR is one.
    + i.e., that there is no association between the disease outcome
    and the exposure variable after adjusting for the confounding variable.

* The text, section 16.2.3, goes into the details
of the derivation, if you are interested. 

* Instead, we'll use R to apply this test (see R demo). 
    

## MH Test on AS Example

\scriptsize

```{r, echo=FALSE}
with(stenosis,mantelhaen.test(x=smoke,y=AS,z=sex,correct=FALSE,
                              conf.level=0.95))
```

\normalsize

* Referring to the p-value reported in the output, we see that, at the 5% level, there is no statistical evidence that smoking affects the odds
of AS, once we adjust for sex.


## Confidence Interval for the Common OR


* R returns a 95% CI of about $(0.69,2.1)$.

* Though not discussed, can extend the 
MH test to arbitrary values of the common OR.
    + We only covered the special case where the 
    value hypothesized is 1; i.e. $H_0: OR=1$ (unassociated) vs. $H_a: OR \not= 1$ (associated).
* The 95% CI is obtained by finding the set of all OR values 
that would be retained in a test at the 5% level.
    + E.G., With the AS data, if we were to test  
    $H_0:OR=0.69$ vs. $H_a:OR \not= 0.69$, we'd 
    retain $H_0$ at 5% level.
    + Similarly, if we were to test  
    $H_0:OR=2.1$ vs. $H_a:OR \not= 2.1$, we'd 
    retain $H_0$ at 5% level.
    + And the same for all values in between.
    + Any values smaller than 0.69 or larger than 2.1 for $H_0$ would lead to rejection with these data.
    
* Thus the 95% CI obtained from the AS data is
(0.69, 2.1). 
    + This is called "inverting the test" 
and is how R gets the CI (but not the text, which we'll skip). 

## Summary

* Inference of an association between an exposure variable and disease outcome should
consider extraneous variables that could be effect
modifiers or confounders.
* We first look for effect modification by testing for homogeneous ORs across the strata of the extraneous variable.
* If an extraneous variable is not an effect modifier, we may
adjust for it as a potential confounder by applying the MH estimator of the common OR.
    + The MH estimator automatically adjusts for the extraneous variable (but is only valid if there is no effect modification).
    +  In this class, we'll omit the material on the approximate CI for the common OR given in section 16.2.2 of the text and instead use the output from R which "inverts" the hypothesis test as described on the previous page.
