---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 16: Multiple $2\times 2$ Contingency Tables'
author: "Jinko Graham"
date: '2018-10-28'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## Example Data: Smoking, Aortic Stenosis and Sex   

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/sten.txt")
stenosis <- read.table(uu,header=TRUE)
head(stenosis)
```

\normalsize

* `AS` is 1 for patients with aortic stenosis (AS)
and 0 otherwise.
* `smoke` is 1 for smokers and 0 for non-smokers
* `sex` is 1 for males and 0 for females.


## Tabulating data in multiway tables

* Can use the `xtabs()` function to get tables of `AS` by `smoke` that are stratified by `sex`:

\scriptsize

```{r}
xtabs(~AS + smoke + sex, data=stenosis)
```

##

* Tables of `AS` by `smoke` without regard to `sex`
    + This is called the "pooled table", as opposed
    to the stratified tables from the previous slide.

\scriptsize

```{r}
xtabs(~AS + smoke, data=stenosis)
```


## Software Notes

* `xtabs()` uses R formulas; e.g., `~ AS + smoke + sex`.
* The order of the variables in the formula specifies the 
order of the margins. 
    + If you switched to, say, `~ AS + sex + smoke`, 
    you would have tables of disease by sex that are stratified by 
    smoking status.


## Test of Homogeneity in the AS Example

* Are the ORs for AS the same across the different sexes? 
Does `sex` modify the effect of smoking on AS?

* R has no built-in function or add-on packages (that I know of) for this test.  You can use one that I wrote. Below are the steps.
    1. For each stratum $i$ of the extraneous variable, get the stratum-specific log-ORs and weights,
    $y_i=\log_e(\widehat{OR}_i)$'s and $w_i=1/se^2(\log_e \widehat{OR}_i)$ using ```stratumStats()```
    2. Pass these stratum-specific statistics to ```homogTest()```,
    which combines them into the homogeneity test 
    statistic and returns a $p$-value for the test of homogeneity 
    of ORs across the strata.

## ```stratumStats()```

* Get the stratum-specific log-OR, $y_i=\log_e(\widehat{OR}_i)$,
and weight $w_i=1/se^2(\log_e \widehat{OR}_i)$,

\scriptsize
```{r}
stratumStats <- function(a,b,c,d) {
  y<-log(a*d/(b*c))
  w <- 1/(1/a+1/b+1/c+1/d)
  return(list(y=y,w=w))
}
```
\normalsize

* Apply ```stratumStats()``` to the first stratum defined by the 
extraneous variable ```sex``` (for females).

\scriptsize
```{r}
s1 <- stratumStats(47,19,29,14)
s1
```
\normalsize

## 

* Then apply ```stratumStats()``` to the 2nd stratum defined by the  ```sex``` (for males).

\scriptsize
```{r}
s2 <- stratumStats(20,24,25,37)
s2
```


\normalsize

*  The R objects ```s1``` and ```s2```
contain the $(y_i, w_i)$ pairs
for the 1st (females) and 2nd (males) stratum of ```sex```, respectively.


## ```homogTest()```

\small 

* homogTest() takes the stratum-specific log-ORs and weights 
in ```s1``` and ```s2``` as arguments and
returns the $X^2$ statistic for homogeneity, along with a $p$-value.

\scriptsize

```{r}
homogStat <- function(s1, s2) {
  Y <- (s1$w*s1$y + s2$w*s2$y)/(s1$w+s2$w)
  X2 <- s1$w*(s1$y-Y)^2 + s2$w*(s2$y-Y)^2
  p <- pchisq(X2,df=1,lower.tail=FALSE)
  return(list(X2=X2,pval=p))
}
```

\small
\vspace*{-0.2in}

* Apply ```homogStat()```:

\scriptsize

```{r}
homogStat(s1,s2)
```

\small
\vspace*{-0.2in}

* We retain the hypothesis of homogeneous ORs.  Insufficient evidence to claim that sex modifies the effect of smoking on AS.
    
# Confounding

## Simpson's Paradox and the AS Example

* As ```sex``` doesn't appear to be an effect
modifier, we can consider whether it is a confounder.

* Do this informally, by checking whether the ORs when we account for ```sex``` are different from the OR when we ignore ```sex```.

\scriptsize
```{r}
OR1<-exp(s1$y) #OR in females
OR1
OR2<-exp(s2$y) #OR in males
OR2
```

\normalsize

* The estimated ORs for smoking and AS, stratified by
sex are about the same: 1.19 for females and 1.23 for males.
    + This suggests that, for both females and males, smoking _slightly_ increases the odds of AS.


## 

* However, the estimated OR from the pooled table (when we ignore ```sex```) is 1.47.
    + This suggests a _stronger_ effect of smoking than
    we obtain from the stratified tables!
    
\scriptsize

```{r}
xtabs(~AS + smoke, data=stenosis)
a<-67; b<-43; c<-54; d<-51
ORpooled<-a*d/(b*c)
ORpooled
```

\normalsize 

* Simpson's paradox: When the effect of an exposure on the odds of the disease outcome differs in the stratified and pooled
analyses.
    
## The MH Common OR Estimate and Test

* The effect of smoking on AS appears to differ between the analyses
that stratify on versus ignore ```sex```.

* We therefore adjust for ```sex``` as a potential confounding variable
by getting the MH estimate of the common (adjusted) OR.

* Once we have the MH estimate of the common OR across the strata of the extraneous variable, we may test $H_0$ that this common OR is one.

* The text, section 16.2.3, goes into the details
of the derivation, if you're interested. 

* Instead, focus on using R to apply this test and interpreting the results.
    + R has a built-in function `mantelhaen.test()` that we can apply.
    

## MH Test applied to the AS Example

* Apply `mantelhaen.test()` to the 
`stenosis` data.
    + As the text doesn't use a continuity correction, to be consistent, we turn off the default setting in the function. 

\scriptsize

```{r}
with(stenosis,mantelhaen.test(x=smoke,y=AS,z=sex,correct=FALSE,
                              conf.level=0.95))
```

\normalsize

* Referring to the p-value reported in the output, we see that, at the 5% level, there is no statistical evidence that smoking affects the odds
of AS, once we adjust for sex.

