---
title: 'Statistics 305: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 17: Correlation'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## Example Data: Child Mortality by Country

* Data on child mortality (number of deaths before
age 5 years, per 1000 live births)
and percentage of children who are immunized for 
diptheria, pertussis and tetanus (DPT) from a random sample of 20 countries (see Table 17.2 of text).

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/mort.txt")
childmort <- read.table(uu,header=TRUE)
head(childmort)
```


## Scatterplots

* `ggplot()` used before to draw histograms
and boxplots.
    + When given quantitative variables it will draw 
    a scatterplot.
* The `aes()` specifies which variable to go on the 
`x` (horizontal) and `y` (vertical) axes.
* `geom_point()` adds the points to the scatterplot.

\scriptsize

```{r,fig.height=1.6,fig.width=3}
library(ggplot2)
ggplot(childmort,aes(x=PercentImmunized,y=Mortality)) +
  geom_point()
```

## Pearson Correlation in R

\scriptsize

```{r}
with(childmort,cor(PercentImmunized,Mortality))
```
\normalsize

* Sample Pearson correlation between `PercentImmunized` and `Mortality` is negative: $r=-0.791$.

## 

\bigskip

* Test if the Pearson population correlation differs from zero  
    +  i.e. Test $H_0: \rho=0$ vs. $H_a: \rho \not= 0$.

\scriptsize

```{r}
with(childmort,cor.test(PercentImmunized,Mortality))
```

\normalsize

\vspace*{-.1in}

* Reject $H_0$ at level 0.05; there is statistical evidence 
of correlation between a country's immunization rate and mortality.     
    + In particular, `Mortality` and `PercentImmunized` appear to be negatively correlated.

## Spearman Correlation in R

\scriptsize

```{r}
with(childmort,cor(PercentImmunized,Mortality,method="spearman"))
```

\normalsize

*  Note: Spearman's $r_s=-0.54$ is closer to zero
than Pearson's $r=-0.79$.

\bigskip

* Next, test whether the
population Spearman correlation coefficient differs
from zero;
i.e., test $H_0: \rho_s=0$
vs. $H_a:\rho_s \not= 0$.

## 

* Test Spearman's correlation coefficient:

\scriptsize

```{r}
with(childmort,cor.test(PercentImmunized,Mortality,
                    method="spearman",exact=FALSE))
```

\normalsize

\vspace*{-.2in}

* By default, ```cor.test()``` uses another null distribution than the $t$-distribution suggested by the text. 
    + To disable the default and use the $t$- distribution to get the $p$-value as in the text,  set ```exact=FALSE.```


## 

\scriptsize

```{r, echo=FALSE}
with(childmort,cor.test(PercentImmunized,Mortality,
                    method="spearman",exact=FALSE))
```

\normalsize

* There is statistical evidence that the population 
Spearman correlation is not 0 (at level $\alpha=0.05$).

* `Mortality` and `PercentImmunized` appear to be negatively correlated, even when we take into account outlying countries through the Spearman's rank-correlation test.
