---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 17: Correlation'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## Example Data: Child Mortality by Country

* Introduce ideas by focusing on an example data set.

* Data on child mortality (number of deaths before
age 5 years, per 1000 live births)
and percentage of children who are immunized for 
diptheria, pertussis and tetanus (DPT) from a random sample of 20 countries (see Table 17.2 of text).

\scriptsize

```{r,echo=FALSE}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/mort.txt")
childmort <- read.table(uu,header=TRUE)
head(childmort)
```


## Scatterplots

* For displaying a relationship between two quantitative variables.
* Each individual is represented by one point, comprised of 
a coordinate on the $x$-axis and a coordinate on the $y$-axis.

\scriptsize

```{r,fig.height=2,fig.width=3, echo=FALSE}
library(ggplot2)
ggplot(childmort,aes(x=PercentImmunized,y=Mortality)) +
  geom_point()
```


## Interpreting Scatterplots

* The **direction** of the relationship -- positive, negative, or
no relationship
* The **form** of the relationship -- patterns, such as linear or curved trends or even regional clusters in this particular example.
* The **strength** of the relationship -- how tightly the data fall around the apparent trend

* The child mortality data give the impression of a fairly weak,
negative linear relationship.
    + However, if we were to exclude the countries with 
    immunization rates $<$ 50% a pattern would not be obvious.

## Overview of Correlation

* Correlation is a measure of the strength of a **linear** 
(as opposed to curved, circular etc.) relationship between two quantitative variables.

* The most commonly-used measure of correlation is the 
Pearson correlation coefficient.

* Example trends with their Pearson correlation coefficients:
\vspace*{-.15in}
\begin{center}
\includegraphics[height=3in]{512pxCorrEgs.png}
\end{center}
\vspace*{-.07in}
\scriptsize Created by Denis Boigelot, 2011, and distributed under a CC-BY 2.0 license.

## Pearson Correlation Coefficient

* Suppose we have a simple random sample of data of size $n$, taken
from some larger population. 

* The Pearson correlation coefficient, $r$, is
a measure of the strength and direction of a _linear_ association between quantitative variables in the sample. 
* Always between $-1$ and $1$. Close to $\pm 1$ suggests a
strong linear relationship.
* Negative $r$ suggests a negative association. Positive $r$, a positive association.
* The Pearson correlation coefficient is:
$$
r = \frac{1}{n-1} \sum_{i=1}^n \left( 
\frac{x_i - \overline{x}}{s_x} \right) \left(
\frac{y_i - \overline{y}}{s_y} \right)
$$
(The formula can be simplified slightly -- see page 400 of the text.)

## Hypothesis Test of Correlation

* $r$ estimates the _population_ Pearson correlation, $\rho$,
which we can think of
as the Pearson correlation for the _entire population_ of (large) size $N$:
$$
\rho = \frac{1}{N-1} \sum_{i=1}^N \left( 
\frac{x_i - \mu_x}{\sigma_x} \right) \left(
\frac{y_i - \mu_y}{\sigma_y} \right)
$$
* The null hypothesis $H_0: \rho=0$ versus 
$H_a: \rho \not= 0$ can be tested with the test statistic
$$
t = \frac{r}{\sqrt{(1-r^2)/(n-2)}} = r \sqrt{\frac{n-2}{1-r^2}}
$$
which, under $H_0$, has an approximate $t$-distribution with $n-2$ df.

## Testing correlation

* Sample correlation between `PercentImmunized` and `Mortality` is negative: $r=-0.791$.


* Let's test if the population correlation differs from zero  
    +  i.e. Test $H_0: \rho=0$ vs. $H_a: \rho \not= 0$.

\vspace*{-.1in}

\scriptsize

```{r, echo=FALSE}
with(childmort,cor.test(PercentImmunized,Mortality))
```

\small

\vspace*{-.1in}

* According to the test, there is strong 
statistical evidence that the
population correlation is not 0.

* But could this be due to outlier countries, such
    as the three with low immunization rates? 

## 

* Pearson correlation coefficient is sensitive to outliers; e.g.:

\vspace*{-.15in}
\begin{center}
\includegraphics[height=3in]{CorrOutlier.png}
\end{center}
\vspace*{-.15in}
\scriptsize{\hspace{.3in} From Gravetter and Wallnau, 8th Ed.}

\normalsize

* An alternative measure that is less sensitive to outliers
is the Spearman Rank-Correlation coefficient.



## Spearman's Rank Correlation Coefficient

* To reduce the impact of outliers, replace the
values with **ranks**.
* For a sample $x_1, x_2, \ldots, x_n$, let $x_{r1},x_{r2}, \ldots,
x_{rn}$ denote the ranks; E.G. 
    + If $x_1$ is the $5$th-largest value in the ordered list of $x$'s, then its
rank is 5; i.e., $x_{r1}=5$. 
    + If $x_2$ is the $n$th-largest value (i.e. smallest) in the ordered list of $x$'s, then its rank is $n$; i.e. $x_{r2}=n$.
    + If $x_3$ is the $1$st-largest value (i.e. largest) in the ordered list of the $x$'s, then its 
    rank is $1$; i.e. $x_{r3}=1$, etc.
* **Spearman's rank-correlation coefficient** is the Pearson correlation 
of the **ranks**:
$$
r_s = \frac{1}{n-1} \sum_{i=1}^n \left( 
\frac{x_{ri} - \overline{x}_r}{s_{rx}} \right) \left(
\frac{y_{ri} - \overline{y}_r}{s_{ry}} \right), \;\; \mbox{where}
$$
$\bar{x}_r$, $\bar{y}_r$ are sample means and 
$s_{rx}$,  $s_{ry}$ are sample SDs of ranks.
* The text (pg 405) shows that this formula can be simplified to depend only on the 
differences between ranks of the $(x,y)$ pairs, but this is not our focus.


## Hypothesis Test of Spearman's Correlation

* The _population_ Spearman correlation coefficient, $\rho_s$, is 
the Spearman correlation for the entire population.
* To test $H_0: \rho_s=0$
vs. $H_a:\rho_s \not= 0$, about the presence of any rank-based correlation in the population, we can use the statistic
$$
t = \frac{r_s}{\sqrt{(1-r_s^2)/(n-2)}} = r_s \sqrt{\frac{n-2}{1-r_s^2}},
$$
which has an approximate $t$-distribution with $n-2$ df under $H_0$.
    + The text considers this approximation to be reliable for $n\geq 10$.

## Application to Child Mortality Data

\scriptsize

```{r,echo=FALSE}
with(childmort,cor(PercentImmunized,Mortality,method="spearman"))
```

\normalsize

* The sample Spearman correlation coefficient  of $-0.54$ is closer to zero
than the sample Pearson correlation coefficient of $-0.79$.

* Let's test to see if there's any evidence that the
population Spearman correlation coefficient differs
from zero.
    + i.e., test $H_0: \rho_s=0$
vs. $H_a:\rho_s \not= 0$.

* We have a sample of $n=20$ countries and so,
according to the text, approximating
the null distribution of the test statistic with t-distribution should be OK.

##

\bigskip

\scriptsize

```{r,echo=FALSE}
with(childmort,cor.test(PercentImmunized,Mortality,
                    method="spearman",exact=FALSE))
```

\normalsize

* There is statistical evidence that the population 
Spearman correlation differs from 0 (at level $\alpha=0.05$).
    + `Mortality` and `PercentImmunized` appear to be negatively correlated, even when outlying countries are taken into
    account through a rank-based correlation test.
