---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 18, part 3: Prediction intervals, $r^2$ and Residual Plots'
author: "Jinko Graham"
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## Load the data, fit the regression model, get CIs

* We'll continue working with the data on low-birthweight babies.

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu) #load the data
lfit <- lm(headcirc ~ gestage,data=lbwt) #fit the regression model
lCI <- predict(lfit,interval="confidence") #get the 95% CI
#Now make a dataframe of the results
lbwtCIs<-data.frame(gestage=lbwt$gestage, lCI )
head(lbwtCIs)
```

\normalsize

* Next, we'll compare these CIs to _prediction intervals_ $\ldots$.


## 95% PIs for example data

\scriptsize

```{r}
lPI <- predict(lfit,interval="prediction")
lbwtPIs <- data.frame(gestage=lbwt$gestage, lPI) 
head(lbwtPIs,n=3)
```

\normalsize

* Predicted values $\hat{y}$ and
lower and upper limits of PI are in
the columns `fit`, `lwr` and `upper`, respectively.

## PIs are wider than CIs

* E.G. Compare 
for gestational age 29 weeks:

\scriptsize
```{r}
lbwtCIs[1,] #95% CI for gestational age 29 weeks
lbwtPIs[1,] #95% PI for gestational age 29 weeks
```

\normalsize

* The CI **estimates a parameter**, namely the population-mean head circumference of a baby of gestational age 29 weeks. 
    + The CI is an interval ***estimate*** of a ***parameter***.

* The PI **predicts a future random value**, namely the head circumference of a new baby of gestational age 29 weeks. 
    + The PI is an interval ***prediction*** of a future ***random value***.


## $r^2$ in simple linear regression

* For the regression of head circumference on 
gestational age, we can
get $r^2$, the coefficient of determination as follows:


\scriptsize
```{r}
with(summary(lfit), r.squared)
```




## Plots of residuals vs. fitted values

* In the low-birthweight babies data, the regression of
head circumference on gestational age gives:

\scriptsize

```{r, fig.height=2.9,fig.width=3.5}
plot(lfit,which=1, add.smooth=FALSE)
```

\normalsize

## 

* R's `plot()` function will plot the regression-modelling 
output in `lfit`. 
   + Recall that we got `lfit` by applying the `lm()`
   function to the dataframe `lbwt` in
   \footnotesize 
   `lfit <- lm(headcirc ~ gestage,data=lbwt)`
   \normalsize
   
* For regression-model objects such as `lfit`
the `plot()` function has 6
different diagnostic plots, specified by the
`which` argument.
    + The first (`which=1`) is the plot of the residuals _vs_ fitted values.
* The resulting plot labels the three most extreme 
residuals (i.e. residuals farthest from zero) by their case number in the `lbwt` dataframe: 29, 31 and 33.

* These cases can be viewed in the context of the dataframe:

\scriptsize
```{r}
lbwt[c(29,31,33),]
```

