---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'R Demo for Chapter 18, part 2: Inference in Simple Linear Regression'
author: "Jinko Graham"
date: '2018-11-04'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## Load the Data and Fit the Regression Model

* Load the data on low-birthweight babies:

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_17/Data/lbwt.csv")
lbwt <- read.csv(uu)
```

\normalsize

* Fit the regression model and print out the fitted regression coefficients:

\scriptsize

```{r}
lfit <- lm(headcirc ~ gestage,data=lbwt)
coefficients(lfit)
```

\normalsize

* The regression parameters are estimated by $\hat{\alpha}=3.91$ (`intercept`) and $\hat{\beta}=0.78$ (`gestage`).

## Testing Example

* For the low-birthweight data, the model summary includes 
the p-value from the tests of $H_0: \beta=0$ _vs._
$H_a:\beta \not= 0$

\scriptsize

```{r}
summary(lfit)
```

\normalsize 

##  

\bigskip

* We're interested specifically in the `coefficients` component of the summary.


* Can extract the `coefficients` component with the `$` operator.

\scriptsize

```{r}
summary(lfit)$coefficients
```

\normalsize 

* The test statistic value is about 12.37 and the p-value is tiny.
* Strong statistical evidence for an association between gestational age and
head circumference.

## Software Notes

* The output of the `summary()` function includes a lot
of components that we are not yet ready for.
    + However 
    `summary(lfit)$coefficients` 
    extracts just
    the coefficients table.
* Statistics related to a particular
coefficient are in the row of the
table labelled by the name of the
explanatory variable.
    + E.G., Below the summaries related to the slope 
    of the regression line are in the row labelled `gestage`.
    
\scriptsize

```{r}
summary(lfit)$coefficients
```

\normalsize

## CI Example

* We can use the `confint()` function in R to extract a confidence interval.

\scriptsize

```{r}
confint(lfit,conf.level=0.95)
```

\normalsize

* The 95% CI for $\beta$ is about $(0.65,0.91)$. 

* Interpret:
"With 95% confidence, we estimate that a
  one-week increase in gestational age
  is associated with an increase in head circumference
  of between 0.65 to 0.91 cm."
    

## CIs at Observed Values of Explanatory Variable

* Use the `predict()` function to get predictions and confidence
intervals for each observed value of the explanatory variable.

* Default coverage probability or level for the CI is $C=0.95$

\scriptsize

```{r}
lpred <- predict(lfit,interval="confidence")
head(lpred)
```

\normalsize

* In a given row, 1st entry
is fitted value $\hat{y}$, 2nd entry is lower bound of CI and last entry is upper bound of CI.


* Attach `lpred` to the corresponding values for $y$ (`headcirc`) and $x$ (`gestage`), to create a new R object called `lbwtFits` $\ldots$

## 

\bigskip

\scriptsize

```{r}
lbwtFits <- data.frame(headcirc=lbwt$headcirc,gestage=lbwt$gestage,lpred)
head(lbwtFits)
```

\normalsize

* The values of the response, $y$,
are in the column `headcirc`.

* The values of the explanatory variable, $x$, are in the column `gestage`.

* The fitted values $\hat{y}$ are in the column `fit`.

* The lower limits of the CIs are in the column `lwr`
and the upper limits are in the column `upr`.


## 90% CIs at New Values of Explanatory Variable

* Suppose that we want 90% CI's at new values of the explanatory variable, such as `gestage` of 25.5 and 30.5 weeks.

* Create a dataset with the new values of the explanatory 
variables and pass this to `predict()`.
* Specify the level of the confidence interval with the `level` argument.


\scriptsize

```{r}
newdat <- data.frame(gestage = c(25.5,30.5))
predict(lfit,newdata = newdat, interval="confidence", level=.90)
```

\normalsize

* The fitted values $\hat{y}$ of `headcirc` for 
`gestage`s of 25.5 and 30.5 are in the column `fit` and are about 23.8 and 27.7, respectively.

* The lower limits of the 90% CIs are in the column `lwr`
and the upper limits are in the column `upr`.


## Adding CIs to a Scatterplot

* By default, pointwise 95% CIs around the fitted regression line are added by `ggplot() + geom_smooth(method="lm")`

\scriptsize

```{r,fig.height=2,fig.width=4}
library(ggplot2)
ggplot(lbwt, aes(x=gestage,y=headcirc)) + 
   geom_point() + 
   geom_smooth(method="lm")
```

\normalsize

##

* To change the default and add, e.g., pointwise 99% CIs around the fitted regression line, use the argument `level` in `geom_smooth()`: 

    + `geom_smooth(method="lm", level=.99)` for 99% CIs.

\scriptsize

```{r,fig.height=2,fig.width=4}
library(ggplot2)
ggplot(lbwt, aes(x=gestage,y=headcirc)) + 
   geom_point() + 
   geom_smooth(method="lm", level=.99)
```
