---
title: "Assignment 2, Question 4"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 15, #21, with modification. Beer-drinking questionnaires. Two different questionnaires addressing drinking habits
were administered to the same set of men and women aged 50-65 years from a certain community. The first questionnaire format concerned
general diet and lifestyle while the second format specifically targeted alcohol use. For each questionnaire, the responses were processed to classify a respondent as either a non-drinker or drinker of beer.  Categories for the general questionnaire are saved under the variable `genques` and those for the questionnaire targeting alcohol use are saved under the variable `alcques`.
The data are available as a plain text file; see the R code below for accessing it.

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/alcohol.txt")
alc <- read.table(uu,header=TRUE)
head(alc) 
```

a) Are the drinking statuses on the two different types of questionnaires associated? Use a 5\% level in your test. What do you conclude?


**Answer:**

```{r}
# Put your R commands here.
```

*Insert your written comments here.*

b) Is the type of questionnaire associated with drinking status? Use a 5\% level and continuity correction in your test. What do you conclude?

**Answer:**

```{r}
# Put your R commands here.
```

*Insert your written comments here.*
