---
title: "Assignment 2, Question 1"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 15, #10 but with modification. Survey of Italian physicians on breast cancer. Physicians in different specialties were asked about
their recommendations for radical (mastectomy) vs.
conservative (lumpectomy) surgery. In particular, 
would they recommend radical surgery regardless
of a patient's age (RR), conservative surgery for 
younger patients only (CY), or conservative surgery regardless of age (CR)?


To answer this question we need a data-frame 
that contains the results of the survey.  
Create one called `survey` with 3 
variables `RR`, `CY` and `CR` as follows:
```{r}
survey <- data.frame(
  RR=c(6,23,2,1,1), 
  CY=c(22,61,3,12,12), 
  CR=c(42,127,54,43,31))
row.names(survey)<-c("Internal", "Surgery", "Radiotherapy",
              "Oncology", "Gynecology")
survey
```

(a) At the 0.05 level, test the null hypothesis
that there is no association between physician
specialty and recommended treatment. (Note: Ignore
the warning message from R about the chi-squared approximation.)

**Answer:**

```{r}
# Add your R commands here.  
```

*Insert your written comments here.*

(b) Report the expected cell
counts in the table under the null hypothesis. Comment on whether the chi-square
approximation is reliable for these data.

**Answer:**

```{r}
# Add your R commands here.  
```

*Insert your written comments here.*


(c) What do you conclude?

**Answer:**

*Insert your written comments here.*

