---
title: "Assignment 2, Question 2"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 15, #12 but with modification. Intra-observer
variability in assessing cervical smears. This study
screened 3325 pap-smear slides for the presence or
absence of abnormal squamous cells. Each slide
was screened by a particular observer and then
rescreened 6 months later by the same observer.
The results of the study are shown below:

<table cellpadding="10">
<tr> 
 <td> &nbsp; </td>  
 <td colspan="2" align="center"> &nbsp; <b>2nd screen</b> &nbsp; </td>
 <td> &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; <b>1st screen</b> &nbsp; </td> 
 <td> &nbsp; Present &nbsp; </td> 
 <td> &nbsp; Absent &nbsp; </td> 
 <td> &nbsp; Total &nbsp; </td>
</tr>
<tr>
 <td> &nbsp; Present &nbsp; </td> 
 <td> &nbsp; 1763 &nbsp; </td> 
 <td> &nbsp; 489 &nbsp; </td> 
 <td> &nbsp; 2252 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Absent &nbsp; </td> 
 <td> &nbsp; 403 &nbsp; </td> 
 <td> &nbsp; 670 &nbsp; </td> 
 <td> &nbsp; 1073 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Total &nbsp; </td> 
 <td> &nbsp; 2166 &nbsp; </td> 
 <td> &nbsp; 1159 &nbsp; </td> 
 <td> &nbsp; 3325 &nbsp; </td> 
</tr>
</table>
<p>


(a) Do these data support the null hypothesis
that there is no association between the time of 
screening and diagnosis? 

Obtain the value of the test statistic 
with continuity correction on your hand calculator but below,
when you are asked to insert your written comments, be sure to provide the formula you used in R Markdown's math notation. E.G. Say the numbers were 102 and 137. Then in Markdown the test statistic would be written as
`$(|102 - 137|-1)^2/(102+137)=4.84$`. In your rendered RMarkdown
document to be turned in, this would show up nicely formatted.
 Note that you will still need an R code chunk to find the
 $p$-value.
 
 
 **Answer:**

```{r}
# Add your R commands here.  
```

*Insert your written comments here.*

The data could also be displayed in the following
manner:
<table cellpadding="10">
<tr> 
 <td> &nbsp; </td>  
 <td colspan="2" align="center"> &nbsp; <b>Screening</b> &nbsp; </td>
 <td> &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; <b>Abnormal cells </b> &nbsp; </td> 
 <td> &nbsp; First &nbsp; </td> 
 <td> &nbsp; Second &nbsp; </td> 
 <td> &nbsp; Total &nbsp; </td>
</tr>
<tr>
 <td> &nbsp; Present &nbsp; </td> 
 <td> &nbsp; 2252 &nbsp; </td> 
 <td> &nbsp; 2166 &nbsp; </td> 
 <td> &nbsp; 4418 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Absent &nbsp; </td> 
 <td> &nbsp; 1073 &nbsp; </td> 
 <td> &nbsp; 1159 &nbsp; </td> 
 <td> &nbsp; 2232 &nbsp; </td> 
</tr>
<tr>
 <td> &nbsp; Total &nbsp; </td> 
 <td> &nbsp; 3325 &nbsp; </td> 
 <td> &nbsp; 3325 &nbsp; </td> 
 <td> &nbsp; 6650 &nbsp; </td> 
</tr>
</table>
<p>


(b) Is there anything wrong with this presentation?
Please describe briefly.

**Answer:**

*Insert your written comments here.*

(c) If you didn't know there was anything wrong with the
presentation in part b, how would you analyze the data? For instance, what type of test would you apply and what would be the null hypothesis that you'd be testing? (Note: You don't actually have to do the test; just say what you would do.)

**Answer:**

*Insert your written comments here.*


