---
title: "Assignment 2, Question 3"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 15, #18. Alcohol and miscarriages. A study investigating the association between spontaneous abortion (miscarriage) and alcohol consumption obtained the following
data:

<table cellpadding="10">
<tr> 
 <td> <b>drinks/week</b> &nbsp; </td>
 <td> <b>pregnancies</b> &nbsp; </td>
 <td> <b>miscarriages</b> &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 0 &nbsp; </td>
 <td> &nbsp; 33164 &nbsp; </td>
 <td> &nbsp; 6793 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 1-2 &nbsp; </td>
 <td> &nbsp; 9099 &nbsp; </td>
 <td> &nbsp; 2068 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 3-6 &nbsp; </td>
 <td> &nbsp; 3069 &nbsp; </td>
 <td>  &nbsp; 776 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 7-20 &nbsp; </td>
 <td> &nbsp; 1527 &nbsp; </td>
 <td>  &nbsp; 456 &nbsp; </td>
</tr>
<tr> 
 <td> &nbsp; 21+ &nbsp; </td>
 <td> &nbsp; 287 &nbsp; </td>
 <td> &nbsp; 98 &nbsp; </td>
</tr>
</table>
<p>

For parts a-d below, you may **modify** this R code chunk as necessary (and then run it) if you find it helpful.
```{r, message=FALSE, warning=FALSE}
mydf <- data.frame(group=c("0", "1-2", "3-6", "7-20", "21+"),
                   preg=c(33164,9099,3069,1527,287),
                   mis=c(6793,2068,776,456,98))

library(dplyr)
mydf <- mutate(mydf,ok=preg-mis, prmis = mis/preg, odds = (prmis/(1-prmis)))
odds0 <- unlist(select(filter(mydf, group=="0"), odds))
a <- unlist(select(filter(mydf, group=="0"), ok))
b <- unlist(select(filter(mydf, group=="0"), mis))
mydf <- mutate(mydf, OR=odds/odds0, se.logOR = sqrt(1/a + 1/b + 1/ok + 1/mis))
C <- 0.80 #confidence level or coverage probability
critval <- qnorm( (1-C)/2, lower.tail=FALSE)
mydf <- mutate(mydf, lowerCI = round(exp(log(OR) - critval*se.logOR),3), upperCI =  
round(exp(log(OR) + critval*se.logOR),3))
mydf <- mutate(mydf, prmis=round(prmis,3), odds=round(odds,3), OR=round(OR,3), se.logOR=round(se.logOR,3))
mydf[mydf$group=="0",c("se.logOR", "lowerCI", "upperCI")] <- c(NA, NA, NA)
mydf
```

(a) For each level of alcohol consumption, estimate
the probability that a woman who becomes pregnant
will have a miscarriage.

**Answer:**

*Insert your written comments here.*

(b) For each category of alcohol use, estimate the
relative odds of miscarriage 
for women who consume some amount of alcohol versus
those who do not consume any.

**Answer:**

*Insert your written comments here.*

(c) In each case, calculate a 95\% confidence 
interval for the odds ratio

**Answer:**

*Insert your written comments here.*

(d) What do you conclude?

**Answer:**

*Insert your written comments here.*

