---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 20, part 2: Inference in Logistic Regression'
author: "Jinko Graham"
date: '2018-11-13'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/bpd.csv")
bpd <- read.csv(uu)
```

## Inference in Logistic Regression

* In logistic regression, the log-odds of the outcome
is modeled by the straight-line relationship $\alpha + \beta_1 X_1$.

* The intercept $\alpha$ 
is not typically of interest; instead we focus on
$\beta_1$ because it summarizes the effect of $X_1$ on $Y$. 
* It turns out that the sampling distribution of 
$\hat\beta_1$ is approximately normal with mean $\beta_1$ and 
SD that depends on $\alpha$ and $\beta_1$.
* Let  $SE(\hat\beta_1)$ denote the SE of 
$\hat{\beta}_1$, obtained
by inserting parameter estimates $\hat{\alpha}$ and $\hat{\beta}_1$
into the SD formula for $\hat{\beta}_1$.
* For large samples, the pivotal quantity has a standard normal distribution; i.e.,
$$ 
\frac{\hat\beta_1 - \beta_1}{SE(\hat\beta_1)} \sim N(0,1)
$$
* Hypothesis tests and CIs follow in the usual way.

## Dataset

* We'll be working with the `bpd` dataframe of low-birthweight 
babies from the neonatal ICU of a large hospital:

\scriptsize

```{r}
head(bpd)
```

\normalsize

* The `bpd` dataframe has a variable `bpd`
indicating whether the baby had bronchopulmonary dysplasia.
    + This condition results from damage to the lungs caused by a respirator and long-term use of oxygen. 
    + Most infants recover, but some may have long-term breathing difficulty.

## Logistic Regression of BPD on Birth Weight

\scriptsize

```{r}
bfit <- glm(bpd~birthwt,data=bpd,family=binomial())
summary(bfit)$coefficients
```

\small

\vspace*{-.15in}
* $\beta_1$ is the increase in the log-odds
of BPD associated with a one-gram increase in birthweight.

* If $\beta_1=0$, then the log-odds of BPD don't change with birthweight and so BPD and birthweight are not associated.

* To assess whether BPD is associated
with birthweight, we test the hypotheses that
$H_0:\beta_1 = 0$ _vs._ $H_a: \beta_1 \not= 0$

* The estimate of $\beta_1$ is $\hat\beta_1 = -0.0042$ with SE $0.00064$.

* The test statistic is 
$$ \frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)}
= \frac{-0.0042-0}{0.00064} = -6.6,$$   
which gives a tiny $p$value. Strong statistical evidence that BPD
is associated with birth weight.


## Approximate 95\% CIs 

* We will obtain 95\% CI's for the following:
    + $\beta_1$, the increase in the log-odds
of BPD associated with a one-gram increase in birthweight, and
    + $e^{\beta_1}$, the factor by which the odds of BPD changes
    with a one-gram increase in birthweight (an OR)

* An approximate 95\% CI for 
$\beta_1$ is 
$$\hat{\beta}_1 \pm z^* \! \times \! SE(\hat{\beta}_1) = \hat{\beta}_1 \pm 1.96\times SE(\hat{\beta}_1);$$
i.e., estimate $\pm$ margin of error.

* For the BPD data, the 95\% CI for $\beta_1$ is
$$ -0.0042 \pm 1.96\times 0.00064 =
(-0.0055,-0.0029)$$
    

* The 95\% CI for $e^{\beta_1}$
is obtained by exponentiating and so is
    $$(e^{-0.0055},e^{-0.0029}) = (.995,.997)$$
    
## But CIs from R are obtained differently

* The `confint()` function in R, when applied to a `glm()`-fitted object (such as `bfit`), gets the CIs for logistic-regression coefficients differently.

* Its CIs are based on inverting hypothesis tests: 
    + E.G., A 95% CI
    is the set of all $\beta_1$ values, $b$, retained
    in a test of $H_0: \beta_1=b$ _vs._ $H_a: \beta_1 \not= b$ 
    at the 5% level, with the data at hand.
    + Recall: Same approach used by `mantelhaen.test()` to get its CI (see Ch16 notes, pg 23).

\scriptsize

```{r}
confint(bfit) # CIs for alpha and beta1
exp(confint(bfit)["birthwt",]) # CI for OR parameter e^{beta1}
```


## Comparing CI methods

* In this example, the CI for $e^{\beta_1}$ from 
applying `confint()` to the `glm()`-fitted object
is (0.994,0.997), which is very similar to 
the CI of (0.995,0.997) 
form the pivotal-quantity method.
* Know how to extract what is needed from the coefficients summary 
below to calculate a CI for the logistic-regression coefficients
using the pivotal-quantity method.

\scriptsize

```{r}
summary(bfit)$coefficients
```

\normalsize

+ Need the coefficient estimate, its SE and the critical value $z^*$.
+ The critical value will be provided; for 95% CIs it is 1.96.

## Interpreting the CI

\small

* Also, know how to interpret the CI from `confint()`.
    + e.g., the `birthwt` row of `confint(bfit)` gives us the
    95% CI for $\beta_1$, the slope coefficient for `birthwt`. 

\scriptsize

```{r}
confint(bfit)["birthwt",]
```

\small

\vspace*{-.12in}
* To get the 95% CI for the OR, $e^{\beta_1}$,
exponentiate the above CI for $\beta_1$:

\scriptsize

```{r}
exp(confint(bfit)["birthwt",])
```

\small

\vspace*{-.14in}
* Interpretation: "With 95% confidence, we estimate that an increase
in birthweight of 1 gram is associated with a
change in the odds of BPD by a factor 
of between 0.994 and 0.997."
    

## Binary Explanatory Variable

\normalsize

* The `bpd` dataframe also has a column for
the toxemia status of the baby's mother
(1 if she was toxic and 0 if not)

\scriptsize

```{r}
head(bpd, n=3)
```

\normalsize

* Logistic regression of BPD on the binary 
variable toxemia:

\scriptsize

```{r}
bfit2 <- glm(bpd~toxemia,data=bpd,family=binomial)
summary(bfit2)$coefficients
```

## Testing for association between toxemia and BPD.

\scriptsize

```{r}
summary(bfit2)$coefficients
```

\normalsize

* The estimate of the `toxemia` 
coefficient is $\hat\beta_1 = -0.772$ with SE $0.482$.
\bigskip

* The test statistic for testing $H_0:\beta_1 = 0$ 
_vs._ $H_a:\beta_1 \not= 0$ is 
$$\frac{\hat{\beta}_1 - 0}{SE(\hat{\beta}_1)} =
\frac{-0.772- 0}{0.482} = -1.6,$$ 
which gives a reported $p$value of $p=0.11$.
+ At the 5\% level, there is insufficient statistical evidence 
    to conclude that BPD is associated with toxemia.

## Confidence Intervals for the Toxemia Effect

* Recall:  For a binary exposure, $X_1$, $e^{\beta_1}$
is the odds-ratio (OR) for exposed _vs._ unexposed groups.

* In this example, $e^{\beta_1}$ is the ratio 
of the odds of BPD given toxemia divided by the odds of BPD given no toxemia

\scriptsize

```{r}
confint(bfit2)["toxemia",] #CI for beta1, the log-OR
exp(confint(bfit2)["toxemia",]) # CI for OR
```

\normalsize

* An approximate 95\% CI for this OR 
is $(0.164,1.12)$: