---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 20, part 1: Logistic Regression Models'
author: "Jinko Graham"
date: '2018-11-25'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Introduction to Logistic Regression

* In logistic regression we study the effect of explanatory variables 
on the odds of a binary outcome.
* This is a generalization of the analyses of odds ratios we have
studied before.
* Think of the binary outcome $Y$ as disease status (0=non-disease; 1=disease).
* The explanatory variables $X_1, \ldots, X_q$ could be categorical (e.g., exposures), or
quantitative variables.


## Example Data

\small

* In 223 low-birthweight infants sampled from the 
neonatal ICU of a large hospital, 76 were diagnosed with
bronchopulmonary dysplasia (BPD; $Y=1$) and 
147 were not ($Y=0$).
* Birth weight ($X_1$) is believed to influence the risk of BPD. Grouping the birth weights into 3 categories, we obtain:
\footnotesize
\begin{center}
\begin{tabular}{r|cccc} 
birthweight(in g) &  BPD & no BPD & odds BPD & log-odds$^*$ BPD \\ \hline
0-950    & 49 & 19 & $49/19=2.58$ & $~~0.95$ \\
951-1350 & 18 & 62 & $18/62=0.29$ &  $-1.24$ \\ 
1351-1750&  9 & 66 & $~9/66=0.14$ & $-1.99$ \\  \hline
\multicolumn{5}{l}{$^*$ Use the natural logarithm.}
\end{tabular}
\end{center}
\small

* Consider the ratio of the odds at two values of $X_1$.
* To get the log of this odds ratio (the log-OR), we take the difference between the two log-odds.
* E.G. Consider the odds of BPD in babies with birthweight $<950g$ relative to babies with birthweight between $1351$-$1750g$. The log-OR is $0.95-(-1.99)=0.95+1.99=2.94$.


## The Logistic Regression Model

* We may model the log-odds of $Y=1$ (e.g. BPD) as a function of $X_1$ (e.g. birthweight):
$$
\log \left[ \frac{p}{1-p} \right] = \alpha + \beta_1 X_1, \quad {\rm where}
$$
    + log is the _natural logarithm_ and 
    + $p$ is the probability of $Y=1$ given $X_1$.
    
* Let $LO = \alpha + \beta_1 X_1$ be the linear predictor
for the log-odds. 
    + The logistic-regression parameters are $\alpha$ and $\beta_1$.

* It can be shown that 
$$
p = \frac{e^{LO}}{1+ e^{LO}}.
$$
    + i.e., the probability $p$ is the _logistic function_ of the log-odds.
    
## Graph of the Logistic Function

\small 

* $y$-axis is $p$; $x$-axis is $LO$; curve is function $p = \frac{e^{LO}}{1+ e^{LO}}.$

\scriptsize

```{r,echo=FALSE,fig.height=2.5, fig.width=4}
lo <- seq(from=-5,to=5,length=40)
plot(lo,exp(lo)/(1+exp(lo)),xlab="LO (log odds)",ylab="p",
     type="l")
```

\small

* On the $y$-axis, $p$ is constrained to be between 0 and 1 
and, on the $x$-axis, $LO$ is unconstrained.
    +  As $LO$ gets large and negative, $p$ approaches 0.
    +  As $LO$ gets large and positive, $p$ approaches 1.
    +  At $LO=0$, $p=1/2$. ($LO=0 \iff {\rm odds}=1.$)
    
  
    
    
## Fitting the model to the data

* To fit this logistic-regression model to the data and get the
parameter estimates, we use
a technique called the method of _maximum likelihood_. 
    + Likelihood methods for fitting statistical models to data 
    and obtaining parameter estimates are beyond the scope of this course.
    + Instead, see STAT 475 on Applied Discrete Data Analysis, for
    which STAT 305 is a pre-requisite.
    
* The slope parameter $\beta_1$ for $X_1$ summarizes the association
between $Y$ and $X_1$ in the population.

* For large sample sizes, the CLT kicks in and allows us to make approximate inference about the 
slope parameter $\beta_1$ using our fitted model.


## Review of Natural Logarithms and Exponents

* Recall that if $a$ is
the natural logarithm of $z$, written $a=\log(z)$, then
$e^a = e^{\log(z)} = z$.

* The logarithm of $1$ is always zero; 
    + e.g, $0=\log(1)$ and $e^0 = 1$.

* Sums of exponents are multiples; that is, 
$e^{a+b} = e^ae^b$.

* Differences of exponents are ratios; that is, 
$e^{a-b} = e^a/e^b$.
    + We will make use of this as $e^a/e^b = e^{a-b}$.

## Interpretation of $\beta_1$

* A one-unit (insert relevant units) increase in $X_1$ 
is associated with a change of $\beta_1$ 
in the log-odds of the outcome, or a $e^{\beta_1}$-fold
change in the odds of the outcome. 

* Need to take care with negative parameter values.

* E.G. Let's say that $\beta_1=-2$ and $X_1$ is measured
in grams. Then:
\begin{quote}
A one-gram increase in $X_1$ is associated with a change
of $-2$ in the log-odds of the outcome, or a $e^{-2}=0.135$-fold
change in the odds of the outcome. 
\end{quote}

* Rephrase to be shorter. E.G., if $X_1$ is birthweight and the outcome is BPD:
\begin{quote}
A one-gram increase in birthweight is associated with a $0.135$-fold
change in the odds of BPD. 
\end{quote}


## Mathematical justification of interpretation

* Let $p_1$ be the probability of Y = 1 given $X_1=x_1$.

* When $X_1=x_1$, 
we have log-odds
$$
\log \left[ \frac{p_1}{1-p_1} \right] = \alpha + \beta_1 x_1.
$$

* Let $p_2$ be the probability of Y = 1 given $X_1=x_1+1$.

* When $X_1=x_1+1$, we have log-odds 
$$
\log \left[ \frac{p_2}{1-p_2} \right] = \alpha + \beta_1 (x_1+1) = 
\alpha + \beta_1 x_1 + \beta_1.
$$

* The odds at $X_1=x_1$ and $X_1=x_1+1$ are, respectively,
$$
\frac{p_1}{1-p_1} = e^{\alpha + \beta_1 x_1}
\quad \mbox{and} \quad
\frac{p_2}{1-p_2}  = e^{\alpha + \beta_1 x_1 + \beta_1},
$$

* Hence the odds-ratio for $X_1=x_1+1$ relative to $X_1=x_1$ is
$$
\left ( \frac{p_2}{1-p_2} \right )  \bigg/ 
\left ( \frac{p_1}{1-p_1} \right )  = 
e^{\alpha + \beta_1 x_1 + \beta_1 - (\alpha + \beta_1 x_1)} = e^{\beta_1}.
$$


## Interpretation of $\beta_1$ for a Binary Exposure

* If $X_1$ is a binary exposure that takes values 1 for exposed
and 0 for unexposed,
a one-unit increase in $X_1$ means going from
unexposed to exposed.
* Set $x_1=0$ on the previous slide to find that
$e^{\beta_1}$ is the odds ratio for the exposed subjects
relative to the unexposed subjects.
* In the homework assignment, you will be asked to interpret 
fitted coefficients from a logistic regression on a binary 
exposure variable.

## BPD Example

* Let's read in the BPD data and look at it:

\scriptsize

```{r}
uu <- url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/bpd.csv")
bpd <- read.csv(uu)
head(bpd)
```

## Fit the Logistic Regression of BPD on Birth Weight

* To fit a logistic-regression model to the data, we use
the `glm()` function in R.

* Similar to the `lm()` function, `glm()` also requires 
a model formula.

    * The model formula is `bpd ~ birthwt`.
    * Response `bpd` on left-hand side of the ~ in the formula and explanatory
variable `birthwt` on the right-hand side are columns in the dataframe `bpd`.


\scriptsize

```{r}
bfit <- glm(bpd~birthwt,data=bpd,family=binomial)
coefficients(bfit)
```

\normalsize


* To three significant digits, the estimated parameters are
$\hat{\alpha}=4.03$ and $\hat{\beta}_1 = -0.00423$

## Software Notes

* When we fitted 
linear-regression models to data using the method of least squares, we used the _linear model_ 
or `lm()` function in R.

* When we fit logistic-regression models to data using the method
of maximum-likelihood,
we use the _generalized linear model_ or `glm()` function in R.

* In fact, `glm()` fits several types of models,
including linear and logistic regression. 

* Specify the type of model in `glm()` with the `family` option. 
    + Regular linear regression  is `family=gaussian` (default); same as using `lm()`.
    + Logistic regression is `family=binomial`.
        
+ **BEWARE:** Omitting the `family=binomial` argument in
    `glm()` will fit a linear regression to binary-outcome data.
    + ***For binary-outcome data, the fitted model will be nonsense.*** 
    
## Interpretation of Birth-Weight Effect

* To three significant digits, $\hat{\beta}_1 = -0.00423$
* Model the log-odds, but interpret in terms of the odds.
* We estimate that a one-gram 
increase in birth weight is associated with a $-0.00423$ change in
the log-odds of BPD.
* Report: "A one-gram increase in birth weight is associated with an
   estimated $e^{-0.00423}=0.996$-fold change in the odds of BPD."

* As one-gram units are too fine-grained, can instead work with a 100-gram 
increase in birth weight.
    + Then a 100-gram increase in birth weight is associated with an
estimated $100\times -0.00423=-0.423$ change in the log-odds of BPD.
    + Report: "A 100-gram increase in birth weight is 
    associated with an estimated $e^{-0.423}=0.655$-fold change in the
    odds of BPD."

## Log-Odds of BPD _vs._ Birthweight

* The logistic-regression model specifies a straight-line 
relationship between the log-odds of BPD
and `birthwt`.

\scriptsize

```{r,echo=FALSE,fig.height=2.8, fig.width=4}
newdat <- data.frame(birthwt = seq(from=450,to=1730,length=50))
library(dplyr)
newdat <- mutate(newdat,
           logodds = predict(bfit,newdata=newdat,type="link"),
           probability = predict(bfit,newdata=newdat,type="response"))
plot(newdat$birthw,newdat$logodds,
     xlab="birthwt (g)",ylab="Log-Odds of BPD",type="l",xaxt="n",yaxt="n")
# Lines to mark fitted value at birthwt=800 vs 1000 and 
# 1500 vs 1700
bb <- c(800,1000,1500,1700)
lo <- predict(bfit, newdata=data.frame(birthwt=bb),type="link")
axis(side=1,at=bb,labels=as.character(bb),cex.axis=.8)
axis(side=2,at=lo,labels=as.character(round(lo,3)),las=2,cex.axis=.7)
lines(c(0,bb[1]),c(lo[1],lo[1]),lty=3) 
lines(c(bb[1],bb[1]),c(-4,lo[1]), lty=3)
lines(c(0,bb[2]),c(lo[2],lo[2]),lty=3)
lines(c(bb[2],bb[2]),c(-4,lo[2]),lty=3)
lines(c(0,bb[3]),c(lo[3],lo[3]), lty=3)
lines(c(bb[3],bb[3]),c(-4,lo[3]), lty=3)
lines(c(0,bb[4]),c(lo[4],lo[4]), lty=3) 
lines(c(bb[4],bb[4]),c(-4,lo[4]), lty=3)
```
\small

\vspace*{-.15in}

* E.G.: A 200g increase in `birthwt` is associated with 
an estimated $200 \times \hat{\beta}_1=200\times -0.00423=-0.846$ 
change in log-odds of BPD. 
    + From plot, estimated log-odds of BPD in an 800g baby is $0.651$.
    + So, for a 1000g baby, it is $0.651-0.846=-0.195$
    
## Odds of BPD _vs._ Birthweight

\scriptsize

```{r,echo=FALSE,fig.height=3,fig.width=4}
plot(newdat$birthw,newdat$probability/(1-newdat$probability),
     xlab="birthwt (g)",
     ylab="Odds of BPD",type="l",xaxt="n",yaxt="n",ylim=c(0,2))
# Lines to mark fitted value at birthwt=600 vs 700 and 
# 1500 vs 1600
p <- predict(bfit, newdata=data.frame(birthwt=bb),type="response")
od<-p/(1-p)
axis(side=1,at=bb,labels=as.character(bb),cex.axis=.8)
axis(side=2,at=od,labels=as.character(round(od,3)),las=2,cex.axis=.5)
lines(c(0,bb[1]),c(od[1],od[1]),lty=3)
lines(c(bb[1],bb[1]),c(-4,od[1]), lty=3)
lines(c(0,bb[2]),c(od[2],od[2]), lty=3)
lines(c(bb[2],bb[2]),c(-4,od[2]), lty=3)
lines(c(0,bb[3]),c(od[3],od[3]), lty=3)
lines(c(bb[3],bb[3]),c(-4,od[3]), lty=3)
lines(c(0,bb[4]),c(od[4],od[4]), lty=3)
lines(c(bb[4],bb[4]),c(-4,od[4]), lty=3)
```

\normalsize

* Exponeniate to get odds. E.G.A 200g increase in `birthwt` is
associated with an estimated $e^{200\times-0.00423}=e^{-0.846}$
or $0.429$-fold change in the odds of BPD. 
    + From plot, estimated odds of BPD in an 800g baby are $e^{0.651}=1.92$.
    + For a 1000g baby, the odds are $e^{0.651-0.846}=e^{-0.195}=0.823$

## Probability of BPD _vs._ Birthweight

\scriptsize

```{r,echo=FALSE,fig.height=3, fig.width=4}
plot(newdat$birthw,newdat$probability,
     xlab="birthwt (g)",
     ylab="Probability of BPD",type="l",xaxt="n",yaxt="n")
# Lines to mark fitted value at birthwt=600 vs 700 and 
# 1500 vs 1600
p <- predict(bfit, newdata=data.frame(birthwt=bb),type="response")
axis(side=1,at=bb,labels=as.character(bb),cex.axis=.8)
axis(side=2,at=p,labels=as.character(round(p,3)),las=2,cex.axis=.7)
lines(c(0,bb[1]),c(p[1],p[1]),lty=3) 
lines(c(bb[1],bb[1]),c(-4,p[1]),lty=3)
lines(c(0,bb[2]),c(p[2],p[2]),lty=3)
lines(c(bb[2],bb[2]),c(-4,p[2]), lty=3)
lines(c(0,bb[3]),c(p[3],p[3]), lty=3)
lines(c(bb[3],bb[3]),c(-4,p[3]), lty=3)
lines(c(0,bb[4]),c(p[4],p[4]), lty=3)
lines(c(bb[4],bb[4]),c(-4,p[4]), lty=3)
```

\normalsize

* Saw that estimated log-odds of BPD in an 800g and 1000g baby are, respectively, $0.651$ and $0.651-0.846=-0.195$.
    + Therefore, corresponding probabilities of BPD are $e^{0.651}/(1+e^{0.651})=0.657$ for 800g babies and
    $e^{-0.195}/(1+e^{-0.195})=0.451$ for 1000g babies.



## Predicted Log-Odds and Probability of BPD

\small 

* We can use the `predict()` function to estimate
the log-odds or the probability of the outcome
at new values of the explanatory variable.

* The range of `birthwt` (in grams) in the `bpd` dataset is:

\scriptsize

```{r}
range(bpd$birthwt)
```

\small

* Let's consider new values of `birthwt` towards the extremes of this range,
for values 450.5g and 1729.5g

\scriptsize 

```{r}
newdat <- data.frame(birthwt = c(450.5,1729.5))
library(dplyr)
newdat <- mutate(newdat,
           logodds = predict(bfit,newdata=newdat,type="link"),
           probability = predict(bfit,newdata=newdat,type="response"))
newdat
```

## Software Notes

In the above calls to `predict()`:

\bigskip

* specifying the `type` argument as **`type=link`** requests predictions on the scale of the linear predictor; 
    + i.e. on the **log-odds scale**,
    + possible values of the log-odds are between $-\infty$ and $\infty$.

\bigskip

* specifying the `type` argument as **`type=response`** requests predictions on the scale of the response; 
    + i.e. on the **probability scale**,
    + possible values are between $0$ and $1$.  

## Fitting a Logistic Regression to Case-Control Data

+ The study of low-birthweight babies takes a simple random sample from a hospital ICU to see which babies have BPD.

+ But what if, instead, we had a case-control study.
    + A case-control study does not take a SRS from the population
but rather separate SRS's from cases and from controls.
    + Cases are typically over-sampled relative to their frequency in the population.
    
+ This is called _biased sampling_ and the case-control
study design is called a _biased sampling design_.

+ The biased sampling leads to biased estimates of the
intercept parameter $\alpha$ in the linear predictor and therefore of the log odds, odds and 
probabilities.
    + We can't estimate any of these on an absolute scale.
    
+ Fortunately, we **can**
    estimate the _changes_ in the log odds and odds because estimates of the slope parameter $\beta_1$ turn out to be unbiased. 

##

* Since the estimates of $\beta_1$ are not biased by the case-control
sampling: 
    + $\hat{\beta}_1$ can still be interpreted as
    the estimated effect of a one-unit increase in $X_1$ on the log-odds
    of the disease outcome.
    + $e^{\hat{\beta}_1}$ can still be interpreted as the estimated odds-ratio
    describing the multiplicative change resulting from a one-unit
    increase in $X_1$.

* The association between the binary disease outcome $Y$ and 
the explanatory variable $X_1$ is our main interest, and
$e^{\hat{\beta}_1}$ estimates an odds-ratio that
describes this association. 


