---
title: 'Statistics 305/605: Introduction to Biostatistical Methods for Health Sciences'
subtitle: 'Chapter 20, part 3: Multiple Logistic Regression'
author: "Jinko Graham"
date: '2017-11-13'
output: 
  beamer_presentation:
    includes:
      in_header: ../header_pagenum.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE,fig.height=3,fig.width=5)
```

## Heart Data For Multiple Logistic Regression

* To study the association between
atherosclerotic heart
disease (AHD) and maximum heart rate (MaxHR),
investigators randomly
sampled 303 patients presenting with chest pain
at a large hospital.

* They recorded information on the patients'
    + age in years, 
    + sex (1=male, 0=female), and 
    + MaxHR, the maximum heart rate in beats per minute
    + AHD diagnosis (1=Yes, 0=No) based on a coronary angiogram. 

\scriptsize

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/hrt.csv")
heart <- read.csv(uu)
head(heart)
```

\normalsize



## Multiple Logistic Regression

* Multiple logistic regression allows us to
    + investigate statistical interaction between explanatory variables
    + adjust for potential confounders
* Example: Could `sex` or `age` modify the relationship between
`MaxHR` and the odds of `AHD`?
* If not, does either variable confound the relationship between 
`MaxHR` and the odds of `AHD`?

## Multiple Logistic Regression Model

* We may model the log-odds of AHD as a function of 
$q$ explanatory variables $X_1, X_2, \ldots, X_q$; i.e.,
$$
\log \left[ \frac{p}{1-p} \right] = \alpha + \beta_1 X_1 + \beta_2 X_2 + \ldots \beta_q X_q,
$$
    + where:
        + $\log$ is the natural logarithm and 
        + $p$ is the probability of AHD given $X_1$,\ldots,$X_q$.
* Letting $LO =\alpha + \beta_1 X_ + \beta_2 X_2+ \ldots \beta_q X_q$, 
we have the logistic function:
$$
p = \frac{e^{LO}}{1+ e^{LO}}
$$


    

    

## Statistical Interaction 

- Interaction terms such as $X_1X_2$ appear as products of main effect terms such as $X_1$ and $X_2$:
    - $\log\left[\frac{p}{1-p}\right] = \alpha + \beta_1 X_1 + \beta_2X_2 + \beta_3 X_1X_2$
    
- In general, we allow the slopes for $X_1$ to depend
on the value of the modifying variable $X_2$.

- Let's focus on the two values $X_2=x_2$ and $X_2=x_2+1$ that are one unit apart:
    - Line for $X_2 = x_2$ baseline group:
        - $\log\left[\frac{p}{1-p}\right] = \alpha + \beta_1 X_1 + \beta_2x_2 + \beta_3 X_1x_2$
        - intercept is $\alpha + \beta_2 x_2$ and 
        - slope for $X_1$ is $\beta_1 + \beta_3 x_2$
    - Line for $X_2 = x_2+1$ group:  
        - $\log\left[\frac{p}{1-p}\right] = \alpha + \beta_1 X_1 + \beta_2(x_2+1) + \beta_3 X_1(x_2+1)$
        - intercept is $\alpha+\beta_2 (x_2+1)$ and 
        - slope for $X_1$ is $\beta_1+\beta_3(x_2+1)$
    - Difference between the slopes for $X_1$ in the two groups is 
   $\beta_1+\beta_3(x_2+1)  - (\beta_1+\beta_3x_2) = \beta_3$.
        
##

- The interaction coefficient $\beta_3$ is the difference between 
the slopes for $X_1$ in two
    groups that are defined by a one-unit increase in $X_2$.
    - If $X_2$ is a binary variable such as `sex`, a one-unit
    increase is from $X_2=0$ to $X_2=1$; i.e., from female to male.
- If $\beta_3=0$ then the slopes are the same.
    - Therefore, $X_2$ does **not** modify the effect of $X_1$ on $Y$.
- To assess statistical interaction between $X_1$ and $X_2$, test the hypotheses $H_0:\beta_3=0$ _vs._ $H_a:\beta_3 \not= 0$.

## Example: Interaction between MaxHR and Age

* Fitting ht model with `MaxHR`-by-`Age` interaction
gives the following table of coefficients:

\scriptsize

```{r}
hfit2 <- glm(AHD ~ MaxHR+Age+MaxHR:Age,data=heart,family=binomial)
summary(hfit2)$coefficients
```
\normalsize

\vspace*{-.12in}

* Estimated $\hat \beta_3=.00256$ difference between slopes for `MaxHR` in 2 groups defined by a one-unit change in `Age`.

* E.G. One group of patients of median `Age`  56 years
and another of patients of `Age` 57 years.

    + In patients of `Age` 57 years, the slope for `MaxHR` 
    is estimated to be 0.00256 more than in patients of `Age` 56 years.

## Effect of `MaxHR` on `AHD` in patients of different ages

\small

- The linear predictor or log-odds is 
$$ \log \left (\frac{p}{1-p} \right )
= \alpha + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2. $$

- In patients aged 57 years, simplifies to
$$\alpha + \beta_1 X_1 + \beta_2 57 + \beta_3 X_1 57
 =  \alpha + \beta_2 57 + (\beta_1 + \beta_3 57) X_1$$ 
<!--
\begin{eqnarray*} 
\log \left (\frac{p}{1-p} \right )
& = & \alpha + \beta_1 X_1 + \beta_2 57 + \beta_3 X_1 57. \\
& = & \alpha + \beta_2 57 + (\beta_1 + \beta_3 57) X_1 
\end{eqnarray*}
-->

- Slope for `maxHR` ($X_1$) in patients aged 57 years is 
$\beta_1 + \beta_3 57$; slope in patients aged 56 years
is $\beta_1 + \beta_3 56$.

- We estimate that the effect of `MaxHR` on the log-odds of `AHD` is:
    - $\hat{\beta}_1 + \hat{\beta_3} 57 = -0.18417+0.00256 \times 57 = -0.03825$ in patients aged **57** years, and
    - $\hat{\beta}_1 + \hat{\beta_3} 56 = -0.18417+0.00256 \times 56 = -0.04081$ in patients aged **56** years

- i.e., an increase of one beat-per-minute in `MaxHR` is associated with estimated decreases of 
    - $0.038$ in the log-odds of `AHD` in patients aged 57 years, and
    - $0.041$ in the log-odds of `AHD` in patients aged 56 years.

<!-- Heads up: by age 72 the estimated MaxHR slope is positive. This is not very obvious from the results, so hopefully no one will notice -->

## Does `Age` modify the effect of `MaxHR` on `AHD`?

* To address this question, test for statistical interaction
between `MaxHR` and `Age` at the 5\% level.

\scriptsize

```{r}
summary(hfit2)$coefficients
```

\normalsize 

* Compare the $p$-value for the interaction term 
`MaxHR:Age` to the level 0.05.
    - Since the $p$-value of 0.001 is less than the level, we reject the null hypothesis
    of no interaction.

* Conclude that `Age` **does** modify the effect of 
`MaxHR` on `AHD`.

# Confounding 

## Sex as a potential confounding variable

* `Age` has been declared to modify
the effect of `MaxHR` on `AHD` and so we can't
consider it as a potential confounding variable

* However, the binary variable `Sex` is not 
declared as an effect modifier at level 5\% (results not shown).
    + We may thus consider `Sex` as a potential confounding variable.

* We fit a model with `MaxHR` and `Sex` main effects,
and a model with a main effect only for `MaxHR`:

\scriptsize

```{r}
coefficients(glm(AHD~MaxHR+Sex,data=heart,family=binomial))
coefficients(glm(AHD~MaxHR,data=heart,family=binomial))
```

##

* We find that the estimated MaxHR effect changes by only

$$ \frac{|-0.0451 - (-0.0434)|}{|-0.0451|} \times 100\% = 3.7\%.$$ 
    
+ Notice that we use the estimate from the larger model in 
    the denominator.
* As the change is less than 10\%, we follow
convention and declare that `Sex` is not a confounder.
* A logistic regression model with a
main effect for `MaxHR` is therefore sufficient
and is summarized by:


\scriptsize

```{r}
hfitFinal <- glm(AHD~MaxHR,data=heart,family=binomial)
summary(hfitFinal)$coefficients
```

<!---
## 95\% CI for OR

\scriptsize

```{r}
hfitFinal <- glm(AHD~MaxHR,data=heart,family=binomial)
summary(hfitFinal)$coefficients
```

\normalsize

* A 95% CI for the OR parameter is as follows.

\scriptsize

```{r}
exp(confint(hfitFinal)["MaxHR",])
```

\normalsize

* We estimate with 95% confidence that an increase in 
MaxHR of 1 beat per minute is associated with 
a decrease in the odds of AHD by a factor of between
a 0.945 and 0.969.


## Test for an Effect of Sex on AHD

* To get more practice interpreting models with 
a binary exposure, we suppose 
our interest is in the effect of sex on 
the odds of AHD.
* An investigation of possible variables that could modify or confound
the relationship between sex and AHD 
turns up none (results not shown).
* A summary of the model for the odds of AHD as a function 
of sex is as follow.

\scriptsize

```{r}
sfit <- glm(AHD~Sex,data=heart,family=binomial)
summary(sfit)$coefficients
```

\normalsize

* A test of the sex effect is significant at the 5% level.
    + There is strong statistical evidence that the odds
    of AHD varies by sex.

## Point and Interval Estimates of the Sex Effect

* From the model summary on the previous slide,
the OR estimate for the odds of AHD in males _versus_ females
is $e^{1.27}= 3.56$.
    + We estimate that males have about a 3.56-times higher odds
    of AHD than females.
* A 95% confidence interval for the odds ratio of AHD in 
males _versus_ females is about (2.12,6.16)  (see R code below). 
    + With 95% confidence we estimate that males have 
    a 2.12 to 6.16 times higher odds of AHD than females.

\scriptsize

```{r}
confint(sfit)["Sex",] # CI for log-odds ratio
exp(confint(sfit)["Sex",]) # CI for odds ratio
```
-->

## Model checking and residual diagnostics

* There are measures of goodness-of-fit and residual diagnostics
for logistic regression.

    + However, these are difficult to interpret and beyond the scope of the course.

* See Stat 475 **Applied Discrete Data Analysis** if interested.

    + Stat 305 is a pre-requisite for Stat 475.