---
title: "Assignment 1, Question 1"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To answer this question we need a data frame 
that contains the numbers of hysterectomies for the
15 Swiss doctors.
We can create a data frame called `hyst` with one 
variable `numHyster`, as follows:
```{r}
hyst <- data.frame(numHyster = c(20,25,25,27,28,31,33,34,
                                 36,37,44,50,59,85,86))
```

Notice that I'm cutting-and-pasting from the `ch2-3Demo.Rmd`
demo. You
will find lots of useful R code in the demos that 
you can use, with little or no modification.

(a) Draw a histogram of the data. Do you see any outliers?

**Answer:**

```{r}
library(ggplot2)
# Add commands to draw a histogram here -- see the
# lecture notes for inspiration.
```

*Write your comments about outliers here*

(b) Compute the mean, standard deviation, median, 25th
percentile and 75th percentile. Comment on the difference
between the mean and the median, and the difference 
between the SD and the IQR.

**Answer:**

```{r}
library(dplyr)
# Use the summarize command as in the demo in ch2-3Demo.Rmd.
```

*Add your comments here*

