---
title: "Assignment 1, Question 3"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 10, problem 16. 
Include histograms of PDI and MDI scores. 
Note: There is some ambiguity in the question about whether 
we are supposed to use the data on all 143 babies or just 
the 70 babies with "trtment" variable equal to one. 
Please use the data on all 143 babies. (The "trtment" 
variable indicates the type of surgery that the babies 
obtained to correct the heart abnormality that they 
were born with.)

**Answer to histogram question**

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/heart.txt")
heart <- read.table(uu,header=TRUE)
head(heart)
library(ggplot2)
# Put your histogram commands here
```

(a) 
**Answer** 

Use the `t-test()` function to perform the test,
and then interpret the results.
```{r}
# t.test commands go here
```

*Interpretation here*


(b) 
**Answer**

```{r}
# t.test commands go here
```

*Interpretation here*



(c) 
**Answer**
