---
title: "Assignment 1, Question 3"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 10, problem 16, but with modification. 
Specifically, we will use the data on all 143 babies with
congenital heart disease rather than just the 70 babies 
who undergo reparative heart surgery during the first 3 
months of life (and so have "trtment" variable equal to one).

The Bayley Scales of Infant 
Development yield scores on 2 indices -- the Psychomotor 
Development Index (PDI) and the Mental Development Index (MDI)
-- which can be used to assess a child's level of functioning
in each of these areas at approximately one year of age.
The Bayley Scales were
administered to a sample of one-year-old infants born with
congenital heart disease. 

Before you do parts (a) - (d), draw a histogram of the PDI and
MDI scores. The histogram should be handed in with your assignment. 


**Answer to histogram question**

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/heart.txt")
heart <- read.table(uu,header=TRUE)
head(heart)
library(ggplot2)
# Put your histogram commands here
```

(a) At the 0.05 level of significance, test the null hypothesis
that the mean PDI score for children born with congenital
heart disease is equal to 100, the mean score for healthy
children. Use a 2-sided test. What is the pvalue? What
do you conclude?

**Answer** 

Use the `t-test()` function to perform the test,
and then interpret the results.
```{r}
# t.test commands go here
```

*Add your comments on interpretation here*


(b) Conduct the analogous test of hypothesis for the mean MDI
score. What do you conclude?

**Answer**

```{r}
# t.test commands go here
```

*Add your comments on interpretation here*


(c) Construct 95\% CIs for the true mean PDI score
and the true mean MDI score for this population of
children with congenital heart disease. Does either of
these intervals contain the value 100? Would you have
expected that they would?

**Answer**

*Add your interpretation here of the CI results in the t.test( ) output above.*
