---
title: "Assignment 1, Question 4"
author: "Your Name Here; Your Student ID number Here"
date: 'Put date here in YYYY-MM-DD format'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Chapter 11, problem 14. 

```{r}
uu<-url("http://people.stat.sfu.ca/~jgraham/Teaching/S305_18/Data/lowbwt.txt")
lowbwt <- read.table(uu,header=TRUE)
head(lowbwt)
```

(a) 
You are to plot a histogram of systolic blood
pressure (SBP) with observations **grouped by sex**.
To group by sex, we need to create a categorical variable out
of the numeric `sex` variable in the dataset. 

```{r, fig.height=3,fig.width=4}
library(dplyr)
lowbwt <- mutate(lowbwt,sexCateg = factor(sex))
```

Refer to the example histograms in the `ch11Demo.Rmd` demo
to see how to "fill" the bars of your histogram 
with different colours for different values of `sexCateg`.

**Answer** 

```{r}
library(ggplot2)
# histogram commands go here
```

*Comment on normality goes here*



(b) 
See the `ch11Demo.Rmd` demo for an example two-sample 
t-test with the `t.test()` function.

**Answer**

```{r}
# t.test commands go here
```

*Interpretation here*
